### 2022-09-07 doc/project1.tex
Make tex files

\documentclass{article}
	\usepackage[utf8]{inputenc}
	\usepackage{mathtools}
	\usepackage{amssymb}
	\usepackage{amsmath}
	\usepackage{bm}
	\usepackage{float}
	\usepackage{listings}
	\usepackage{nicefrac}
	\usepackage{xcolor}
	\usepackage[T1]{fontenc}
	\usepackage{newtxmath,newtxtext}
	\author{Jonatan Hoffmann Hanssen \and Eric Emanuel Reber}
	\title{Project 1 FYS-STK3155}

\begin{document}
	\maketitle

\section*{Abstract}

\end{document}

### 2022-09-08 doc/project1.tex

@@
 
+\section*{a}
+
+We have defined $\bm{y}$ as a function of a matrix multiplication $X\bm{\beta}$
+plus an error vector $\bm{\epsilon}$. This means that each element of the vector
+$\bm{y}$ can be expressed as follows:
+
+\[ y_i = \sum_j x_{ij}\beta_j + \epsilon_i \]
+
+If we take the expectation value of this expression we get the following:
+
+\[ E[y_i] = E\left[\sum_j x_{ij}\beta_j + \epsilon_i\right] \]
+\[ E[y_i] = E\left[\sum_j x_{ij}\beta_j\right] + E[\epsilon_i] \]
+
+However, the elements of $X$ are not stochastic, and neither are the elements of $\beta$
+the first expectation value is simply the sum itself. Furthermore, $\epsilon$ is 
+explicitly defined as a normal distribution $N(0,\sigma^2)$, and will by definition have
+the expectation value $0$. Therefore, we end up with the final expression:
+
+\[ E[y_i] = \sum_j x_{ij}\beta_j = \bm{X}_{i,*}\bm{\beta}\]
+
+We can use expectation values to calculate the variance as well:
+
+\[ var[y_i] = E\left[(y_i - E[y_i])^2\right] \]
+
 \end{document}

### 2022-09-08 doc/project1.tex
Add solution to first to tasks

@@
 \[ var[y_i] = E\left[(y_i - E[y_i])^2\right] \]
+\[ = E\left[y_i^2 - 2E[y_i]y_i + E[y_i]^2\right] \]
 
+Distributing the outer expectation value function:
+
+\[ var[y_i] = E[y_i^2] - 2E\left[E[y_i]\right]E[y_i] + E\left[E[y]^2\right] \]
+\[ = E[y_i^2] - 2E\left[E[y_i]\right]E[y_i] + E\left[E[y]^2\right] \]
+\[ = E[y_i^2] - 2E\left[E[y_i]\right]E[y_i] + E\left[E[y]\right]E\left[E[y]\right] \]
+
+The result of calculating the expectation value is non-stochastic. This means that
+$E[E[X]] = E[X]$. From this it follows that
+
+\[ = E[y_i^2] - 2E[y_i]E[y_i] + E[y_i]E[y_i] \]
+\[ = E[y_i^2] - E[y_i]E[y_i] \]
+\[ = E[y_i^2] - E[y_i]^2 \]
+
+The expectation value in the second summand has been proven to be equal 
+to $\bm{X}_{i,*}\bm{\beta}$ above. We therefore have
+
+\[ var[y_i] = E[y_i^2] - (\bm{X}_{i,*}\bm{\beta})^2 \]
+\[ = E\left[(X_{i,*}\bm{\beta})^2 + X_{i,*}\bm{\beta}\epsilon_i + \epsilon_i^2\right] - (\bm{X}_{i,*}\bm{\beta})^2 \]
+\[ = E\left[(X_{i,*}\bm{\beta})^2\right] + E[X_{i,*}\bm{\beta}\epsilon_i] + E[\epsilon_i^2] - (\bm{X}_{i,*}\bm{\beta})^2 \]
+
+$X_{i,*}\bm{\beta}$ and $\epsilon_i$ are both scalars. Therefore the expectation value can be written as
+$E[X_{i,*}\bm{\beta}]E[\epsilon_i]$. However, $E[\epsilon_i]$ is by definition $0$, because $\epsilon_i$ is
+defined as a normal distribution of mean $0$ and variance $\sigma^2$. Furthermore, the expectation value of the
+non-stochastic $(X_{i,*}\bm{\beta})^2$ is simply the expression itself. We therefore have
+
+\[ var[y_i] = (X_{i,*}\bm{\beta})^2 + E[X_{i,*}\bm{\beta}]\cdot 0 + E[\epsilon_i^2] - (\bm{X}_{i,*}\bm{\beta})^2 \]
+\[ = E[\epsilon_i^2] \]
+
+We can prove that this is equal to the variance of $\epsilon_i$:
+
+\[ E[\epsilon_i^2] = \frac{1}{n} \sum_i \epsilon_i^2\]
+\[ var[\epsilon_i] = \frac{1}{n} \sum_i (\epsilon_i - \bar{\epsilon_i})^2\]
+\[ var[\epsilon_i] = \frac{1}{n} \sum_i (\epsilon_i - 0)^2\]
+\[ var[\epsilon_i] = \frac{1}{n} \sum_i \epsilon_i^2 = E[\epsilon_i^2] \]
+
+And of course, we know that the variance of $\epsilon_i$ by definition is $\sigma^2$.
+
+\[ var[y_i] = E[\epsilon_i^2] = var[\epsilon_i] = \sigma^2 \]
+
+Proof of $var(\beta) = \sigma^{2}(X^{T}X)^{-1}$
+$$var(\beta) = var((X^{T}X)^{-1}X^{T}y)$$
+Because we are assuming $(X^{T}X)^{-1}X^{T}$ to be deterministic and $y$ to be stochastic, we can rewrite the $var(\beta)$ as following:
+$$var(\beta) = (X^{T}X)^{-1}X^{T}var(y)((X^{T}X)^{-1}X^{T})^{T}$$
+$$ = (X^{T}X)^{-1}X^{T}var(y)(X^{T})^{T}((X^{T}X)^{-1})^{T}$$
+Since $(X^{T})^{T} = X$ and $(X^{T}X)^{-1})^{T}$ = $(X^{T}X)^{T})^{-1} = (X^{T}X)^{-1}$
+$$ = (X^{T}X)^{-1}X^{T}var(y)X(X^{T}X)^{-1} $$
+We know var(y) = $\sigma^{2}$, and since it is a scalar it is commutative. Thus we may move it freely
+$$ = \sigma^{2}(X^{T}X)^{-1}X^{T}X(X^{T}X)^{-1} $$
+$$ = \sigma^{2}(X^{T}X)^{-1} $$
+$\blacksquare$
 \end{document}

### 2022-09-08 doc/project1.tex
Add solution to first to tasks

@@
 
-Proof of $var(\beta) = \sigma^{2}(X^{T}X)^{-1}$
-$$var(\beta) = var((X^{T}X)^{-1}X^{T}y)$$
-Because we are assuming $(X^{T}X)^{-1}X^{T}$ to be deterministic and $y$ to be stochastic, we can rewrite the $var(\beta)$ as following:
-$$var(\beta) = (X^{T}X)^{-1}X^{T}var(y)((X^{T}X)^{-1}X^{T})^{T}$$
-$$ = (X^{T}X)^{-1}X^{T}var(y)(X^{T})^{T}((X^{T}X)^{-1})^{T}$$
-Since $(X^{T})^{T} = X$ and $(X^{T}X)^{-1})^{T}$ = $(X^{T}X)^{T})^{-1} = (X^{T}X)^{-1}$
-$$ = (X^{T}X)^{-1}X^{T}var(y)X(X^{T}X)^{-1} $$
-We know var(y) = $\sigma^{2}$, and since it is a scalar it is commutative. Thus we may move it freely
-$$ = \sigma^{2}(X^{T}X)^{-1}X^{T}X(X^{T}X)^{-1} $$
-$$ = \sigma^{2}(X^{T}X)^{-1} $$
-$\blacksquare$
 \end{document}

### 2022-09-08 doc/project1.tex

--- 2022-09-08 doc/project1.tex
@@
 
+Proof of $var(\beta) = \sigma^{2}(X^{T}X)^{-1}$
+$$var(\beta) = var((X^{T}X)^{-1}X^{T}y)$$
+Because we are assuming $(X^{T}X)^{-1}X^{T}$ to be deterministic and $y$ to be stochastic, we can rewrite the $var(\beta)$ as following:
+$$var(\beta) = (X^{T}X)^{-1}X^{T}var(y)((X^{T}X)^{-1}X^{T})^{T}$$
+$$ = (X^{T}X)^{-1}X^{T}var(y)(X^{T})^{T}((X^{T}X)^{-1})^{T}$$
+Since $(X^{T})^{T} = X$ and $(X^{T}X)^{-1})^{T}$ = $(X^{T}X)^{T})^{-1} = (X^{T}X)^{-1}$
+$$ = (X^{T}X)^{-1}X^{T}var(y)X(X^{T}X)^{-1} $$
+We know var(y) = $\sigma^{2}$, and since it is a scalar it is commutative. Thus we may move it freely
+$$ = \sigma^{2}(X^{T}X)^{-1}X^{T}X(X^{T}X)^{-1} $$
+$$ = \sigma^{2}(X^{T}X)^{-1} $$
+$\blacksquare$
 \end{document}

### 2022-09-08 doc/project1.tex
add proof of expectation value

--- 2022-09-08 doc/project1.tex
@@
 
+\Proof of $E[\hat{\beta}] = \beta$
+$$E[\hat{\beta}] = E[(X^{T}X)^{-1}X^{T}y]$$
+Because we are assuming $(X^{T}X)^{-1}X^{T}$ to be deterministic, $E[(X^{T}X)^{-1}X^{T}] = (X^{T}X)^{-1}X^{T}$. Moreover, we know $E[y] = y$, and we can thus rewrite the equation as follows:
+$$E[(X^{T}X)^{-1}X^{T}y] = (X^{T}X)^{-1}X^{T}y = \beta$$
+$\blacksquare$
+\\
+\\[ var[y_i] = E\left[(y_i - E[y_i])^2\right] \]
+
 Proof of $var(\beta) = \sigma^{2}(X^{T}X)^{-1}$

### 2022-09-08 doc/project1.tex

@@
 
-\Proof of $E[\hat{\beta}] = \beta$
+Proof of $E[\hat{\beta}] = \beta$
 $$E[\hat{\beta}] = E[(X^{T}X)^{-1}X^{T}y]$$
@@
 \\
-\\[ var[y_i] = E\left[(y_i - E[y_i])^2\right] \]
 

### 2022-09-29 doc/project1.tex
Add beautiful proof

@@
 
-We have defined $\bm{y}$ as a function of a matrix multiplication $X\bm{\beta}$
+$\bm{y}$ is defined as a $f\bm{x} + \bm{\epsilon}$. This can be approximated as a matrix multiplication $X\bm{\beta}$
 plus an error vector $\bm{\epsilon}$. This means that each element of the vector
@@
 However, the elements of $X$ are not stochastic, and neither are the elements of $\beta$
+and therefore
 the first expectation value is simply the sum itself. Furthermore, $\epsilon$ is 
@@
 \[ var[y_i] = E[y_i^2] - 2E\left[E[y_i]\right]E[y_i] + E\left[E[y]^2\right] \]
-\[ = E[y_i^2] - 2E\left[E[y_i]\right]E[y_i] + E\left[E[y]^2\right] \]
+% \[ = E[y_i^2] - 2E\left[E[y_i]\right]E[y_i] + E\left[E[y]^2\right] \]
 \[ = E[y_i^2] - 2E\left[E[y_i]\right]E[y_i] + E\left[E[y]\right]E\left[E[y]\right] \]
@@
 
-\[ = E[y_i^2] - 2E[y_i]E[y_i] + E[y_i]E[y_i] \]
+\[ var[y_i] = E[y_i^2] - 2E[y_i]E[y_i] + E[y_i]E[y_i] \]
 \[ = E[y_i^2] - E[y_i]E[y_i] \]
@@
 
+\section{b}
+
 Proof of $E[\hat{\beta}] = \beta$
@@
 $\blacksquare$
+
+\section{c}
+
+\subsection{Proof of the bias-variance decomposition}
+
+Given data $\bm{y} = f(\bm{x}) + \bm{\epsilon}$ and a OLS model $\bm{\tilde{y}} = X\bm{\beta}$, we define
+the Mean Squared Error as follows
+
+\[C(X,\beta) = \frac{1}{n}\sum_{i=0}^{n-1}(y_i - \tilde{y_i})^2\]
+
+This is simply the expectation value of $(\bm{y} - \bm{\tilde{y}})^2$
+
+\[ C(X,\beta) = E\left[(\bm{y} - \bm{\tilde{y}})^2\right] \]
+
+To derive the bias variance decomposition, we now add $E[\bm{\tilde{y}}] - E[\bm{\tilde{y}}]$ 
+as a sneaky trick. We will also replace $\bm{y}$ with our definition $\bm{y} = \bm{f} + \bm{\epsilon}$
+
+\[ C(X,\beta) = E\left[(\bm{f} + \bm{\epsilon} - E[\bm{\tilde{y}}] + E[\bm{\tilde{y}}] - \bm{\tilde{y}})^2\right] \]
+\[ = E\left[(\bm{f} + \bm{\epsilon} - E[\bm{\tilde{y}}])^2 + 2\left[(\bm{f} + \bm{\epsilon} - E[\bm{\tilde{y}}])(E[\bm{\tilde{y}}] - \bm{\tilde{y}})\right] +(E[\bm{\tilde{y}}] - \bm{\tilde{y}})^2\right] \]
+\[ = E\left[(\bm{f} + \bm{\epsilon} - E[\bm{\tilde{y}}])^2\right] + 2E\left[(\bm{f} + \bm{\epsilon} - E[\bm{\tilde{y}}])(E[\bm{\tilde{y}}] - \bm{\tilde{y}})\right] + E\left[(E[\bm{\tilde{y}}] - \bm{\tilde{y}})^2\right] \]
+
+Let us focus our attention on the middle term first.
+
+\[ 2E\left[(\bm{f} + \bm{\epsilon} + E[\bm{\tilde{y}}])(E[\bm{\tilde{y}}] - \bm{\tilde{y}})\right] \]
+
+Expanding the multiplication:
+
+\[ 2E\left[\bm{f}E[\bm{\tilde{y}}] + \bm{\epsilon}E[\bm{\tilde{y}}] - E[\bm{\tilde{y}}]E[\bm{\tilde{y}}] - \bm{f}\bm{\tilde{y}} - \bm{\epsilon}\bm{\tilde{y}} + E[\bm{\tilde{y}}]\bm{\tilde{y}}\right] \]
+
+Distributing the expectation value:
+
+\[ 2\left[E[\bm{f}E[\bm{\tilde{y}}]] + E[\bm{\epsilon}E[\bm{\tilde{y}}]] - E[E[\bm{\tilde{y}}]E[\bm{\tilde{y}}]] - E[\bm{f}\bm{\tilde{y}}] - E[\bm{\epsilon}\bm{\tilde{y}}] + E[E[\bm{\tilde{y}}]\bm{\tilde{y}}]\right] \]
+
+We know that $\bm{\tilde{y}}$ and $\bm{f}$ are non-stochastic, and we know that the expectation value of $\bm{\epsilon}$ is $0$. We also know that 
+$\bm{\epsilon}$ and $\bm{\tilde{y}}$ are independent, which means we can write $E[\bm{\epsilon}\bm{\tilde{y}}]$ as $E[\bm{\epsilon}]E[\bm{\tilde{y}}]$ We therefore have
+
+\[ 2\left[\bm{f}\bm{\tilde{y}} + 0 - \bm{\tilde{y}}\bm{\tilde{y}} - \bm{f}\bm{\tilde{y}} - 0 + \bm{\tilde{y}}\bm{\tilde{y}}\right] = 0 \]
+
+We then have
+
+\[ C(X,\beta) = E\left[(\bm{f} + \bm{\epsilon} - E[\bm{\tilde{y}}])^2\right] + E\left[(E[\bm{\tilde{y}}] - \bm{\tilde{y}})^2\right] \]
+
+The second term is by definition the variance of model.
+
+\[ C(X,\beta) = E\left[(\bm{f} + \bm{\epsilon} - E[\bm{\tilde{y}}])^2\right] + var[\bm{\tilde{f}}] \]
+
+And now the final term
+
+\[ E\left[(\bm{f} + \bm{\epsilon} - E[\bm{\tilde{y}}])^2\right] \]
+
+\[ = E\left[\bm{f}^2 + \bm{f}\bm{\epsilon} - \bm{f}E[\bm{\tilde{y}}] + \bm{f\epsilon} + \bm{\epsilon}^2 - \bm{\epsilon}E[\bm{\tilde{y}}] - \bm{f}E[\bm{\tilde{y}}] - \bm{\epsilon}E[\bm{\tilde{y}}] + E[\bm{\tilde{y}}]^2 \right] \]
+
+After distributing the expectation value we get the following, keeping in mind that $E[\bm{f}] = E[\bm{y} - \bm{\epsilon}] = E[\bm{y}] - E[\bm{\epsilon}] = \bm{y}$
+
+\[ \bm{y}^2 - 2\bm{y}\bm{\tilde{y}} + E[\bm{\tilde{y}}]^2 + \sigma^2\]
+
+\[ (\bm{y} + E[\bm{\tilde{y}}]^2 + \sigma^2 \]
+
+We see that the quadratic form is equal to $bias[\bm{\tilde{y}}]^2$. We can then conclude that the cost function can be rewritten as follows:
+
+\[ C(X,\beta) = bias[\bm{\tilde{y}}]^2 + var[\bm{\tilde{f}}] + \sigma^2 \]
+
+% The first term here represents the bias, the second term is equal to zero and the last term represents the variance.
+
 \end{document}

### 2022-09-29 src/task_b.py
Groundbreaking change

@@ -114,8 +114,8 @@ plt.subplot(222)
 
 plt.plot(MSE_train, label="train implementation")
 plt.plot(MSE_test, label="test implementation")
-plt.plot(MSE_train_sk, "r--", label="train ScikitLearn") 
-plt.plot(MSE_test_sk, "g--", label="test ScikitLearn") 
+plt.plot(MSE_train_sk, "r--", label="train ScikitLearn")
+plt.plot(MSE_test_sk, "g--", label="test ScikitLearn")
 plt.ylabel("MSE score")
 plt.xlabel("Polynomial degree")
 plt.legend()

### 2022-10-04 doc/project1.tex
Start on method

@@
 
+\section*{Introduction}
+
+This is done by assuming that we can model our observed values as a linear combination of continous functions of the input values.
+For example, we can assume that our observed values can be approximately modeled by different polynomial degrees of the input values, and try to find the linear coefficients which minimize the difference between our observed values and the values we predict.
+This approach is essentially
+
+\section{Method}
+\subsection{Preprocessing: Creating the design matrix and normalizing our data}
+
+We have two datasets, one synthetic and one real.
+Regardless of the source of the data, we create the design matrix in a similar manner, as the relation between our input data and our target values are the same for both datasets, that is, for each combination of $x$ and $y$, we have a height $z$.
+For the Franke function this height is no more than the height over the $xy$-plane, but for our real data, this value represents the elevation in meters above ocean level for a terrain in Norway.
+As we want to do a polynomial fit of the data, we create our design matrix such that each row $i$ has the following values:
+
+\[ 1 \;\; x_i \;\; y_i \; x_i^2 \;\; x_i y_i \;\; y_i^2 \;\; ...\]
+
+These must be prepared in different ways to give us the best results.
+For the Franke function, we claim that no normalization is needed, as we generate the data ourselves to be between $0$ and $1$.
+Given that we are doing a polynomial fit, every element in our design matrix will be a multiplication of $x^i$ and $y^i$ for different $i$'s and $j$'s, which, given our choice of range for $x$ and $y$ guarantees that every element is between $0$ and $1$.
+This
+
+\subsection{Linear regression}
+
+In this paper, we use three forms of linear regression to approximate both a synthetic dataset (the so called Franke Function) and a real dataset comprised of elevation data of a landscape in Norway.
+We use th
+
 \section*{a}
 
-$\bm{y}$ is defined as a $f\bm{x} + \bm{\epsilon}$. This can be approximated as a matrix multiplication $X\bm{\beta}$
+$\bm{y}$ is defined as the $f(\bm{x}) + \bm{\epsilon}$. This can be approximated as a matrix multiplication $X\bm{\beta}$
 plus an error vector $\bm{\epsilon}$. This means that each element of the vector

### 2022-10-05 doc/paper/paper.tex
Add good paper layout files

%%% use twocolumn and 10pt options with the asme2ej format
\documentclass[twocolumn,10pt]{asme2ej}

\usepackage{graphicx} %% for loading jpg figures

%% The class has several options
%  onecolumn/twocolumn - format for one or two columns per page
%  10pt/11pt/12pt - use 10, 11, or 12 point font
%  oneside/twoside - format for oneside/twosided printing
%  final/draft - format for final/draft copy
%  cleanfoot - take out copyright info in footer leave page number
%  cleanhead - take out the conference banner on the title page
%  titlepage/notitlepage - put in titlepage or leave out titlepage
%  
%% The default is oneside, onecolumn, 10pt, final

\title{Linear regression for approximating real and synthetic data}

%%% first author
\author{Jonatan H. Hanssen
    \affiliation{
	Bachelor Student, Robotics and Intelligent Systems\\
	The faculty of Mathmatics and Natural Sciences\\
	Department of Informatics\\
    Email: jonatan.hanssen@outlook.com
    }	
}

\author{Eric E. Reber
    \affiliation{
	Bachelor Student, Robotics and Intelligent Systems\\
	The faculty of Mathmatics and Natural Sciences\\
	Department of Informatics\\
    Email: jonatan.hanssen@outlook.com
    }	
}

%%% second author
%%% remove the following entry for single author papers
%%% add more entries for additional authors
%%% third author
%%% remove the following entry for single author papers
%%% add more entries for additional authors

\begin{document}

\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
{\it This is the abstract.
This article illustrates preparation of ASME paper using 
\LaTeX2\raisebox{-.3ex}{$\epsilon$}.
An abstract for an ASME paper should be less than 150 words and is normally in italics.
%%% 
Please use this template to test how your figures will look on the printed journal page of the Journal of Mechanical Design.  The Journal will no longer publish papers that contain errors in figure resolution.  These usually consist of unreadable or fuzzy text, and pixilation or rasterization of lines.  This template identifies the specifications used by JMD some of which may not be easily duplicated; for example, ASME actually uses Helvetica Condensed Bold, but this is not generally available so for the purpose of this exercise Helvetica is adequate.  However, reproduction of the journal page is not the goal, instead this exercise is to verify the quality of your figures. Notice that this abstract is to be set in 9pt Times Italic, single spaced and right justified.  
}
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \begin{nomenclature}
% \entry{A}{You may include nomenclature here.}
% \entry{$\alpha$}{There are two arguments for each entry of the nomemclature environment, the symbol and the definition.}
% \end{nomenclature}
%
% The primary text heading is  boldface and flushed left with the left margin.  The spacing between the  text and the heading is two line spaces.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

Linear regression is a simple method in machine learning, used to create a fit of a dataset of which it can produce continuous predictions. This is achieved by assuming that we can model the output values as a linear combination of continuous functions of the input values. For example, we can assume that the observed values in our dataset can be approximated by different polynomial degrees of the input values and find the linear coefficients $\beta$ which minimize the difference between our observed and predicted values. This general idea that we can create a model which approximates our observed data with a minimal difference between observed and predicted values is a problem at the very heart of machine learning.

However, in order to create an optimal model, we require optimal parameters which cannot be known a priori. Thus, we require extensive testing and tuning of parameters to inform our selection. This presents a major obstacle, as we must for example find the optimal polynomial degree $N$ for which our model best approximates the observed values at the same time as we tune the hyperparameter $\lambda$ responsible for an adequate regularisation. We must find these parameters which lie at an intersection, where all parameters yield the minimal error in unity.

Furthermore, we will compare different models of linear regression, namely Ordinary Least Squares (OLS), Ridge and Lasso regression, and employ resampling methods, namely bootstrap and cross-validation to ensure that we draw conclusive results as these help us reduce the randomness in our parameters based off how we split our data. <source?>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Method}

% The heading is boldface with upper and lower case letters. 
% If the heading should run into more than one line, the run-over is not left-flushed.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Preprocessing: Creating the design matrix and normalizing our data}

We have two datasets, one synthetic and one real. Regardless of the source of the data, we create the design matrix in a similar manner, as the relation between our input data and our target values are the same for both datasets, that is, for each combination of $x$ and $y$, we have a height $z$. For the Franke function this height is no more than the height over the $xy$-plane, but for our real data, this value represents the elevation in meters above ocean level for a terrain in Norway. As we want to do a polynomial fit of the data, we create our design matrix such that each row $i$ has the following values:

\[ X_{i*} = 1 \;\; x_i \;\; y_i \; x_i^2 \;\; x_i y_i \;\; y_i^2 \;\; ... \;\; x_i^p y_i^p\]

While the creating of the design matrix is equivalent for the two datasets, the preprocessing of the data afterwards is not. For the Franke function, we claim that no normalization is needed, as we generate the data ourselves to be between $0$ and $1$. Given that we are doing a polynomial fit, every element in our design matrix will be a multiplication of $x^p$ and $y^k$ for different $p$'s and $k$'s, which, given our choice of range for $x$ and $y$, guarantees that every element is between $0$ and $1$. However, we have included a centering of the data, which has no effect on the Ordinary Least Squares regression, but which is used when applying Ridge and Lasso regression, to avoid punishing the intercept.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Third-Level Heading.}

The third-level of heading follows the style of the second-level heading.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Use of SI Units}

An ASME paper should use SI units.  When preference is given to SI units, the U.S. customary units may be given in parentheses or omitted. When U.S. customary units are given preference, the SI equivalent {\em shall} be provided in parentheses or in a supplementary table.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Footnotes\protect\footnotemark}
\footnotetext{Examine the input file, asme2ej.tex, to see how a footnote is given in a head.}

Footnotes are referenced with superscript numerals and are numbered consecutively from 1 to the end of the paper\footnote{Avoid footnotes if at all possible.}. Footnotes should appear at the bottom of the column in which they are referenced.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Mathematics}

Equations should be numbered consecutively beginning with (1) to the end of the paper, including any appendices.  The number should be enclosed in parentheses and set flush right in the column on the same line as the equation.  An extra line of space should be left above and below a displayed equation or formula. \LaTeX\ can automatically keep track of equation numbers in the paper and format almost any equation imaginable. An example is shown in Eqn.~(\ref{eq_ASME}). The number of a referenced equation in the text should be preceded by Eqn.\ unless the reference starts a sentence in which case Eqn.\ should be expanded to Equation.

\begin{equation}
f(t) = \int_{0_+}^t F(t) dt + \frac{d g(t)}{d t}
\label{eq_ASME}
\end{equation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Figures}
\label{sect_figure}

All figures should be positioned at the top of the page where possible.  All figures should be numbered consecutively and centered under the figure as shown in Fig.~\ref{figure_ASME}. All text within the figure should be no smaller than 7~pt. There should be a minimum two line spaces between figures and text. The number of a referenced figure or table in the text should be preceded by Fig.\ or Tab.\ respectively unless the reference starts a sentence in which case Fig.\ or Tab.\ should be expanded to Figure or Table.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%% begin figure %%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\begin{center}
\setlength{\unitlength}{0.012500in}%
\begin{picture}(115,35)(255,545)
\thicklines
\put(255,545){\framebox(115,35){}}
\put(275,560){Beautiful Figure}
\end{picture}
\end{center}
\caption{The caption of a single sentence does not have period at the end}
\label{figure_ASME} 
\end{figure}
%%%%%%%%%%%%%%%% end figure %%%%%%%%%%%%%%%%%%% 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In the following subsections, I have inserted figures that have been provided by authors in order to demonstrate what to avoid.  In each case the authors provided figures that are 3.25in wide and 600dpi in the .tif graphics format.  The papers containing these figures have been held from production due to their poor quality.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The 1st Example of Bad Figure}

%%%%%%%%%%%%%%%% begin figure %%%%%%%%%%%%%%%%%%%
%%% 3.34in is the maximum width you can have for a figure
\begin{figure} 
\centerline{\includegraphics[width=3.34in]{figure/FMANU_MD_05_1107_11.jpg}}
\caption{Example taken from a paper that was held from production because the image quality is poor.  ASME sets figures captions in 8pt, Helvetica Bold.}
\label{fig_example1.jpg}
\end{figure}
%%%%%%%%%%%%%%%% end figure %%%%%%%%%%%%%%%%%%%

In order to place the figure in this template using MSWord, select Insert Picture from File, and use wrapping that is top and bottom. Make sure the figure is 3.25in wide.

Figure~`\ref{fig_example1.jpg}
was taken from a recent paper that was held from publication, because the text is fuzzy and unreadable. It was probably obtained by taking a screen shot of the computer output of the authors software. This means the original figure was 72dpi (dots per inch) on a computer screen. There is no way to improve the quality such a low resolution figure.

In order to understand how poor the quality of this figure is, please zoom in slightly, say to 200\%.  Notice that while the font of the paper is clear at this size, the font in the figures is fuzzy and blurred.  It is impossible to make out the small symbol beside the numbers along the abscissa of the graph.  Now consider the labels Time and Cost. They are clearly in fonts larger that the text of the article, yet the pixilation or rasterization, associated with low resolution is obvious. This figure must be regenerated at higher resolution to ensure quality presentation.

The poor quality of this figure is immediately obvious on the printed page, and reduces the impact of the research contribution of the paper, and in fact detracts from the perceived quality of the journal itself.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The 2nd Example of Bad Figure}

%%%%%%%%%%%%%%%% begin figure %%%%%%%%%%%%%%%%%%%
\begin{figure} 
\centerline{\includegraphics[width=3.34in]{figure/FMANU_MD_05_1272_5.jpg}}
\caption{While this figures is easily readable at a double column width of 6.5in, when it is shrunk to 3.25in column width the text is unreadable. This paper was held from production.}
\label{fig_example2.jpg}
\end{figure}
%%%%%%%%%%%%%%%% end figure %%%%%%%%%%%%%%%%%%%

Figure~\ref{fig_example2.jpg}
demonstrates a common problem that arises when a figure is scaled down fit a single column width of 3.25in.  The original figure had labels that were readable at full size, but become unreadable when scaled to half size.  This figure also suffers from poor resolution as is seen in the jagged lines the ovals that form the chain.

This problem can be addressed by increasing the size of the figure to a double column width of 6.5in, so the text is readable.  But this will not improve the line pixilation, and a large low resolution figure is less desirable than a small one.  This also significantly expands the length of the paper, and may cause it to exceed the JMD nine page limit.  Additional pages require page charges of \$200 per page.  It is best to regenerate the figure at the resolution that ensures a quality presentation.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The 3rd Example of Bad Figure}
%%%%%%%%%%%%%%%% begin figure %%%%%%%%%%%%%%%%%%%
\begin{figure} 
\centerline{\includegraphics[width=3.25in]{figure/FMANU_MD_04_1274_13.jpg}}
\caption{Another example of a figure with unreadable text.  Even when the paper was expanded to double column width the text as shown in Fig.~\ref{fig_example4.jpg} was of such low quality that the paper was held from production.}
\label{fig_example3.jpg}
\end{figure}
%%%%%%%%%%%%%%%% end figure %%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%% begin figure %%%%%%%%%%%%%%%%%%%
%%% the maximum width in double column is 6.85in
\begin{figure*} 
\centerline{\includegraphics[width=6.85in]{figure/FMANU_MD_04_1274_13.jpg}}
\caption{A figure expanded to double column width the text from Figure~\ref{fig_example3.jpg}}
\label{fig_example4.jpg}
\end{figure*}
%%%%%%%%%%%%%%%% end figure %%%%%%%%%%%%%%%%%%%
An author provided the high resolution image 
in Fig.~\ref{fig_example3.jpg}
that was sized to a single column width of 3.25in.  Upon seeing the poor quality of the text, the publisher scaled the image to double column width as shown in Fig.~\ref{fig_example4.jpg} 
at which point it took half of a page.  The publisher went on to do this for all eight figures generating four pages of figures that the author did not expect. ASME stopped production of the paper even with the larger figures due to the pixilation of the font.

Clearly the text in this figure is unreadable, and it is doubtful that the author can print the output in a way that it is readable.  This is a problem that the author must solve, not the publisher.

As you might expect, I have many more examples, but in the end the author is the best judge of what is needed in each figure.  ASME simply requires that the image meet a minimum standard for font and line quality, specifically the font should be the appropriate size and not be blurred or pixilated, and that lines should be the appropriate weight and have minimal, preferably no, pixilation or rasterization.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Tables}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%% begin table   %%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[t]
\caption{Figure and table captions do not end with a period}
\begin{center}
\label{table_ASME}
\begin{tabular}{c l l}
& & \\ % put some space after the caption
\hline
Example & Time & Cost \\
\hline
1 & 12.5 & \$1,000 \\
2 & 24 & \$2,000 \\
\hline
\end{tabular}
\end{center}
\end{table}
%%%%%%%%%%%%%%%% end table %%%%%%%%%%%%%%%%%%% 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

All tables should be numbered consecutively  and centered above the table as shown in Table~\ref{table_ASME}. The body of the table should be no smaller than 7 pt.  There should be a minimum two line spaces between tables and text.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Citing References}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The ASME reference format is defined in the authors kit provided by the ASME.  The format is:

\begin{quotation}
{\em Text Citation}. Within the text, references should be cited in  numerical order according to their order of appearance.  The numbered reference citation should be enclosed in brackets.
\end{quotation}

The references must appear in the paper in the order that they were cited.  In addition, multiple citations (3 or more in the same brackets) must appear as a `` [1-3]''.  A complete definition of the ASME reference format can be found in the  ASME manual \cite{asmemanual}.

The bibliography style required by the ASME is unsorted with entries appearing in the order in which the citations appear. If that were the only specification, the standard {\sc Bib}\TeX\ unsrt bibliography style could be used. Unfortunately, the bibliography style required by the ASME has additional requirements (last name followed by first name, periodical volume in boldface, periodical number inside parentheses, etc.) that are not part of the unsrt style. Therefore, to get ASME bibliography formatting, you must use the \verb+asmems4.bst+ bibliography style file with {\sc Bib}\TeX. This file is not part of the standard BibTeX distribution so you'll need to place the file someplace where LaTeX can find it (one possibility is in the same location as the file being typeset).

With \LaTeX/{\sc Bib}\TeX, \LaTeX\ uses the citation format set by the class file and writes the citation information into the .aux file associated with the \LaTeX\ source. {\sc Bib}\TeX\ reads the .aux file and matches the citations to the entries in the bibliographic data base file specified in the \LaTeX\ source file by the \verb+\bibliography+ command. {\sc Bib}\TeX\ then writes the bibliography in accordance with the rules in the bibliography .bst style file to a .bbl file which \LaTeX\ merges with the source text.  A good description of the use of {\sc Bib}\TeX\ can be found in \cite{latex, goosens} (see how two references are handled?).  The following is an example of how three or more references \cite{latex, asmemanual,  goosens} show up using the \verb+asmems4.bst+ bibliography style file in conjunction with the \verb+asme2ej.cls+ class file. Here are some more \cite{art, blt, ibk, icn, ips, mts, mis, pro, pts, trt, upd} which can be used to describe almost any sort of reference.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}
The only way to ensure that your figures are presented in the ASME Journal of Mechanical Design in the way you feel is appropriate and meets the requirement for quality presentation is for you to prepare a double column version of the paper in a form similar to that used by the Journal.

This gives you the opportunity to ensure that the figures are sized appropriately, in particular that the labels are readable and match the size of the text in the journal, and that the line weights and resolutions have no pixilation or rasterization.  Poor quality figures are immediately obvious on the printed page, and this detracts from the perceived quality of the journal.

I am pleased to provide advice on how to improve any figure, but this effort must start with a two-column version of the manuscript. Thank you in advance for your patience with this effort, it will ensure quality presentation of your research contributions.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussions}
This template is not yet ASME journal paper format compliant at this point.
More specifically, the following features are not ASME format compliant.
\begin{enumerate}
\item
The format for the title, author, and abstract in the cover page.
\item
The font for title should be 24 pt Helvetica bold.
\end{enumerate}

\noindent
If you can help to fix these problems, please send us an updated template.
If you know there is any other non-compliant item, please let us know.
We will add it to the above list.
With your help, we shall make this template 
compliant to the ASME journal paper format.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{acknowledgment}
ASME Technical Publications provided the format specifications for the Journal of Mechanical Design, though they are not easy to reproduce.  It is their commitment to ensuring quality figures in every issue of JMD that motivates this effort to have authors review the presentation of their figures.

Thanks go to D. E. Knuth and L. Lamport for developing the wonderful word processing software packages \TeX\ and \LaTeX. We would like to thank Ken Sprott, Kirk van Katwyk, and Matt Campbell for fixing bugs in the ASME style file \verb+asme2ej.cls+, and Geoff Shiflett for creating 
ASME bibliography stype file \verb+asmems4.bst+.
\end{acknowledgment}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% The bibliography is stored in an external database file
% in the BibTeX format (file_name.bib).  The bibliography is
% created by the following command and it will appear in this
% position in the document. You may, of course, create your
% own bibliography by using thebibliography environment as in
%
% \begin{thebibliography}{12}
% ...
% \bibitem{itemreference} D. E. Knudsen.
% {\em 1966 World Bnus Almanac.}
% {Permafrost Press, Novosibirsk.}
% ...
% \end{thebibliography}

% Here's where you specify the bibliography style file.
% The full file name for the bibliography style file 
% used for an ASME paper is asmems4.bst.
\bibliographystyle{asmems4}

% Here's where you specify the bibliography database file.
% The full file name of the bibliography database for this
% article is asme2e.bib. The name for your database is up
% to you.
\bibliography{asme2e}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\appendix       %%% starting appendix
\section*{Appendix A: Head of First Appendix}
Avoid Appendices if possible.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Appendix B: Head of Second Appendix}
\subsection*{Subsection head in appendix}
The equation counter is not reset in an appendix and the numbers will
follow one continual sequence from the beginning of the article to the very end as shown in the following example.
\begin{equation}
a = b + c.
\end{equation}

\end{document}

### 2022-10-05 doc/paper/paper.tex
More on paper

@@
 	Department of Informatics\\
-    Email: jonatan.hanssen@outlook.com
+    Email: jonatahh@ifi.uio.no
+    Email: jonatanindahouse@gmail.com
     }	
@@
 	Department of Informatics\\
-    Email: jonatan.hanssen@outlook.com
+    Email: ericer@ifi.uio.no
     }	
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{Method}
 
+\subsection{Theory}
+
+This sections covers the theoretic background which underlies the methods used in this paper.
+
+\subsubsection{Ordinary Least Squares}
+
+\subsubsection{Proofs}
 % The heading is boldface with upper and lower case letters. 
@@
 However, we have included a centering of the data, which has no effect on the Ordinary Least Squares regression, but which is used when applying Ridge and Lasso regression, to avoid punishing the intercept.
+The centering is done by
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{Third-Level Heading.}
+
+\subsection{Implementation of linear regression}
 

### 2022-10-05 doc/paper/paper.tex
Add linear regresison theory

@@
 %%% use twocolumn and 10pt options with the asme2ej format
-\documentclass[twocolumn,10pt]{asme2ej}
+\documentclass[twocolumn,10pt,cleanfoot]{asme2ej}
 
 \usepackage{graphicx} %% for loading jpg figures
+\usepackage{bm}
+\usepackage{nicefrac}
+\usepackage{mathtools}
+\usepackage{amssymb}
+\usepackage{amsmath}
+\usepackage{parskip}
 
@@
     Email: jonatahh@ifi.uio.no
-    Email: jonatanindahouse@gmail.com
     }
@@
 }
-
-%%% second author
-%%% remove the following entry for single author papers
-%%% add more entries for additional authors
-%%% third author
-%%% remove the following entry for single author papers
-%%% add more entries for additional authors
 
@@
 
-\subsubsection{Ordinary Least Squares}
+\subsubsection{Linear regression}
 
-\subsubsection{Proofs}
-% The heading is boldface with upper and lower case letters. 
-% If the heading should run into more than one line, the run-over is not left-flushed.
+To describe how linear regression works, let us first describe the problem it aims to solve.
+Assume we have a dataset, which contains a series of inputs $\{x_i\}$, and corresponding target values $\{y_i\}$.
+Our assumption is that there is some relationship $y_i = f(x_i) + \epsilon_i$ where $\epsilon_i \sim N(0,\sigma^2)$, and we wish to create a model which approximates this relationship.
+More specifically, we wish to find a vector $\beta$ which gives us the "best" approximation $\bm{y_{pred}}$ of $\bm{y}$, by using the parameters in the following equation:
+
+\begin{equation}
+X \bm{\beta} = \bm{\tilde{y}}
+\end{equation}
+
+$X$ is called the design matrix, and is an $n$ by $p$ matrix created by our input data and our choice of basis functions, where $n$ is the amount of different input states we have (for our problem, $n$ is the number of distinct $x$ values times the number of distinct $y$ values) and $p$ is the number of features we use for our fit (in our case, $p$ is based on the polynomial degree we use).
+The element $X_{ij}$ is calculated by evaluating the basis function corresponding the coloumn $j$ for the input data corresponding to the row $j$.
+As we can see, each row of this matrix is scalar multiplied by the vector $\beta$, giving us a linear combination of the values of $X_{i*}$ for each $y_i$ in the target vector $\bm{y}$.
+This vector is called the parameter vector, and finding the optimal values $\hat{\bm{\beta}}$ is the problem we wish to solve.
+But for this, we need to define what an "optimal" prediction $\bm{\tilde{y}}$ is.
+
+\subsubsection{The cost function}
+
+The define the optimal prediction, we use a cost function, and the choice of cost function greatly influences what optimal parameters we get.
+FOOTNOTE: It is here important to keep in mind that the cost function used when fitting the model (training) rarely is the same as the one used for testing our prediction, as described in <somewhere>.
+An intuitive cost function is the mean squared error (MSE).
+This function simply computes the average distance between our prediction and the target value, squared:
+
+\begin{equation}
+	C_{mse}(\tilde{\bm{y}}) = \sum_{i=0}^{n-1}(y_i-\tilde{y_i})^2
+\end{equation}
+
+We may now plug in the definition of $\bm{\tilde{y}}$.
+
+\begin{equation}
+	C_{mse}(\bm{\beta}) = \sum_{i=0}^{n-1}(y_i-X_{i*}\bm{\beta})^2
+\end{equation}
+
+We see that we have a function of our parameters $\bm{\beta}$, which describe how much our prediction deviates from our target values.
+Our problem now is to optimize this function, that is, find the values of $\beta$ which give MSE the lowest value.
+Luckily for us, it can be proven that the second derivative of this function is always positive (WHERE?), which means that we can find the global minimum by solving the equation
+
+Hastie 45
+
+\begin{gather}
+\frac{\delta}{\delta \beta} \sum_{i=0}^{n-1}(y_i-X_{i*}\bm{\beta})^2 = 0 \\
+ - \frac{2}{n} X^T(\bm{y} - X\bm{\beta}) = 0
+\end{gather}
+
+Solving this for $\beta$ gives us the so called normal equation, an analytical solution\footnote{This solution requires that $X^TX$ is non-singular, which may not be the case.
+This is solved by using the singular value decomposition of the design matrix, which we will not go into here} for the optimal $\hat{\beta}$.
+
+% \begin{equation}
+\begin{gather}
+- \frac{2}{n} X^T(\bm{y} - X\bm{\beta}) = 0 \\
+X^T(\bm{y} - X\bm{\beta}) = 0 \\
+X^T\bm{y} - X^TX\bm{\beta} = 0 \\
+X^T\bm{y} = X^TX\bm{\beta} \\
+\bm{\hat{\beta}} = (X^TX)^{-1}X^T\bm{y}
+\end{gather}
+
+The process we have just described, with MSE as the cost function, is the what is known as Ordinary Least Squares (OLS) regression.
+Using different cost functions when fitting the model leads to different equations for $\bm{\hat{\beta}}$.
+This leads us to the two next forms of linear regression we use in this paper.
+
+\subsubsection{Ridge Regression}
+
+	A common problem with OLS is what is known as overfitting (we will discuss this in greater detail in later sections).
+A symptom of overfitting is that the values in the parameter vector $\bm{\beta}$ grow very large, and vary wildly for small changes in our dataset.
+To counteract this, we can add to the cost function a term which increases with the values in $\bm{\beta}$.
+Optimizing the cost function would now not only mean finding the smallest difference between target and predicted values, but also keeping the parameters small.
+We now define a new cost function
+
+\begin{equation}
+	C_{lasso}(\bm{\beta}) = \sum_{i=0}^{n-1}(y_i-X_{i*}\bm{\beta})^2 + \lambda \sum_{j=0}^{p-1} \beta_j^2
+\end{equation}
+
+	$||\beta||_2^2$ is equal to $\beta^T\beta$ and is simply the norm of the parameter vector, squared.
+With the new hyperparameter $\lambda \in (0, \inf)$ , we can decide how harshly we should punish large values in our parameter vector.
+Finding an optimal $\lambda$ is an important step in the process of finding the best parameters for our model.
+
+A similar derivation as done in the previous section gives us the normal equation for Ridge regression:
+
+\begin{equation}
+	\bm{\hat{\beta}} = (X^TX + \lambda I)^{-1}X^T\bm{y}
+\end{equation}
+
+	We see that this equation, with a $\lambda > 0$ avoids the problem of an inversion of a non-singular matrix, and gives us an analytical solution for $\bm{\hat{\beta}}$ yet again.
+\footnote{We should cherish this analytical solution, as it may well be the last time we see a cost function that can be optimized analytically.} By using ridge regresison, we can control the complexity of our model and effectively hinder overfitting.
+
+\subsubsection{Lasso Regression}
+
+	Another cost function is one where we add the L1-norm (the sum of the absolute values of the parameters), instead of the L2-norm. The new cost function is as follows.
+
+\begin{equation}
+	C_{lasso}(\bm{\beta}) = \sum_{i=0}^{n-1}(y_i-X_{i*}\bm{\beta})^2 + \lambda \sum_{j=0}^{p-1} |\beta_j|
+\end{equation}
+
+However, the derivative of the absolute value is not continous, so there is sadly no analytical solution for the optimal parameters with this form of linear regression.
+We must then solve the problem numerically, which can be done by using Gradient Descent.
+Like Ridge regression, this form of regression controls the complexity of the model, with the added benifit that Lasso can drive parameters to exactly zero.
 
@@
 
-\[ X_{i*} = 1 \;\; x_i \;\; y_i \; x_i^2 \;\; x_i y_i \;\; y_i^2 \;\; ... \;\; x_i^p y_i^p\]
+\begin{equation}
+X_{i*} = 1 \;\; x_i \;\; y_i \; x_i^2 \;\; x_i y_i \;\; y_i^2 \;\; ... \;\; x_i^p y_i^p
+\end{equation}
 
@@
 \section{Footnotes\protect\footnotemark}
-\footnotetext{Examine the input file, asme2ej.tex, to see how a footnote is given in a head.}
-
-Footnotes are referenced with superscript numerals and are numbered consecutively from 1 to the end of the paper\footnote{Avoid footnotes if at all possible.}.
-Footnotes should appear at the bottom of the column in which they are referenced.
 

### 2022-10-05 src/task_b.py
change redundancies, add resampling and bias-variance method in report

@@ -22,7 +22,6 @@ z = FrankeFunction(x, y)
 N = 20
 scaling = True
 
-# Do the linear_regression
 z += 0.05 * np.random.standard_normal(z.shape)
 X, X_train, X_test, z_train, z_test = preprocess(x, y, z, N, 0.2)
 betas, z_preds_train, z_preds_test, z_preds = linreg_to_N(
@@ -40,7 +39,7 @@ OLS_scikit = LinearRegression(
 _, z_preds_train_sk, z_preds_test_sk, _ = linreg_to_N(
     X, X_train, X_test, z_train, z_test, N, scaling=scaling, model=OLS_scikit
 )
-
+assert np.allclose(z_preds_test_sk[:3, :], z_preds_test[:3,:])
 # X, X_train, X_test, z, z_train, z_test = normalize_task_g(
 #     X, X_train, X_test, z, z_train, z_test
 # )

### 2022-10-05 src/utils.py

@@ -110,7 +110,6 @@ def crossval(
     for k in range(K):
         if k == K - 1:
             # if we are on the last, take all thats left
-            X_test = np.zeros((X.shape[0] - K * chunksize, X.shape[1]))
             X_test = X[k * chunksize :, :]
             z_test = z[k * chunksize :]
         else:

### 2022-10-05 doc/paper/paper.tex
Implementation

@@
 \usepackage{parskip}
+\usepackage{listings}
+\usepackage{xcolor}
 
+\definecolor{codegreen}{rgb}{0,0.6,0}
+\definecolor{codegray}{rgb}{0.5,0.5,0.5}
+\definecolor{codepurple}{rgb}{0.58,0,0.82}
+\definecolor{backcolour}{rgb}{0.95,0.95,0.92}
+
+\lstdefinestyle{mystyle}{
+    backgroundcolor=\color{backcolour},
+    commentstyle=\color{codegreen},
+    keywordstyle=\color{magenta},
+    numberstyle=\tiny\color{codegray},
+    stringstyle=\color{codepurple},
+    basicstyle=\ttfamily\footnotesize,
+    breakatwhitespace=false,
+    breaklines=true,
+    captionpos=b,
+    keepspaces=true,
+    numbers=left,
+    numbersep=5pt,
+    showspaces=false,
+    showstringspaces=false,
+    showtabs=false,
+    tabsize=2
+}
+
+\lstset{breaklines=true, style=mystyle}
 %% The class has several options
@@
 
+With the theory explained, we can go on to describe how we have applied it to our problem.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{Preprocessing: Creating the design matrix and normalizing our data}
-
 We have two datasets, one synthetic and one real.
@@
 
+	Below is the python code used to create the design matrix, which can be found in \texttt{src/utils.py}.
+
+\lstinputlisting[language=Python, firstline=27, lastline=42]{../../src/utils.py}
+
 	While the creating of the design matrix is equivalent for the two datasets, the preprocessing of the data afterwards is not.
@@
 However, we have included a centering of the data, which has no effect on the Ordinary Least Squares regression, but which is used when applying Ridge and Lasso regression, to avoid punishing the intercept.
-The centering is done by
+The centering is done as follows: First, the first coloumn of the design matrix (the intercept coloumn) is discarderd.
+Then we calculate the mean for the target values and the design matrix (the mean for the design matrix is a vector with mean values for each coloumn).
+We subtract this from the design matrix and target values when performing the fit giving us a vector $\bm{\beta}$.
+Afterwards we calculate the intercept with the following equation
+
+\begin{equation}
+	intercept = mean(\bar{y} - \bm{\bar{X}_c}^T \bm{\beta})
+\end{equation}
+
+	where $\bar{y}$ is the mean value of $\bm{y}$ and $\bm{\bar{X}_c}$ is a vector of size $p$ with $\bar{X}_i = mean(X_i)$ (the mean of each coloumn in $X$).
+We finally create our prediction by using the $\bm{\beta}$, $X$ and intercept we calculated above:
+
+\begin{equation}
+	\bm{\hat{y}} = X \bm{\beta} + intercept
+\end{equation}
+
+The code for this, taken from the \texttt{evaluate\_model} function in \texttt{src/utils.py} is as follows: \footnote{Some intermediate parts of our implementation not related to this have been removed for readability}
+
+\begin{lstlisting}[language=Python]
+X_train = X_train[:, 1:]
+z_train_mean = np.mean(z_train, axis=0)
+X_train_mean = np.mean(X_train, axis=0)
+beta = model((X_train - X_train_mean), (z_train - z_train_mean))
+intercept = np.mean(z_train_mean - X_train_mean @ beta)
+z_pred_train = X_train @ beta + intercept
+\end{lstlisting}
+
+Where \texttt{model} calculates \texttt{beta} using the normal equations for OLS.
+
+For the real elevation data, normalization is definitely required, as the values for $x$, $y$ reach several hundred and $z$ reaches over a thousand.
+For polynomial degree $10$ and an $x$ value which reaches $400$, we will have a design matrix with values from $0$ to $400^{10}$.
+This large range of values makes our linear regression perform poorly (WHY?).
+We have therefore chosen to scale our design matrix using Scikit's \texttt{StandardScaler}, which removes the mean value and divides by the variance for each coloumn, giving us a design matrix where each feature coloumn has a mean value of $0$ and variance of $1$.
+The calculation of mean and variance is done on the training data, and applied to both the training and test data.
+We have done the same with the values for $z$.
+This gives a much more stable model.
+The code for doing this on the design matrix is as follows, taken from \texttt{src/utils.py}:
+
+\lstinputlisting[language=Python, firstline=238, lastline=245]{../../src/utils.py}
 

### 2022-10-05 doc/paper/paper.tex

@@
 \end{abstract}
+
+% \tableofcontents
+% \newpage
 
@@
 
-\subsection{Preprocessing: Creating the design matrix and normalizing our data}
+\subsection{Implementation}
+\subsubsection{Preprocessing: Creating the design matrix and normalizing our data}
 We have two datasets, one synthetic and one real.
@@
 
-\subsection{Implementation of linear regression}
+\subsubsection{Implementation of linear regression}
 
-The third-level of heading follows the style of the second-level heading.
+Linear regression for OLS and Ridge are implemented by using NumPy's \texttt{numpy.linalg.pinv} and the normal equations.
+We have implemented them as functions which take in the target values and the design matrix and return $\bm{\hat{\beta}}$.
+They are called from a wrapper function called \texttt{evaluate\_model} which takes care of the scaling and works with both SciKit models and our own.
+Below are the functions, taken from \texttt{src/utils.py}:
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Use of SI Units}
+\lstinputlisting[language=Python, firstline=53, lastline=64]{../../src/utils.py}
 
-An ASME paper should use SI units.
-When preference is given to SI units, the U.S.
-customary units may be given in parentheses or omitted.
-When U.S.
-customary units are given preference, the SI equivalent {\em shall} be provided in parentheses or in a supplementary table.
+For repeated calls of these functions with different polynomial degrees, we use the function \texttt{linreg\_to\_N}, which passes the correctly sliced design matrix to the regression functions and saves the predictions and various performance metrics for each polynomial degree for easy plotting.
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Footnotes\protect\footnotemark}
+For Lasso regression, we use SciKit's implementation. We also use SciKit's implementation for OLS and Ridge for comparisons with our own implementation.
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsubsection{Implementation of resampling methods}
+
+Crossvalidation is implemented in the function \texttt{crossval} in \texttt{src/utils.py}.
+We split the data into k folds by slicing with NumPy taking consecutive rows of the design matrix and target vector.
+But before we do this, we resample the data with Scikit's \texttt{resample}.
+This is important because our $x$'s and $y$'s are not normal distributions but consecutive values between $0$ and $1$, and therefore consecutive rows correspond specific areas of our data, which is not wanted.
+
 \section{Mathematics}

### 2022-10-05 src/task_b.py
make better plots, clean up code

@@ -5,6 +5,7 @@ task b: plots 3D surface plot of Franke Function next to our approximated polyno
         for, and scaling for if we center our data or not. The noise parameter controls how much noise we add to our
         Franke Function output. Note that these plots are without resampling as specified in task b.
 """
+import matplotlib.pyplot as plt
 
 # Our own library of functions
 from utils import *
@@ -19,8 +20,8 @@ z = FrankeFunction(x, y)
 zs = SkrankeFunction(x, y)
 
 # parameters
-betas_to_plot = 5
-noise = 0.05
+betas_to_plot = 9
+noise = 0.1
 N = 20
 scaling = False
 
@@ -93,24 +94,26 @@ plt.show()
 
 # ---------------- PLOTTING GRAPHS --------------
 plt.subplot(221)
+plt.suptitle(f"Datapoints = {len(x)*len(y)}, Parameters: N = {N}, noise = {noise}, scaling = {scaling}", fontsize=6)
 if betas_to_plot <= betas.shape[0]:
     for beta in range(betas_to_plot):
         data = betas[beta, :]
         data[data==0] = np.nan
-        plt.plot(data, label=f"beta{beta}", marker ="o")
-    plt.xlabel("Polynomial degree")
+        plt.plot(data, label=f"beta{beta}", marker="o", markersize=3)
+    plt.xlabel("Polynomial degree (N)")
     plt.ylabel("Beta value")
     plt.legend()
     plt.title("Beta progression")
 
 plt.subplot(222)
 
+print(f"Minimal MSE_test value = {np.min(MSE_test)} for N = {np.argmin(MSE_test)}")
 plt.plot(MSE_train, label="train implementation")
 plt.plot(MSE_test, label="test implementation")
 plt.plot(MSE_train_sk, "r--", label="train ScikitLearn")
 plt.plot(MSE_test_sk, "g--", label="test ScikitLearn")
 plt.ylabel("MSE score")
-plt.xlabel("Polynomial degree")
+plt.xlabel("Polynomial degree (N)")
 plt.ylim(0, 0.1)
 plt.legend()
 plt.title("MSE scores over model complexity")
@@ -121,7 +124,7 @@ plt.plot(R2_test, label="test implementation")
 plt.plot(R2_train_sk, "r--", label="train ScikitLearn")
 plt.plot(R2_test_sk, "g--", label="test ScikitLearn")
 plt.ylabel("R2 score")
[…]

### 2022-10-05 doc/paper/paper.tex
More more

@@
 \begin{abstract}
-{\it This is the abstract.
-This article illustrates preparation of ASME paper using 
-\LaTeX2\raisebox{-.3ex}{$\epsilon$}.
-An abstract for an ASME paper should be less than 150 words and is normally in italics.
-%%% 
-Please use this template to test how your figures will look on the printed journal page of the Journal of Mechanical Design.
-The Journal will no longer publish papers that contain errors in figure resolution.
-These usually consist of unreadable or fuzzy text, and pixilation or rasterization of lines.
-This template identifies the specifications used by JMD some of which may not be easily duplicated; for example, ASME actually uses Helvetica Condensed Bold, but this is not generally available so for the purpose of this exercise Helvetica is adequate.
-However, reproduction of the journal page is not the goal, instead this exercise is to verify the quality of your figures.
-Notice that this abstract is to be set in 9pt Times Italic, single spaced and right justified.
+{\it We use Ordinary Least Squares, Ridge and Lasso regression on a synthetic dataset generated by the Franke Function, and on a real dataset of elevation data from a terrain in Norway.
+We explore different regularization values and polynomial degrees and find that that MODEL performs best, with a MSE of NUMBER for polynomial degree 6.
+We find that linear regression performs well on the synthetic data, but falls short when approximating elevation data with large variances in height.
 }
@@
 The define the optimal prediction, we use a cost function, and the choice of cost function greatly influences what optimal parameters we get.
-FOOTNOTE: It is here important to keep in mind that the cost function used when fitting the model (training) rarely is the same as the one used for testing our prediction, as described in <somewhere>.
+\footnote{It is here important to keep in mind that the cost function used when fitting the model (training) rarely is the same as the one used for testing our prediction, as described in <somewhere>}.
 An intuitive cost function is the mean squared error (MSE).
@@
 \begin{equation}
-	intercept = mean(\bar{y} - \bm{\bar{X}_c}^T \bm{\beta})
+	intercept = mean(\bar{y} - \bar{X}_c^T \bm{\beta})
 \end{equation}
@@
 
-Crossvalidation is implemented in the function \texttt{crossval} in \texttt{src/utils.py}.
-We split the data into k folds by slicing with NumPy taking consecutive rows of the design matrix and target vector.
+We use resampling methods to better estimate our performance metrics.
+We use both cross validation and the bootstrap method.
+Cross validation is implemented in the function \texttt{crossval} in \texttt{src/utils.py}.
+We split the data into $K$ folds by slicing with NumPy taking consecutive rows of the design matrix and target vector.
 But before we do this, we resample the data with Scikit's \texttt{resample}.
 This is important because our $x$'s and $y$'s are not normal distributions but consecutive values between $0$ and $1$, and therefore consecutive rows correspond specific areas of our data, which is not wanted.
+We then fit on $K-1$ folds, and test on the last.
+We return the average error.
+
+We implement bootstrap by repeatedly resampling our data with replacement and performing the prediction.
+We use the predictions returned from hundreds of resamplings to calculate the bias and variance of our model.
+
+\subsubsection{Franke Function}
+
+We use the Franke Function with an added stochastic noise of to test our models. We explore how higher and lower levels of noise affect the model.
+
+\section{Results}
+
+\subsection{Franke Function}
+
+To gain insight into how the three different regression methods work, we experiment with synthetic data generated by the Franke function.
+
+\subsubsection{Ordinary Least Squares}
+
+We find that OLS creates a good fit for the Franke Function.
+We see few signs of overfitting, even for large polynomial degrees.
+This is most likely because the Franke function is a sum of exponents, which can be expanded into polynomials.
+We would expect to see higher polynomial degrees give us better and better prediction, when there is no noise that could give rise to overfitting.
+We find that our own implementation of OLS follows Scikit's implementation up to around polynomial degree DEGREE, at which point scikit's MSE diverges.
+We chalk this up to numerical errors and differing implementation of the normal equations, as we simply use the normal equation exactly, while scikit uses an implementation which is probably computationally more efficient but gives slight differences when comparing to the normal equation.
+See Fig.
+1 for a comparison between the Franke function and our prediction, which shows that they are essentially identical.
+Fig.
+2 shows the train and test MSE by polynomial degree, and shows that we are not overfitting even for polynomial degree 20.
+
+However, when we introduce a stochastic noise, we see that overfitting happens rapidly.
+Without a regularization hyperparameter, we see that the parameters fluctuate wildly and our prediction suffers.
+Fig.
+3 shows our prediction for polynomial degree, where we clearly see that we have overfit the model severly.
+See fig.
+NUMBER in the appendix for test MSE for more and more noisy target values.
+
+By using bootstrap resampling, we can study the bias-variance-tradeoff, which shows how the variance between different bootstraps iterations increases as our model overfits our noisy data.
+See Fig.
+3.
+Clearly, we need to introduce regularization if we want a model that does not overfit.
+
+\subsubsection{Ridge}
+
+Using Ridge regression, we see from Fig.
+NUMBER that larger and larger values of $\lambda$ efficiently prevents overfitting and reduces the variance between predictions.
+Using Scikit's gridsearch, we find the optimal $\lambda$ to be NUMBER and the optimal polynomial degree to be DEGREE, with an average MSE of NUMBER, which is lower than OLS for the Franke function with an added stochastic noise $\epsilon \sim N(0,0.05)$.
+Ridge regression successfully reduces the model complexity and prevents overfitting for higher polynomial degrees.
+
+\subsubsection{Lasso}
+
+Turning our attention to Lasso regression, we also find that overfitting can be avoided with a good choice of $\lambda$.
+Comparing the three methods, we see that for noisy target data, Ridge and Lasso are superior to OLS, as seen in Fig.
+NUMBER.
+
+\subsection{Elevation data}
+
+With the confidence that our methods work as expected on synthetic data, we turn our attention to the real elevation data.
+
+\newpage
 

### 2022-10-05 src/noise_plot.py
add noise_plot.py, clean up code

@@ -1,0 +1,55 @@
+# Our own library of functions
+from utils import *
+
+np.random.seed(42069)
+
+# make data
+x = np.arange(0, 1, 0.05)
+y = np.arange(0, 1, 0.05)
+x, y = np.meshgrid(x, y)
+z = FrankeFunction(x, y)
+zs = SkrankeFunction(x, y)
+
+# parameters
+N = 10
+noises = np.linspace(0, 0.2, 5)
+scaling = False
+bootstraps = 100
+
+def noise_plot(x, y, z, N, noise, scaling, bootstraps):
+    # add noise
+    z += noise * np.random.standard_normal(z.shape)
+    X, X_train, X_test, z_train, z_test = preprocess(x, y, z, N, 0.2)
+
+    # results
+    test_errors = np.zeros(N)
+
+    # bootstrap
+    for n in range(N):
+        print(n)
+        l = int((n + 1) * (n + 2) / 2)  # Number of elements in beta
+        z_preds_test = bootstrap(
+            X[:, :l],
+            X_train[:, :l],
+            X_test[:, :l],
+            z_train,
+            z_test,
+            bootstraps,
+            scaling=scaling,
+        )
+
+        MSE_test, _ = scores(z_test, z_preds_test)
+        test_errors[n] = np.mean(MSE_test)
+
+    return test_errors
+
+for noise in noises:
+    test_errors = noise_plot(x, y, z, N, noise, scaling, bootstraps)
+    # plot
+    plt.plot(test_errors, label=f"test error noise={round(noise)}", marker="o", markersize=3)
+plt.ylim(0, 0.4)
+plt.xlabel("Polynomial degree")
+plt.ylabel("MSE score")
+plt.title("MSE for different levels of noise in model")
+plt.legend()
+plt.show()

### 2022-10-05 src/task_b.py

@@ -78,7 +78,7 @@ ax = fig.add_subplot(122, projection="3d")
 surf = ax.plot_surface(
     x,
     y,
-    np.reshape(z_preds[:, N], z.shape),
+    np.reshape(z_preds[:, np.argmin(MSE_test)], z.shape),
     cmap=cm.coolwarm,
     linewidth=0,
     antialiased=False,
@@ -87,46 +87,46 @@ surf = ax.plot_surface(
 ax.set_zlim(-0.10, 1.40)
 ax.zaxis.set_major_locator(LinearLocator(10))
 ax.zaxis.set_major_formatter(FormatStrFormatter("%.02f"))
-ax.set_title("Polynomial fit of Franke Function")
+ax.set_title(f"Best polynomial fit of Franke Function, N={np.argmin(MSE_test)}")
 fig.colorbar(surf, shrink=0.5, aspect=5)
 
 plt.show()
 
 # ---------------- PLOTTING GRAPHS --------------
-plt.subplot(221)
 plt.suptitle(f"Datapoints = {len(x)*len(y)}, Parameters: N = {N}, noise = {noise}, scaling = {scaling}", fontsize=6)
 if betas_to_plot <= betas.shape[0]:
     for beta in range(betas_to_plot):
         data = betas[beta, :]
         data[data==0] = np.nan
         plt.plot(data, label=f"beta{beta}", marker="o", markersize=3)
-    plt.xlabel("Polynomial degree (N)")
-    plt.ylabel("Beta value")
-    plt.legend()
-    plt.title("Beta progression")
+        plt.xlabel("Polynomial degree (N)")
+        plt.ylabel("Beta value")
+        plt.title("Beta progression")
+        plt.legend()
+    plt.show()
 
-plt.subplot(222)
 
 print(f"Minimal MSE_test value = {np.min(MSE_test)} for N = {np.argmin(MSE_test)}")
-plt.plot(MSE_train, label="train implementation")
-plt.plot(MSE_test, label="test implementation")
-plt.plot(MSE_train_sk, "r--", label="train ScikitLearn")
-plt.plot(MSE_test_sk, "g--", label="test ScikitLearn")
+plt.suptitle(f"Datapoints = {len(x)*len(y)}, Parameters: N = {N}, noise = {noise}, scaling = {scaling}", fontsize=6)
+plt.plot(MSE_train, label="train implementation", marker="o", markersize=3)
[…]

### 2022-10-05 src/task_c.py

@@ -1,41 +1,50 @@
 # Our own library of functions
 from utils import *
 
+np.random.seed(42069)
+
+# make data
 x = np.arange(0, 1, 0.05)
 y = np.arange(0, 1, 0.05)
 x, y = np.meshgrid(x, y)
 z = FrankeFunction(x, y)
-# z = SkrankeFunction(x, y)
-N = 12
+zs = SkrankeFunction(x, y)
+
+# parameters
+N = 20
+noise = 0.05
+scaling = False
 bootstraps = 100
 
-np.random.seed(42069)
-
-z += 0.05 * np.random.standard_normal(z.shape)
+# add noise
+z += noise * np.random.standard_normal(z.shape)
 X, X_train, X_test, z_train, z_test = preprocess(x, y, z, N, 0.2)
 
+# results
 errors = np.zeros(N)
 biases = np.zeros(N)
 variances = np.zeros(N)
 
+# bootstrap
 for n in range(N):
     print(n)
     l = int((n + 1) * (n + 2) / 2)  # Number of elements in beta
-    z_preds = bootstrap(
+    z_preds_test = bootstrap(
         X[:, :l],
         X_train[:, :l],
         X_test[:, :l],
         z_train,
         z_test,
         bootstraps,
-        scaling=True,
+        scaling=scaling,
     )
 
-    error, bias, variance = bias_variance(z_test, z_preds)
+    error, bias, variance = bias_variance(z_test, z_preds_test)
     errors[n] = error
     biases[n] = bias
     variances[n] = variance
 
+# plot
 print(f"Minimal MSE_test value = {np.min(errors)} for N = {np.argmin(errors)}")
 plt.plot(errors, label="error")
 plt.plot(biases, label="biases")

### 2022-10-05 src/utils.py

@@ -79,13 +79,15 @@ def bootstrap(
     model: Callable = OLS,
     lam: float = 0,
 ):
+    z_preds_train = np.empty((z_train.shape[0], bootstraps))
     z_preds_test = np.empty((z_test.shape[0], bootstraps))
 
     for i in range(bootstraps):
         X_, z_ = resample(X_train, z_train)
-        _, _, z_pred_test, _ = evaluate_model(
+        _, z_pred_train, z_pred_test, _ = evaluate_model(
             X, X_, X_test, z_, model, lam=lam, scaling=scaling
         )
+        z_preds_train[:, i] = z_pred_train
         z_preds_test[:, i] = z_pred_test
 
     return z_preds_test
@@ -139,11 +141,11 @@ def crossval(
     return np.mean(errors)
 
 
-def bias_variance(z_test: np.ndarray, z_preds: np.ndarray):
-    MSEs, _ = scores(z_test, z_preds)
+def bias_variance(z_test: np.ndarray, z_preds_test: np.ndarray):
+    MSEs, _ = scores(z_test, z_preds_test)
     error = np.mean(MSEs)
-    bias = np.mean((z_test - np.mean(z_preds, axis=1, keepdims=True).flatten()) ** 2)
-    variance = np.mean(np.var(z_preds, axis=1, keepdims=True))
+    bias = np.mean((z_test - np.mean(z_preds_test, axis=1, keepdims=True).flatten()) ** 2)
+    variance = np.mean(np.var(z_preds_test, axis=1, keepdims=True))
 
     return error, bias, variance
 

### 2022-10-05 doc/paper/paper.tex

@@
 With the confidence that our methods work as expected on synthetic data, we turn our attention to the real elevation data.
+Here we find that all forms of regression struggle to model the data, which shows a limitation of linear regression.
+The data is simply to varied and noisy to be effectively modelled by a polynomial of low degree, and larger degrees are so computationally intensive to calculate that it is not possible to reach the degrees needed to approximate the terrain.
 

### 2022-10-06 doc/paper/paper.tex
More paper

--- 2022-10-05 doc/paper/paper.tex
@@
 With the theory explained, we can go on to describe how we have applied it to our problem.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 This large range of values makes our linear regression perform poorly (WHY?).
-We have therefore chosen to scale our design matrix using Scikit's \texttt{StandardScaler}, which removes the mean value and divides by the variance for each coloumn, giving us a design matrix where each feature coloumn has a mean value of $0$ and variance of $1$.
+We have therefore chosen to scale our design matrix using Scikit's \texttt{MinMaxScaler}, which scales each column to be in the range of $(0,1)$.
 The calculation of mean and variance is done on the training data, and applied to both the training and test data.
@@
 \lstinputlisting[language=Python, firstline=238, lastline=245]{../../src/utils.py}
+
+Another form of "preprocessing" which is strictly practical in nature, is that we have reduced the dimensions of the data to 320 by 640.
+This was done by downscaling the image.
+\footnote{The interpolation was done using "nearest neighbor interpolation", which does not introduce new values into the data.
+An equivalent programmatic solution would be to for instance only look at every 4 pixels} This was simply done because the design matrix becomes several gigabytes in size if one uses the whole dataset, and model fitting takes an unreasonable amount of time.
 
@@
 
+\subsubsection{Ordinary Least Squares}
+
+Here we find some limitations of linear regression.
+We see from Fig.
+NUMBER that the we are able to fit the terrain up to a certain point, gaining an MSE of NUMBER with OLS with polynomial degree DEGREE.
+However, we see that beyond this, we have little to gain.
+This can be seen when we increase the polynomial degree to very large amounts, without seeing any decrease in MSE (Fig.
+NUMBER).
+Looking at the beta progression for selected betas (Fig.
+NUMBER), we see that the parameters stabilize, which implies that we are not overfitting, but simply reaching our best possible prediction for polynomial regression.
+The fact that we are not overfitting is further reinforced by looking at the bias-variance plot in Fig.
+NUMBER, which shows no increase in variance.
+It is reasonable to expect that our model will not overfit, due to the large increase in data we now use for the fit.
+Our synthetic data used $20 x 20 = 400$ datapoints, while we now have $320 x 640 = 204 800$ datapoints.
+This means that for a 50th order polynomial, which will have $1326$ features, our number of features are less than $1\%$ of our datapoints.
+If we decrease the number of datapoints used for training considerably, we can see overfitting start to occur (see Appendix 1).
+So we are not overfitting, but we are not decreasing our MSE either.
+This is likely because the elevation data has a jagged and complex shape, which cannot be accurately recreated, even with a polynomial of degree $50$.
+We should remember that the Franke function can be expanded into a series of polynomials, which is why we can decrease our MSE as much as we like by increasing our model complexity (as long as our data is not noisy).
+The same is not true for our elevation data, and as such we have little to gain from increasing our model complexity after a certain point.
+Clearly, linear regression does not offer a very good tradeoff between computational cost and performance for this data.
+
 \newpage
-
-\section{Mathematics}
-
-Equations should be numbered consecutively beginning with (1) to the end of the paper, including any appendices.
-The number should be enclosed in parentheses and set flush right in the column on the same line as the equation.
-An extra line of space should be left above and below a displayed equation or formula.
-\LaTeX\ can automatically keep track of equation numbers in the paper and format almost any equation imaginable.
-An example is shown in Eqn.~(\ref{eq_ASME}).
-The number of a referenced equation in the text should be preceded by Eqn.\ unless the reference starts a sentence in which case Eqn.\ should be expanded to Equation.
-
-\begin{equation}
-f(t) = \int_{0_+}^t F(t) dt + \frac{d g(t)}{d t}
-\label{eq_ASME}
-\end{equation}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Citing References}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-The ASME reference format is defined in the authors kit provided by the ASME.  The format is:
-
-\begin{quotation}
-{\em Text Citation}. Within the text, references should be cited in  numerical order according to their order of appearance.  The numbered reference citation should be enclosed in brackets.
-\end{quotation}
-
-The references must appear in the paper in the order that they were cited.
-In addition, multiple citations (3 or more in the same brackets) must appear as a `` [1-3]''.
-A complete definition of the ASME reference format can be found in the  ASME manual \cite{asmemanual}.
-
-The bibliography style required by the ASME is unsorted with entries appearing in the order in which the citations appear.
-If that were the only specification, the standard {\sc Bib}\TeX\ unsrt bibliography style could be used.
-Unfortunately, the bibliography style required by the ASME has additional requirements (last name followed by first name, periodical volume in boldface, periodical number inside parentheses, etc.) that are not part of the unsrt style.
-Therefore, to get ASME bibliography formatting, you must use the \verb+asmems4.bst+ bibliography style file with {\sc Bib}\TeX.
-This file is not part of the standard BibTeX distribution so you'll need to place the file someplace where LaTeX can find it (one possibility is in the same location as the file being typeset).
-
-With \LaTeX/{\sc Bib}\TeX, \LaTeX\ uses the citation format set by the class file and writes the citation information into the .aux file associated with the \LaTeX\ source.
-{\sc Bib}\TeX\ reads the .aux file and matches the citations to the entries in the bibliographic data base file specified in the \LaTeX\ source file by the \verb+\bibliography+ command.
-{\sc Bib}\TeX\ then writes the bibliography in accordance with the rules in the bibliography .bst style file to a .bbl file which \LaTeX\ merges with the source text.
-A good description of the use of {\sc Bib}\TeX\ can be found in \cite{latex, goosens} (see how two references are handled?).
-The following is an example of how three or more references \cite{latex, asmemanual,  goosens} show up using the \verb+asmems4.bst+ bibliography style file in conjunction with the \verb+asme2ej.cls+ class file.
-Here are some more \cite{art, blt, ibk, icn, ips, mts, mis, pro, pts, trt, upd} which can be used to describe almost any sort of reference.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusions}
-The only way to ensure that your figures are presented in the ASME Journal of Mechanical Design in the way you feel is appropriate and meets the requirement for quality presentation is for you to prepare a double column version of the paper in a form similar to that used by the Journal.
-
-This gives you the opportunity to ensure that the figures are sized appropriately, in particular that the labels are readable and match the size of the text in the journal, and that the line weights and resolutions have no pixilation or rasterization.
-Poor quality figures are immediately obvious on the printed page, and this detracts from the perceived quality of the journal.
-
-I am pleased to provide advice on how to improve any figure, but this effort must start with a two-column version of the manuscript.
-Thank you in advance for your patience with this effort, it will ensure quality presentation of your research contributions.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Discussions}
-This template is not yet ASME journal paper format compliant at this point.
-More specifically, the following features are not ASME format compliant.
-\begin{enumerate}
-\item
-The format for the title, author, and abstract in the cover page.
-\item
-The font for title should be 24 pt Helvetica bold.
-\end{enumerate}
-
-\noindent
-If you can help to fix these problems, please send us an updated template.
-If you know there is any other non-compliant item, please let us know.
-We will add it to the above list.
-With your help, we shall make this template 
-compliant to the ASME journal paper format.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{acknowledgment}
-ASME Technical Publications provided the format specifications for the Journal of Mechanical Design, though they are not easy to reproduce.
-It is their commitment to ensuring quality figures in every issue of JMD that motivates this effort to have authors review the presentation of their figures.
-
-Thanks go to D.
-E.
-Knuth and L.
-Lamport for developing the wonderful word processing software packages \TeX\ and \LaTeX.
-We would like to thank Ken Sprott, Kirk van Katwyk, and Matt Campbell for fixing bugs in the ASME style file \verb+asme2ej.cls+, and Geoff Shiflett for creating
-ASME bibliography stype file \verb+asmems4.bst+.
-\end{acknowledgment}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% The bibliography is stored in an external database file
-% in the BibTeX format (file_name.bib).  The bibliography is
-% created by the following command and it will appear in this
-% position in the document. You may, of course, create your
-% own bibliography by using thebibliography environment as in
-%
-% \begin{thebibliography}{12}
-% ...
-% \bibitem{itemreference} D. E. Knudsen.
-% {\em 1966 World Bnus Almanac.}
-% {Permafrost Press, Novosibirsk.}
-% ...
-% \end{thebibliography}
-
-% Here's where you specify the bibliography style file.
-% The full file name for the bibliography style file 
-% used for an ASME paper is asmems4.bst.
-\bibliographystyle{asmems4}
-
-% Here's where you specify the bibliography database file.
-% The full file name of the bibliography database for this
-% article is asme2e.bib. The name for your database is up
-% to you.
-\bibliography{asme2e}
+\bibliographystyle{alpha}
+\bibliography{bibliography}
 

### 2022-10-06 src/g_task_b.py
new data

@@ -1,9 +1,7 @@
 """
-task g: plot terrain, approximate terrain with OLS and calculate MSE, R2
-        beta over model complexity for real data. Performs task_b, so no
-        resampling.
+task b (and task g): plot terrain, approximate terrain with OLS and calculate MSE, R2 beta over model complexity for
+                     real data. Performs task_b, so no resampling.
 """
-import numpy as np
 from imageio import imread
 import sys
 import argparse
@@ -19,160 +17,153 @@ parser = argparse.ArgumentParser(
     description="Compute task g.b"
 )
 
-# filename is positional and required
-parser.add_argument("file", help="The filename to apply filter to")
+# filename is optional
+parser.add_argument("-f", "--file", help="The filename to apply filter to")
 
 # parse arguments and call run_filter
 args = parser.parse_args()
 
-# Load the terrain
-z_terrain1 = np.asarray(imread(args.file), dtype="float64")
-x_terrain1 = np.arange(z_terrain1.shape[0])
-y_terrain1 = np.arange(z_terrain1.shape[1])
-x1, y1 = np.meshgrid(x_terrain1, y_terrain1, indexing="ij")
-
-z_terrain2 = np.asarray(imread("../data/SRTM_data_Norway_2.tif"), dtype="float64")
-x_terrain2 = np.arange(z_terrain2.shape[0])
-y_terrain2 = np.arange(z_terrain2.shape[1])
-x2, y2 = np.meshgrid(x_terrain2, y_terrain2, indexing="ij")
-
-# Show the terrain
-plt.plot()
-plt.subplot(121)
-plt.title("Terrain over Norway 1")
-plt.imshow(z_terrain1)
-plt.xlabel("Y")
-plt.ylabel("X")
-
-plt.subplot(122)
-plt.title("Terrain over Norway 2")
-plt.imshow(z_terrain2)
-plt.xlabel("Y")
-plt.ylabel("X")
-plt.show()
-
-# Highest order polynomial we fit with
+# parameters
 N = 20
-betas_to_plot = 7
+betas_to_plot = 9
+noise = 0.05
+scaling = False
 
-def task_b(x, y, z, N):
+# implemented model under testing
+OLS_model = OLS
+# scikit model under testing
+OLS_scikit = LinearRegression(fit_intercept=False)
+
+if args.file:
+    # Load the terrain
[…]

### 2022-10-06 src/g_task_e_msecomparison.py

@@ -35,7 +35,7 @@ def ridge_mse_compare_resampling(x, y, z, N, K, bootstraps):
     X, X_train, X_test, z_train, z_test = preprocess(x, y, z, N, 0.2)
 
     # normalize data based on train
-    X, X_train, X_test, z, z_train, z_test = normalize_task_g(
+    X, X_train, X_test, z, z_train, z_test = minmax_dataset(
         X, X_train, X_test, z, z_train, z_test
     )
 

### 2022-10-06 src/g_task_f_biasvariance.py

@@ -31,7 +31,7 @@ lambdas = np.logspace(-6, -1, 6)
 def lasso_bias_variance(x, y, z, N, bootstraps, lambdas):
     X, X_train, X_test, z_train, z_test = preprocess(x, y, z, N, 0.2)
 
-    X, X_train, X_test, z, z_train, z_test = normalize_task_g(
+    X, X_train, X_test, z, z_train, z_test = minmax_dataset(
          X, X_train, X_test, z, z_train, z_test
      )
 
@@ -41,7 +41,7 @@ def lasso_bias_variance(x, y, z, N, bootstraps, lambdas):
         plt.suptitle(f"Bias variance tradeoff for lasso regression")
 
         # model under testing
-        model_Lasso = Lasso(lambdas[i], max_iter=100, fit_intercept=False)
+        model_Lasso = Lasso(lambdas[i], max_iter=200, fit_intercept=False)
 
         # arrays for bias-variance
         errors = np.zeros(N)

### 2022-10-06 src/utils.py

@@ -235,7 +235,7 @@ def evaluate_model(
     return beta, z_pred_train, z_pred_test, z_pred
 
 
-def normalize_task_g(X, X_train, X_test, z, z_train, z_test):
+def minmax_dataset(X, X_train, X_test, z, z_train, z_test):
     x_scaler = MinMaxScaler()
     z_scaler = MinMaxScaler()
 

### 2022-10-06 doc/paper/paper.tex
Paper stuff

@@
 
+%%%%%%% FIGURES %%%%%%%%%
+\begin{figure*} 
+\centerline{\includegraphics[width=6.85in]{figure/frankenonoise.png}}
+\caption{A comparison the our prediction and the Franke function with no noise}
+\label{frankenonoise}
+\end{figure*}
+
+\begin{figure} 
+\centerline{\includegraphics[width=3.25in]{figure/frankenonoisemse.png}}
+\caption{Test and train MSE for the Franke function with no noise}
+\label{frankenonoisemse}
+\end{figure}
+
+\begin{figure} 
+\centerline{\includegraphics[width=3.25in]{figure/frankenonoiser2.png}}
+\caption{Test and train R2-score for the Franke function with no noise}
+\label{frankenonoiser2}
+\end{figure}
+
+\begin{figure} 
+\centerline{\includegraphics[width=3.25in]{figure/frankenoisemse.png}}
+\caption{Test and train MSE for the Franke function with noise}
+\label{frankenoisemse}
+\end{figure}
+
+\begin{figure} 
+\centerline{\includegraphics[width=3.25in]{figure/frankenoiser2.png}}
+\caption{Test and train R2-score for the Franke function no noise}
+\label{frankenoiser2}
+\end{figure}
+
 We find that OLS creates a good fit for the Franke Function.
@@
 We would expect to see higher polynomial degrees give us better and better prediction, when there is no noise that could give rise to overfitting.
-We find that our own implementation of OLS follows Scikit's implementation up to around polynomial degree DEGREE, at which point scikit's MSE diverges.
-We chalk this up to numerical errors and differing implementation of the normal equations, as we simply use the normal equation exactly, while scikit uses an implementation which is probably computationally more efficient but gives slight differences when comparing to the normal equation.
-See Fig.
-1 for a comparison between the Franke function and our prediction, which shows that they are essentially identical.
-Fig.
-2 shows the train and test MSE by polynomial degree, and shows that we are not overfitting even for polynomial degree 20.
+See Fig.~\ref{frankenonoise} for a comparison between the Franke function and our prediction, which shows that they are essentially identical.
+Fig.~\ref{frankenonoisemse} and Fig.~\ref{frankenonoiser2} shows the train and test MSE and R2 by polynomial degree, and shows that we are not overfitting even for polynomial degree 25.
+The MSE for test in these plots is the average MSE for 100 bootstraps.
 
@@
 Without a regularization hyperparameter, we see that the parameters fluctuate wildly and our prediction suffers.
-Fig.
-3 shows our prediction for polynomial degree, where we clearly see that we have overfit the model severly.
-See fig.
-NUMBER in the appendix for test MSE for more and more noisy target values.
+Fig.~\ref{frankenoisemse} and Fig.~\ref{frankenoiser2} shows our MSE and R2 up to polynomial degree 25, where we see that we are overfitting severely.
+See Fig.~\ref{frankienoisecomparison} in the appendix for test MSE for more and more noisy target values.
+Here we also see something a bit surprising: We see that our own implementation of OLS follows Scikit's implementation up to around polynomial degree DEGREE, at which point scikit's MSE increases rapidly.
+We chalk this up to numerical errors and differing implementation of the normal equations, as we simply use the normal equation exactly, while scikit uses an implementation which is probably computationally more efficient but gives slight differences when comparing to the normal equation.
+These differences seem to create a large error when the model is overfitted, which we can not adequately explain.
 
@@
 Clearly, linear regression does not offer a very good tradeoff between computational cost and performance for this data.
+
+\subsubsection{Ridge Regression}
+
+As our model is not overfitting, we do not expect Ridge to offer much improvement.
 
@@
 \appendix       %%% starting appendix
-\section*{Appendix A: Head of First Appendix}
-Avoid Appendices if possible.
+\section*{Appendix A: Plots}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Appendix B: Head of Second Appendix}
-\subsection*{Subsection head in appendix}
-The equation counter is not reset in an appendix and the numbers will
-follow one continual sequence from the beginning of the article to the very end as shown in the following example.
-\begin{equation}
-a = b + c.
-\end{equation}
+\begin{figure} 
+\centerline{\includegraphics[width=3.25in]{figure/frankienoisecomparison.png}}
+\caption{Test and train MSE for the Franke function with no noise}
+\label{frankienoisecomparison}
+\end{figure}
 

### 2022-10-06 src/g_task_b.py
Add better plotting for some files

@@ -5,6 +5,7 @@ task b (and task g): plot terrain, approximate terrain with OLS and calculate MS
 from imageio import imread
 import sys
 import argparse
+import matplotlib
 
 # Our own library of functions
 from utils import *
@@ -13,9 +14,7 @@ np.random.seed(42069)
 
 argv = sys.argv[1:]
 
-parser = argparse.ArgumentParser(
-    description="Compute task g.b"
-)
+parser = argparse.ArgumentParser(description="Compute task g.b")
 
 # filename is optional
 parser.add_argument("-f", "--file", help="The filename to apply filter to")
@@ -24,7 +23,7 @@ parser.add_argument("-f", "--file", help="The filename to apply filter to")
 args = parser.parse_args()
 
 # parameters
-N = 20
+N = 25
 betas_to_plot = 9
 noise = 0.05
 scaling = False
@@ -91,6 +90,11 @@ MSE_test_sk, R2_test_sk = scores(z_test, z_preds_test_sk)
 pred_map = z_preds[:, -1].reshape(z.shape)
 
 # ------------ PLOTTING 3D -----------------------
+
+font = {"family": "normal", "size": 25}
+
+matplotlib.rc("font", **font)
+
 fig = plt.figure(figsize=plt.figaspect(0.3))
 
 # Subplot for terrain
@@ -137,7 +141,7 @@ plt.show()
 if betas_to_plot <= betas.shape[0]:
     for beta in range(betas_to_plot):
         data = betas[beta, :]
-        data[data==0] = np.nan
+        data[data == 0] = np.nan
         plt.plot(data, label=f"beta{beta}", marker="o", markersize=3)
         plt.xlabel("Polynomial degree (N)")
         plt.ylabel("Beta value")
@@ -147,21 +151,40 @@ if betas_to_plot <= betas.shape[0]:
 
 print(f"Minimal MSE_test value = {np.min(MSE_test)} for N = {np.argmin(MSE_test)}")
 #    plt.suptitle(f"Datapoints = {len(x)*len(y)}, Parameters: N = {N}, noise = {noise}, scaling = {scaling}", fontsize=6)
-plt.plot(MSE_train, label="train implementation", marker="o", markersize=3)
-plt.plot(MSE_test, label="test implementation", marker="o", markersize=3)
-plt.plot(MSE_train_sk, "r--", label="train ScikitLearn", marker="o", markersize=3)
[…]

### 2022-10-06 src/noise_plot.py

@@ -3,6 +3,11 @@ own idea: make a plot MSE test error for different levels of noise
 """
 # Our own library of functions
 from utils import *
+import matplotlib
+
+font = {"family": "normal", "size": 25}
+
+matplotlib.rc("font", **font)
 
 np.random.seed(42069)
 
@@ -14,10 +19,11 @@ z = FrankeFunction(x, y)
 zs = SkrankeFunction(x, y)
 
 # parameters
-N = 20
+N = 10
 noises = np.linspace(0, 0.2, 5)
 scaling = False
 bootstraps = 100
+
 
 def noise_plot(x, y, z, N, noise, scaling, bootstraps):
     # add noise
@@ -50,7 +56,13 @@ def noise_plot(x, y, z, N, noise, scaling, bootstraps):
 for noise in noises:
     test_errors = noise_plot(x, y, z, N, noise, scaling, bootstraps)
     # plot
-    plt.plot(test_errors, label=f"test error noise={round(noise, 2)}", marker="o", markersize=3)
+    plt.plot(
+        test_errors,
+        label=f"test error noise={round(noise, 2)}",
+        marker="o",
+        markersize=4,
+        linewidth=3,
+    )
 plt.ylim(0, 0.3)
 plt.xlabel("Polynomial degree (N)")
 plt.ylabel("MSE score")

### 2022-10-06 src/task_c.py

@@ -3,6 +3,7 @@ task c: bootstrapped bias-variance trade-off plot for OLS on synthetic data
 """
 # Our own library of functions
 from utils import *
+import matplotlib
 
 np.random.seed(42069)
 
@@ -14,23 +15,28 @@ z = FrankeFunction(x, y)
 zs = SkrankeFunction(x, y)
 
 # parameters
-N = 20
+N = 25
 noise = 0.05
-scaling = False
+scaling = True
 bootstraps = 100
 
 # model under testing
 model = OLS
 
 # add noise
-z += noise * np.random.standard_normal(z.shape)
+# z += noise * np.random.standard_normal(z.shape)
 X, X_train, X_test, z_train, z_test = preprocess(x, y, z, N, 0.2)
 
 # results
-errors = np.zeros(N)
+MSEs_test = np.zeros(N)
+MSEs_train = np.zeros(N)
+MSEs_test_sk = np.zeros(N)
+MSEs_train_sk = np.zeros(N)
+
 biases = np.zeros(N)
 variances = np.zeros(N)
 
+ols_scikit = LinearRegression(fit_intercept=False)
 # bootstrap
 for n in range(N):
     print(n)
@@ -46,18 +52,73 @@ for n in range(N):
         model=model,
     )
 
-    error, bias, variance = bias_variance(z_test, z_preds_test)
-    errors[n] = error
+    z_preds_test_sk = bootstrap(
+        X[:, :l],
+        X_train[:, :l],
+        X_test[:, :l],
+        z_train,
+        z_test,
+        bootstraps,
+        scaling=scaling,
+        model=ols_scikit,
+    )
+
+    _, z_pred_train, _, _ = evaluate_model(
+        X[:, :l], X_train[:, :l], X_test[:, :l], z_train, model
+    )
+
+    _, z_pred_train_sk, _, _ = evaluate_model(
+        X[:, :l], X_train[:, :l], X_test[:, :l], z_train, ols_scikit
+    )
+
+    MSE_test, bias, variance = bias_variance(z_test, z_preds_test)
+    MSE_train = MSE(z_pred_train, z_train)
+
+    MSE_test_sk, _, _ = bias_variance(z_test, z_preds_test_sk)
+    MSE_train_sk = MSE(z_pred_train_sk, z_train)
+
+    MSEs_test[n] = MSE_test
+    MSEs_test_sk[n] = MSE_test_sk
+
+    MSEs_train[n] = MSE_train
+    MSEs_train_sk[n] = MSE_train_sk
+
[…]

### 2022-10-06 src/utils.py

@@ -144,7 +144,9 @@ def crossval(
 def bias_variance(z_test: np.ndarray, z_preds_test: np.ndarray):
     MSEs, _ = scores(z_test, z_preds_test)
     error = np.mean(MSEs)
-    bias = np.mean((z_test - np.mean(z_preds_test, axis=1, keepdims=True).flatten()) ** 2)
+    bias = np.mean(
+        (z_test - np.mean(z_preds_test, axis=1, keepdims=True).flatten()) ** 2
+    )
     variance = np.mean(np.var(z_preds_test, axis=1, keepdims=True))
 
     return error, bias, variance

### 2022-10-06 doc/paper/paper.tex
More paper

@@
 This leads us to the two next forms of linear regression we use in this paper.
+But before this, a...
+
+\subsubsection{Brief statistical intermission}
+
+We must remember that our target values are defined as $y_i = f + \epsilon_i$, that is, they are distorted by a stochastic noise $\epsilon_i \sim N(0,\sigma^2)$.
+We should therefore take a brief intermission to prove some of the statistical qualities of the method we have just described, so as to reassure ourselves that we are justified in applying linear regression to noisy data.
+We will start by showing that
+
+\begin{equation}
+E[y_i] = \sum_j X_{ij}\beta_j
+\end{equation}
+
+The proof is as follows:
+
+$\bm{y}$ is defined as the $f(\bm{x}) + \bm{\epsilon}$.
+We make an assumption that this can be approximated as a matrix multiplication $X\bm{\beta}$ plus an error vector $\bm{\epsilon}$.
+This means that each element of the vector
+$\bm{y}$ can be expressed as follows:
+
+\begin{equation}
+y_i = \sum_j x_{ij}\beta_j + \epsilon_i
+\end{equation}
+
+If we take the expectation value of this expression we get the following:
+
+\begin{gather}
+E[y_i] = E\left[\sum_j x_{ij}\beta_j + \epsilon_i\right] \\
+E[y_i] = E\left[\sum_j x_{ij}\beta_j\right] + E[\epsilon_i]
+\end{gather}
+
+However, the elements of $X$ are not stochastic, and neither are the elements of $\beta$ and therefore the first expectation value is simply the sum itself.
+Furthermore, $\epsilon$ is explicitly defined as a normal distribution $N(0,\sigma^2)$, and will by definition have the expectation value $0$.
+Therefore, we end up with the final expression:
+
+\begin{equation}
+E[y_i] = \sum_j x_{ij}\beta_j = \bm{X}_{i,*}\bm{\beta}
+\end{equation}
+
+Next, we show that the variance of $y_i$ is equal to the variance of the noise:
+
+\begin{gather}
+var[y_i] = E\left[(y_i - E[y_i])^2\right]  \\
+= E\left[y_i^2 - 2E[y_i]y_i + E[y_i]^2\right]
+\end{gather}
+
+Distributing the outer expectation value function:
+
+\begin{gather}
+var[y_i] = E[y_i^2] - 2E\left[E[y_i]\right]E[y_i] + E\left[E[y]^2\right]  \\
+= E[y_i^2] - 2E\left[E[y_i]\right]E[y_i] + E\left[E[y]\right]E\left[E[y]\right]
+\end{gather}
+
+The result of calculating the expectation value is non-stochastic. This means that $E[E[X]] = E[X]$. From this it follows that
+
+\begin{gather}
+var[y_i] = E[y_i^2] - 2E[y_i]E[y_i] + E[y_i]E[y_i]  \\
+= E[y_i^2] - E[y_i]E[y_i]  \\
+= E[y_i^2] - E[y_i]^2
+\end{gather}
+
+The expectation value in the second summand has been proven to be equal to $\bm{X}_{i,*}\bm{\beta}$ above. We therefore have
+
+\begin{gather}
+var[y_i] = E[y_i^2] - (\bm{X}_{i,*}\bm{\beta})^2  \\
+= E\left[(X_{i,*}\bm{\beta})^2 + X_{i,*}\bm{\beta}\epsilon_i + \epsilon_i^2\right] - (\bm{X}_{i,*}\bm{\beta})^2  \\
+= E\left[(X_{i,*}\bm{\beta})^2\right] + E[X_{i,*}\bm{\beta}\epsilon_i] + E[\epsilon_i^2] - (\bm{X}_{i,*}\bm{\beta})^2
+\end{gather}
+
+$X_{i,*}\bm{\beta}$ and $\epsilon_i$ are both scalars.
+Therefore the expectation value can be written as $E[X_{i,*}\bm{\beta}]E[\epsilon_i]$.
+However, $E[\epsilon_i]$ is by definition $0$, because $\epsilon_i$ is defined as a normal distribution of mean $0$ and variance $\sigma^2$.
+Furthermore, the expectation value of the non-stochastic $(X_{i,*}\bm{\beta})^2$ is simply the expression itself.
+We therefore have
+
+\begin{gather}
+var[y_i] = (X_{i,*}\bm{\beta})^2 + E[X_{i,*}\bm{\beta}]\cdot 0 + E[\epsilon_i^2] - (\bm{X}_{i,*}\bm{\beta})^2  \\
+= E[\epsilon_i^2]
+\end{gather}
+
+We can prove that this is equal to the variance of $\epsilon_i$:
+
+\begin{gather}
+E[\epsilon_i^2] = \frac{1}{n} \sum_i \epsilon_i^2 \\
+var[\epsilon_i] = \frac{1}{n} \sum_i (\epsilon_i - \bar{\epsilon_i})^2 \\
+var[\epsilon_i] = \frac{1}{n} \sum_i (\epsilon_i - 0)^2 \\
+var[\epsilon_i] = \frac{1}{n} \sum_i \epsilon_i^2 = E[\epsilon_i^2]
+\end{gather}
+
+And of course, we know that the variance of $\epsilon_i$ by definition is $\sigma^2$.
+
+\begin{gather}
+var[y_i] = E[\epsilon_i^2] = var[\epsilon_i] = \sigma^2 
+\end{gather}
+
+We should also prove some qualities about or parameters. We start by proving that $E[\hat{\beta}] = \beta$
+
+Because we are assuming $(X^{T}X)^{-1}X^{T}$ to be deterministic, $E[(X^{T}X)^{-1}X^{T}] = (X^{T}X)^{-1}X^{T}$. Moreover, we know $E[y] = y$, and we can thus rewrite the equation as follows:
+
+\begin{equation}
+E[(X^{T}X)^{-1}X^{T}y] = (X^{T}X)^{-1}X^{T}y = \beta
+\end{equation}
+
+Next, we prove that $var(\beta) = \sigma^{2}(X^{T}X)^{-1}$.
+
+$$var(\beta) = var((X^{T}X)^{-1}X^{T}y)$$
+
+Because we are assuming $(X^{T}X)^{-1}X^{T}$ to be deterministic and $y$ to be stochastic, we can rewrite the $var(\beta)$ as following:
+
+\begin{gather}
+var(\beta) = (X^{T}X)^{-1}X^{T}var(y)((X^{T}X)^{-1}X^{T})^{T} \\
+= (X^{T}X)^{-1}X^{T}var(y)(X^{T})^{T}((X^{T}X)^{-1})^{T}
+\end{gather}
+
+Since $(X^{T})^{T} = X$ and $(X^{T}X)^{-1})^{T}$ = $(X^{T}X)^{T})^{-1} = (X^{T}X)^{-1}$
+
+\begin{gather}
+= (X^{T}X)^{-1}X^{T}var(y)X(X^{T}X)^{-1}
+\end{gather}
+
+We know var(y) = $\sigma^{2}$, and since it is a scalar it is commutative. Thus we may move it freely
+
+\begin{gather}
+= \sigma^{2}(X^{T}X)^{-1}X^{T}X(X^{T}X)^{-1} \\
+= \sigma^{2}(X^{T}X)^{-1}
+\end{gather}
 
@@
 Like Ridge regression, this form of regression controls the complexity of the model, with the added benifit that Lasso can drive parameters to exactly zero.
+
+\subsubsection{Resampling methods: The bootstrap}
+
+Bootstrap is a resampling method divided into simple steps.
+First, we reshuffle our dataset with replacement, which will provide us with a new dataset.
+Note that we will already have split our data into train and test data before we begin the bootstrap, so when we reshuffle, we will only reshuffle the training data.
+The reshuffled datasets act as our samples.
+
+Secondly, we will perform our linear regression on this new dataset and store the predictions of the test data.
+Thirdly, we repeat for a given large number of iterations.
+Since we use all the test predictions to calculate the MSE, bias and variance, we will then according to the central limit theorem <reference?> obtain a normal distribution of these variables.
+The law of large numbers <reference?> then tells us that the mean sample values of the MSE, bias and variance approach the true value of MSE, bias and variance.
+This is dependent on our assumption that our data is independently and identically distributed.
+
+These approximations of the true value of MSE, bias and variance will become useful when performing our bias-variance trade-off analysis.
+
+\subsubsection{Cross validation}
+
+Cross validation is another resampling technique, in which we reshuffle our data and then split them into k-folds, where k specifies the number of folds.
+One fold is used as our test data, and the rest is used as training data.
+We run our linear regression, then switch up which fold is used for testing until all folds have been used.
+As in bootstrapping, we will approach the true values of for example the mean squared error for our model.
+Theoretically, because both two resampling techniques use the central limit theorem and the law of large numbers, they should return the same values for the MSE, bias and variance given the same dataset.
+Though resampling techniques can be computationally expensive, they make up for it by aiding in model evaluation by supplying us with variables obtained from a larger number of samples.
+
+\subsubsection{Bias-variance trade-off}
+
+A crucial part of model evaluation is ensuring that the model is neither underfitted nor overfitted, which we can determine by observing the relationship between the bias and the variance.
+Bias is defined as $(Bias[\tilde{y}])^2 = (y-E[\tilde{y}])^2$, the difference between the expected value of our prediction and the observed value.
+Models with high bias are usually underfitted, whilst models with low bias are the opposite.
+Variance on the other hand is defined as $var[\tilde{f}] = \frac{1}{n}\sum(\tilde{y_i}-E[\tilde{y}])^2$, how far away predictions are from the mean prediction.
+Models with high variance are usually overfitted to the training data, whilst low variance models are the opposite.
+A good rule of thumb for finding a good fit is looking at the intercept between bias and variance.
+This is because the cost function <PROOF task c>.
+Because the cost function is a sum of the bias and the variance, the lowest error will generally occur when both bias and variance are low.
+This also fits our understanding of the definitions of bias and variance, as we wish to obtain a model that is neither high in bias (underfitted) or high in variance (overfitted).
 
@@
 \centerline{\includegraphics[width=6.85in]{figure/frankenonoise.png}}
-\caption{A comparison the our prediction and the Franke function with no noise}
+\caption{A comparison of our best OLS prediction and the Franke function with no noise, at polynomial degree 10}
 \label{frankenonoise}
@@
 \centerline{\includegraphics[width=3.25in]{figure/frankenonoisemse.png}}
-\caption{Test and train MSE for the Franke function with no noise}
+\caption{Test and train MSE for the Franke function with no noise, using OLS}
 \label{frankenonoisemse}
@@
 
-\begin{figure} 
-\centerline{\includegraphics[width=3.25in]{figure/frankenonoiser2.png}}
-\caption{Test and train R2-score for the Franke function with no noise}
-\label{frankenonoiser2}
-\end{figure}
+We find that OLS creates a good fit for the Franke Function.
+We can increase our polynomial degree to around 20 before we see signs of overfitting.
+\footnote{Here we also see something a bit surprising: We see that our own implementation of OLS follows Scikit's implementation up to around polynomial degree 10, at which point scikit's MSE increases rapidly.
+We chalk this up to numerical errors and differing implementation of the normal equations, as we simply use the normal equation exactly, while scikit uses an implementation which is probably computationally more efficient but gives slight differences when comparing to the normal equation.
+These differences seem to create a large error after a certain size of the design matrix is reached, which we can not adequately explain.} This is most likely because the Franke function is a sum of exponents, which can be expanded into polynomials.
+We would expect to see higher polynomial degrees give us better and better prediction, up to a certain point.
+When we reach degrees higher than around 20, we see that our test error increases slightly, a sign of overfitting.
+This is also to be expected, as even if the continual Franke function could be expanded into a series of polynomials, we are working with a limited dataset of only 400 samples (and only 320 of those in our training set).
+As we reach higher and higher degrees, we are more and more likely to see unlucky train-test-splits where the best fit for the training data leads to large errors for the test data.
+See Fig.~\ref{frankenonoise} for a comparison between the Franke function and our best prediction, which shows that they are essentially identical.
+Fig.~\ref{frankenonoisemse} shows the train and test MSE by polynomial degree.
+The MSE for test in these plots is the average MSE for 100 bootstraps.
+
+However, when we introduce a stochastic noise, we see that overfitting happens rapidly.
+Without a regularization hyperparameter, parameters fluctuate wildly and our prediction suffers.
+Fig.~\ref{frankenoisemse} shows our MSE up to polynomial degree 25 with an added stochastic noise $\epsilon \sim N(0,0.05)$, where we see that we are overfitting severely.
+See Fig.~\ref{frankienoisecomparison} in the appendix for test MSE for more and more noisy target values.
 
@@
 \centerline{\includegraphics[width=3.25in]{figure/frankenoisemse.png}}
-\caption{Test and train MSE for the Franke function with noise}
+\caption{Test and train MSE for the Franke function with noise, using OLS}
 \label{frankenoisemse}
@@
 
+By using bootstrap resampling, we can study the bias-variance-tradeoff, which shows how the variance of our model between different bootstraps iterations increases as our model overfits our noisy data.
+See Fig.~\ref{frankiebiasvariance}.
+Clearly, we need to introduce regularization if we want a model that does not overfit.
+
 \begin{figure} 
-\centerline{\includegraphics[width=3.25in]{figure/frankenoiser2.png}}
-\caption{Test and train R2-score for the Franke function no noise}
-\label{frankenoiser2}
+\centerline{\includegraphics[width=3.25in]{figure/frankiebiasvariance.png}}
+\caption{Bias, variance and test MSE plotted over polynomial degree for our OLS prediction on the Franke function with an added stochastic noise $\epsilon \sim N(0,0.05)$}
+\label{frankiebiasvariance}
 \end{figure}
 
-We find that OLS creates a good fit for the Franke Function.
-We see few signs of overfitting, even for large polynomial degrees.
-This is most likely because the Franke function is a sum of exponents, which can be expanded into polynomials.
-We would expect to see higher polynomial degrees give us better and better prediction, when there is no noise that could give rise to overfitting.
-See Fig.~\ref{frankenonoise} for a comparison between the Franke function and our prediction, which shows that they are essentially identical.
-Fig.~\ref{frankenonoisemse} and Fig.~\ref{frankenonoiser2} shows the train and test MSE and R2 by polynomial degree, and shows that we are not overfitting even for polynomial degree 25.
-The MSE for test in these plots is the average MSE for 100 bootstraps.
+Table.
+~\ref{ols_mse_table} shows the best polynomial degree and MSE for regular and noisy data.
+Here we see that the best degree without noise is 10, which shows that even without noise, the low amount of data leads to a slight overfit even at very small polynomial degrees.
+This is not surprising, as at $N = 10$ we have $66$ features, which is a substantial amount compared to our measly 320 training datapoints.
+We could of course arbitrarily increase our dataset to ludicrous levels, but doing linear regression with this amount of data gives valuable insight into how our performance is dependent on our availiable data.
 
-However, when we introduce a stochastic noise, we see that overfitting happens rapidly.
-Without a regularization hyperparameter, we see that the parameters fluctuate wildly and our prediction suffers.
-Fig.~\ref{frankenoisemse} and Fig.~\ref{frankenoiser2} shows our MSE and R2 up to polynomial degree 25, where we see that we are overfitting severely.
-See Fig.~\ref{frankienoisecomparison} in the appendix for test MSE for more and more noisy target values.
-Here we also see something a bit surprising: We see that our own implementation of OLS follows Scikit's implementation up to around polynomial degree DEGREE, at which point scikit's MSE increases rapidly.
-We chalk this up to numerical errors and differing implementation of the normal equations, as we simply use the normal equation exactly, while scikit uses an implementation which is probably computationally more efficient but gives slight differences when comparing to the normal equation.
-These differences seem to create a large error when the model is overfitted, which we can not adequately explain.
-
-By using bootstrap resampling, we can study the bias-variance-tradeoff, which shows how the variance between different bootstraps iterations increases as our model overfits our noisy data.
-See Fig.
-3.
-Clearly, we need to introduce regularization if we want a model that does not overfit.
+\begin{table}[t]
+\caption{Best MSE and polynomial degree for OLS}
+\begin{center}
+\label{ols_mse_table}
+\begin{tabular}{c l l}
+& & \\ % put some space after the caption
+\hline
+Std of noise & Best MSE & Best degree \\
+\hline
+0 & 0.0004335 & 10\\
+0.05 & 0.0038302 & 7 \\
+0.10 & 0.0139833 & 7 \\
+\hline
+\end{tabular}
+\end{center}
+\end{table}
 
@@
 
-Using Ridge regression, we see from Fig.
-NUMBER that larger and larger values of $\lambda$ efficiently prevents overfitting and reduces the variance between predictions.
+Using Ridge regression, we see from Fig.~\ref{frankiedifflambdas} that larger and larger values of $\lambda$ efficiently prevents overfitting and reduces the variance between predictions.
 Using Scikit's gridsearch, we find the optimal $\lambda$ to be NUMBER and the optimal polynomial degree to be DEGREE, with an average MSE of NUMBER, which is lower than OLS for the Franke function with an added stochastic noise $\epsilon \sim N(0,0.05)$.
 Ridge regression successfully reduces the model complexity and prevents overfitting for higher polynomial degrees.
+
+\begin{figure} 
+\centerline{\includegraphics[width=3.25in]{figure/frankiedifflambdas.png}}
+\caption{Bias, variance and test MSE plotted over polynomial degree for our Ridge prediction on the Franke function with an added stochastic noise $\epsilon \sim N(0,0.05)$, using lambdas of increasing size.
+Here we have reduced the number of datapoints to only 100 to force overfitting earlier.}
+\label{frankiedifflambdas}
+\end{figure}
 
@@
 This is likely because the elevation data has a jagged and complex shape, which cannot be accurately recreated, even with a polynomial of degree $50$.
-We should remember that the Franke function can be expanded into a series of polynomials, which is why we can decrease our MSE as much as we like by increasing our model complexity (as long as our data is not noisy).
+We should remember that the Franke function can be expanded into a series of polynomials, which is why we can expect to get a lower and lower MSE as the polynomial degree increases (so long as we have enough data and it is not noisy).
 The same is not true for our elevation data, and as such we have little to gain from increasing our model complexity after a certain point.
@@
 
-\newpage
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Figures}
-\label{sect_figure}
-
-All figures should be positioned at the top of the page where possible.
+% All figures should be positioned at the top of the page where possible.
 All figures should be numbered consecutively and centered under the figure as shown in Fig.~\ref{figure_ASME}.
@@
 %%%%%%%%%%%%%%%% begin figure %%%%%%%%%%%%%%%%%%%
-\begin{figure}[t]
-\begin{center}
-\setlength{\unitlength}{0.012500in}%
-\begin{picture}(115,35)(255,545)
-\thicklines
-\put(255,545){\framebox(115,35){}}
-\put(275,560){Beautiful Figure}
-\end{picture}
-\end{center}
-\caption{The caption of a single sentence does not have period at the end}
-\label{figure_ASME} 
-\end{figure}
+% \begin{figure}[t]
+% \begin{center}
+% \setlength{\unitlength}{0.012500in}%
+% \begin{picture}(115,35)(255,545)
+% \thicklines
+% \put(255,545){\framebox(115,35){}}
+% \put(275,560){Beautiful Figure}
+% \end{picture}
+% \end{center}
+% \caption{The caption of a single sentence does not have period at the end}
+% \label{figure_ASME} 
+% \end{figure}
 %%%%%%%%%%%%%%%% end figure %%%%%%%%%%%%%%%%%%% 
@@
 
-In the following subsections, I have inserted figures that have been provided by authors in order to demonstrate what to avoid.
-In each case the authors provided figures that are 3.25in wide and 600dpi in the .tif graphics format.
-The papers containing these figures have been held from production due to their poor quality.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{The 1st Example of Bad Figure}
-
 %%%%%%%%%%%%%%%% begin figure %%%%%%%%%%%%%%%%%%%
-%%% 3.34in is the maximum width you can have for a figure
-\begin{figure} 
-\centerline{\includegraphics[width=3.34in]{figure/FMANU_MD_05_1107_11.jpg}}
-\caption{Example taken from a paper that was held from production because the image quality is poor.  ASME sets figures captions in 8pt, Helvetica Bold.}
-\label{fig_example1.jpg}
-\end{figure}
-%%%%%%%%%%%%%%%% end figure %%%%%%%%%%%%%%%%%%%
-
-In order to place the figure in this template using MSWord, select Insert Picture from File, and use wrapping that is top and bottom. Make sure the figure is 3.25in wide.
-
-Figure~`\ref{fig_example1.jpg}
-was taken from a recent paper that was held from publication, because the text is fuzzy and unreadable.
-It was probably obtained by taking a screen shot of the computer output of the authors software.
-This means the original figure was 72dpi (dots per inch) on a computer screen.
-There is no way to improve the quality such a low resolution figure.
-
-In order to understand how poor the quality of this figure is, please zoom in slightly, say to 200\%.
-Notice that while the font of the paper is clear at this size, the font in the figures is fuzzy and blurred.
-It is impossible to make out the small symbol beside the numbers along the abscissa of the graph.
-Now consider the labels Time and Cost.
-They are clearly in fonts larger that the text of the article, yet the pixilation or rasterization, associated with low resolution is obvious.
-This figure must be regenerated at higher resolution to ensure quality presentation.
-
-The poor quality of this figure is immediately obvious on the printed page, and reduces the impact of the research contribution of the paper, and in fact detracts from the perceived quality of the journal itself.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{The 2nd Example of Bad Figure}
-
-%%%%%%%%%%%%%%%% begin figure %%%%%%%%%%%%%%%%%%%
-\begin{figure} 
-\centerline{\includegraphics[width=3.34in]{figure/FMANU_MD_05_1272_5.jpg}}
-\caption{While this figures is easily readable at a double column width of 6.5in, when it is shrunk to 3.25in column width the text is unreadable. This paper was held from production.}
-\label{fig_example2.jpg}
-\end{figure}
-%%%%%%%%%%%%%%%% end figure %%%%%%%%%%%%%%%%%%%
-
-Figure~\ref{fig_example2.jpg}
-demonstrates a common problem that arises when a figure is scaled down fit a single column width of 3.25in.
-The original figure had labels that were readable at full size, but become unreadable when scaled to half size.
-This figure also suffers from poor resolution as is seen in the jagged lines the ovals that form the chain.
-
-This problem can be addressed by increasing the size of the figure to a double column width of 6.5in, so the text is readable.
-But this will not improve the line pixilation, and a large low resolution figure is less desirable than a small one.
-This also significantly expands the length of the paper, and may cause it to exceed the JMD nine page limit.
-Additional pages require page charges of \$200 per page.
-It is best to regenerate the figure at the resolution that ensures a quality presentation.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{The 3rd Example of Bad Figure}
-%%%%%%%%%%%%%%%% begin figure %%%%%%%%%%%%%%%%%%%
-\begin{figure} 
-\centerline{\includegraphics[width=3.25in]{figure/FMANU_MD_04_1274_13.jpg}}
-\caption{Another example of a figure with unreadable text.
-Even when the paper was expanded to double column width the text as shown in Fig.~\ref{fig_example4.jpg} was of such low quality that the paper was held from production.}
-\label{fig_example3.jpg}
-\end{figure}
+% \begin{figure} 
+% \centerline{\includegraphics[width=3.34in]{figure/FMANU_MD_05_1272_5.jpg}}
+% \caption{While this figures is easily readable at a double column width of 6.5in, when it is shrunk to 3.25in column width the text is unreadable. This paper was held from production.}
+% \label{fig_example2.jpg}
+% \end{figure}
 %%%%%%%%%%%%%%%% end figure %%%%%%%%%%%%%%%%%%%
@@
 %%% the maximum width in double column is 6.85in
-\begin{figure*} 
-\centerline{\includegraphics[width=6.85in]{figure/FMANU_MD_04_1274_13.jpg}}
-\caption{A figure expanded to double column width the text from Figure~\ref{fig_example3.jpg}}
-\label{fig_example4.jpg}
-\end{figure*}
+% \begin{figure*} 
+% \centerline{\includegraphics[width=6.85in]{figure/FMANU_MD_04_1274_13.jpg}}
+% \caption{A figure expanded to double column width the text from Figure~\ref{fig_example3.jpg}}
+% \label{fig_example4.jpg}
+% \end{figure*}
 %%%%%%%%%%%%%%%% end figure %%%%%%%%%%%%%%%%%%%
-An author provided the high resolution image 
-in Fig.~\ref{fig_example3.jpg}
-that was sized to a single column width of 3.25in.  Upon seeing the poor quality of the text, the publisher scaled the image to double column width as shown in Fig.~\ref{fig_example4.jpg} 
-at which point it took half of a page.
-The publisher went on to do this for all eight figures generating four pages of figures that the author did not expect.
-ASME stopped production of the paper even with the larger figures due to the pixilation of the font.
-
-Clearly the text in this figure is unreadable, and it is doubtful that the author can print the output in a way that it is readable.  This is a problem that the author must solve, not the publisher.
-
-As you might expect, I have many more examples, but in the end the author is the best judge of what is needed in each figure.
-ASME simply requires that the image meet a minimum standard for font and line quality, specifically the font should be the appropriate size and not be blurred or pixilated, and that lines should be the appropriate weight and have minimal, preferably no, pixilation or rasterization.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Tables}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%% begin table   %%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{table}[t]
-\caption{Figure and table captions do not end with a period}
-\begin{center}
-\label{table_ASME}
-\begin{tabular}{c l l}
-& & \\ % put some space after the caption
-\hline
-Example & Time & Cost \\
-\hline
-1 & 12.5 & \$1,000 \\
-2 & 24 & \$2,000 \\
-\hline
-\end{tabular}
-\end{center}
-\end{table}
-%%%%%%%%%%%%%%%% end table %%%%%%%%%%%%%%%%%%% 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-All tables should be numbered consecutively  and centered above the table as shown in Table~\ref{table_ASME}.
-The body of the table should be no smaller than 7 pt.
-There should be a minimum two line spaces between tables and text.
 
@@
 \centerline{\includegraphics[width=3.25in]{figure/frankienoisecomparison.png}}
-\caption{Test and train MSE for the Franke function with no noise}
+\caption{Test and train MSE for the Franke function with an increasing amount of noise}
 \label{frankienoisecomparison}

### 2022-10-06 doc/paper/paper.tex
Finally done with these equations

@@
 Models with high variance are usually overfitted to the training data, whilst low variance models are the opposite.
+
 A good rule of thumb for finding a good fit is looking at the intercept between bias and variance.
-This is because the cost function <PROOF task c>.
+This is because the cost function can be decomposed into these two elements.
+The proof for this is as follows:
+
+Given data $\bm{y} = f(\bm{x}) + \bm{\epsilon}$ and a OLS model $\bm{\tilde{y}} = X\bm{\beta}$, we define
+the Mean Squared Error as follows
+
+\begin{gather}
+C(X,\beta) = \frac{1}{n}\sum_{i=0}^{n-1}(y_i - \tilde{y_i})^2
+\end{gather}
+
+This is simply the expectation value of $(\bm{y} - \bm{\tilde{y}})^2$
+
+\begin{gather}
+C(X,\beta) = E\left[(\bm{y} - \bm{\tilde{y}})^2\right]
+\end{gather}
+
+To derive the bias variance decomposition, we now add $E[\bm{\tilde{y}}] - E[\bm{\tilde{y}}]$ 
+as a sneaky trick. We will also replace $\bm{y}$ with our definition $\bm{y} = \bm{f} + \bm{\epsilon}$
+
+\begin{gather}
+C(X,\beta) = E\left[(\bm{f} + \bm{\epsilon} - E[\bm{\tilde{y}}] + E[\bm{\tilde{y}}] - \bm{\tilde{y}})^2\right]
+\end{gather}
+%%%%%%%%%%%%%%%%%5
+\begin{multline}
+= E[(\bm{f} + \bm{\epsilon} - E[\bm{\tilde{y}}])^2 + 2[(\bm{f} + \bm{\epsilon} - E[\bm{\tilde{y}}])(E[\bm{\tilde{y}}] \\
+	- \bm{\tilde{y}})] +(E[\bm{\tilde{y}}] - \bm{\tilde{y}})^2]
+\end{multline}
+
+\begin{multline}
+= E[(\bm{f} + \bm{\epsilon} - E[\bm{\tilde{y}}])^2] + \\
+	2E[(\bm{f} + \bm{\epsilon} - E[\bm{\tilde{y}}])(E[\bm{\tilde{y}}] - \bm{\tilde{y}})] + E[(E[\bm{\tilde{y}}] - \bm{\tilde{y}})^2]
+\end{multline}
+
+Let us focus our attention on the middle term first.
+
+\begin{gather}
+2E\left[(\bm{f} + \bm{\epsilon} + E[\bm{\tilde{y}}])(E[\bm{\tilde{y}}] - \bm{\tilde{y}})\right] 
+\end{gather}
+
+Expanding the multiplication:
+
+\begin{gather}
+2E\left[\bm{f}E[\bm{\tilde{y}}] + \bm{\epsilon}E[\bm{\tilde{y}}] - E[\bm{\tilde{y}}]E[\bm{\tilde{y}}] - \bm{f}\bm{\tilde{y}} - \bm{\epsilon}\bm{\tilde{y}} + E[\bm{\tilde{y}}]\bm{\tilde{y}}\right] 
+\end{gather}
+
+Distributing the expectation value:
+
+\begin{gather}
+2\left[E[\bm{f}E[\bm{\tilde{y}}]] + E[\bm{\epsilon}E[\bm{\tilde{y}}]] - E[E[\bm{\tilde{y}}]E[\bm{\tilde{y}}]] - E[\bm{f}\bm{\tilde{y}}] - E[\bm{\epsilon}\bm{\tilde{y}}] + E[E[\bm{\tilde{y}}]\bm{\tilde{y}}]\right]
+\end{gather}
+
+We know that $\bm{\tilde{y}}$ and $\bm{f}$ are non-stochastic, and we know that the expectation value of $\bm{\epsilon}$ is $0$. We also know that 
+$\bm{\epsilon}$ and $\bm{\tilde{y}}$ are independent, which means we can write $E[\bm{\epsilon}\bm{\tilde{y}}]$ as $E[\bm{\epsilon}]E[\bm{\tilde{y}}]$ We therefore have
+
+\begin{gather}
+2\left[\bm{f}\bm{\tilde{y}} + 0 - \bm{\tilde{y}}\bm{\tilde{y}} - \bm{f}\bm{\tilde{y}} - 0 + \bm{\tilde{y}}\bm{\tilde{y}}\right] = 0 
+\end{gather}
+
+We then have
+
+\begin{gather}
+C(X,\beta) = E\left[(\bm{f} + \bm{\epsilon} - E[\bm{\tilde{y}}])^2\right] + E\left[(E[\bm{\tilde{y}}] - \bm{\tilde{y}})^2\right] 
+\end{gather}
+
+The second term is by definition the variance of model.
+
+\begin{gather}
+C(X,\beta) = E\left[(\bm{f} + \bm{\epsilon} - E[\bm{\tilde{y}}])^2\right] + var[\bm{\tilde{f}}] 
+\end{gather}
+
+And now the final term
+
+\begin{gather}
+E\left[(\bm{f} + \bm{\epsilon} - E[\bm{\tilde{y}}])^2\right]
+\end{gather}
+
+\begin{multline}
+= E[\bm{f}^2 + \bm{f}\bm{\epsilon} - \bm{f}E[\bm{\tilde{y}}] + \bm{f\epsilon} + \bm{\epsilon}^2 - \\ 
+	\bm{\epsilon}E[\bm{\tilde{y}}] - \bm{f}E[\bm{\tilde{y}}] - \bm{\epsilon}E[\bm{\tilde{y}}] + E[\bm{\tilde{y}}]^2]
+\end{multline}
+
+After distributing the expectation value we get the following, keeping in mind that $E[\bm{f}] = E[\bm{y} - \bm{\epsilon}] = E[\bm{y}] - E[\bm{\epsilon}] = \bm{y}$
+
+\begin{gather}
+\bm{y}^2 - 2\bm{y}\bm{\tilde{y}} + E[\bm{\tilde{y}}]^2 + \sigma^2 \\
+(\bm{y} + E[\bm{\tilde{y}}]^2 + \sigma^2 
+\end{gather}
+
+We see that the quadratic form is equal to $bias[\bm{\tilde{y}}]^2$. We can then conclude that the cost function can be rewritten as follows:
+
+\begin{gather}
+C(X,\beta) = bias[\bm{\tilde{y}}]^2 + var[\bm{\tilde{f}}] + \sigma^2 
+\end{gather}
+
 Because the cost function is a sum of the bias and the variance, the lowest error will generally occur when both bias and variance are low.

### 2022-10-06 src/g_task_b.py
large commit, refactor and code cleanup, added docs

@@ -1,83 +1,38 @@
 """
-task b (and task g): plot terrain, approximate terrain with OLS and calculate MSE, R2 beta over model complexity for
-                     real data. Performs task_b, so no resampling.
+task b (and task g): plot terrain, approximate terrain with OLS (own implementation and scikit) and calculate MSE, R2 &
+                     beta over model complexity for real data. Performs task_b, so no resampling.
 """
-from imageio import imread
-import sys
-import argparse
-
 # Our own library of functions
 from utils import *
 
 np.random.seed(42069)
 
-argv = sys.argv[1:]
-
-parser = argparse.ArgumentParser(
-    description="Compute task g.b"
-)
-
-# filename is optional
-parser.add_argument("-f", "--file", help="The filename to apply filter to")
-
-# parse arguments and call run_filter
-args = parser.parse_args()
-
 # parameters
 N = 20
 betas_to_plot = 9
+# parameters for synthetic data
 noise = 0.05
 scaling = False
+
+# get data
+X, X_train, X_test, z, z_train, z_test, scaling, x, y, z = read_in_dataset(
+    N, scaling, noise
+)
 
 # implemented model under testing
 OLS_model = OLS
 # scikit model under testing
 OLS_scikit = LinearRegression(fit_intercept=False)
 
-if args.file:
-    # Load the terrain
-    z = np.asarray(imread(args.file), dtype="float64")
-    x = np.arange(z.shape[0])
-    y = np.arange(z.shape[1])
-    x, y = np.meshgrid(x, y, indexing="ij")
+# perform linear regression
+betas, z_preds_train, z_preds_test, z_preds = linreg_to_N(
+    X, X_train, X_test, z_train, z_test, N, scaling=scaling, model=OLS_model
+)
 
-    # split data into test and train
-    X, X_train, X_test, z_train, z_test = preprocess(x, y, z, N, 0.2)
-
-    # normalize data
-    X, X_train, X_test, z, z_train, z_test = minmax_dataset(
-        X, X_train, X_test, z, z_train, z_test
-    )
-
-    # perform linear regression
-    betas, z_preds_train, z_preds_test, z_preds = linreg_to_N(
-        X, X_train, X_test, z_train, z_test, N, model=OLS_model
-    )
-
[…]

### 2022-10-06 src/g_task_c.py

@@ -1,84 +1,57 @@
-from mpl_toolkits.mplot3d import Axes3D
-import matplotlib.pyplot as plt
-from matplotlib import cm
-from matplotlib.ticker import LinearLocator, FormatStrFormatter
-import numpy as np
-from random import random, seed
-from sklearn.model_selection import train_test_split
-from sklearn.linear_model import LinearRegression
-from imageio import imread
-from sklearn.preprocessing import StandardScaler
-
+"""
+task c (and task g): plots bias-variance tradeoff with bootstrap for OLS
+"""
 # Our own library of functions
 from utils import *
 
-# Load the terrain
-z_terrain1 = np.asarray(imread("../data/tiny_SRTM_data_Norway_1.tif"), dtype=float)
-x_terrain1 = np.arange(z_terrain1.shape[0])
-y_terrain1 = np.arange(z_terrain1.shape[1])
-x1, y1 = np.meshgrid(x_terrain1, y_terrain1, indexing="ij")
-
-z_terrain2 = imread("../data/SRTM_data_Norway_2.tif")
-x_terrain2 = np.arange(z_terrain2.shape[0])
-y_terrain2 = np.arange(z_terrain2.shape[1])
-x2, y2 = np.meshgrid(x_terrain2, y_terrain2, indexing="ij")
-
-# Show the terrain
-plt.plot()
-plt.subplot(121)
-plt.title("Terrain over Norway 1")
-plt.imshow(z_terrain1)
-plt.xlabel("Y")
-plt.ylabel("X")
-
-plt.subplot(122)
-plt.title("Terrain over Norway 2")
-plt.imshow(z_terrain2)
-plt.xlabel("Y")
-plt.ylabel("X")
-plt.show()
-
-N = 15
-bootstraps = 100
-scaling = False
-
 np.random.seed(42069)
 
-def task_c(x, y, z, N, scaling, bootstraps):
-    X, X_train, X_test, z_train, z_test = preprocess(x, y, z, N, 0.2)
+# parameters
+N = 20
+bootstraps = 100
+# parameters for synthetic data
+noise = 0.05
+scaling = False
 
-    X, X_train, X_test, z, z_train, z_test = normalize_task_g(
-        X, X_train, X_test, z, z_train, z_test
+# get data
+X, X_train, X_test, z, z_train, z_test, scaling, x, y, z = read_in_dataset(N, scaling, noise)
+
+# result arrays
+errors = np.zeros(N)
[…]

### 2022-10-06 src/g_task_d.py

@@ -47,7 +47,7 @@ np.random.seed(42069)
 def task_d(x, y, z, N, K, bootstraps, scaling):
     X, X_train, X_test, z_train, z_test = preprocess(x, y, z, N, 0.2)
 
-    X, X_train, X_test, z, z_train, z_test = normalize_task_g(
+    X, X_train, X_test, z, z_train, z_test = minmax_dataset(
         X, X_train, X_test, z, z_train, z_test
     )
 

### 2022-10-06 src/g_task_e_betas.py

@@ -1,61 +1,47 @@
 """
 task g: calculate and plot ridge beta value over model complexity
 """
-from imageio import imread
-
 # Our own library of functions
 from utils import *
 
 np.random.seed(42069)
 
-# read in data
-z_terrain1 = imread("../data/tiny_SRTM_data_Norway_1.tif")
-x_terrain1 = np.arange(z_terrain1.shape[0])
-y_terrain1 = np.arange(z_terrain1.shape[1])
-x1, y1 = np.meshgrid(x_terrain1, y_terrain1, indexing="ij")
-
-z_terrain2 = imread("../data/SRTM_data_Norway_2.tif")
-x_terrain2 = np.arange(z_terrain2.shape[0])
-y_terrain2 = np.arange(z_terrain2.shape[1])
-x2, y2 = np.meshgrid(x_terrain2, y_terrain2, indexing="ij")
-
-# Highest order polynomial we fit with
-N = 30
+# Parameters
+N = 20
 lambdas = np.logspace(-15, -9, 6)
 lambdas[-1] = 10000000000
-betas_to_plot = 5
+betas_to_plot = 9
 
-def ridge_betas(x, y, z, N, lambdas, betas_to_plot):
-    # split data into train and test
-    X, X_train, X_test, z_train, z_test = preprocess(x, y, z, N, 0.2)
+# Parameters for synthetic data
+noise = 0.05
+scaling = False
 
-    # normalize data
-    X, X_train, X_test, z, z_train, z_test = normalize_task_g(
-        X, X_train, X_test, z, z_train, z_test
+# get data
+X, X_train, X_test, z, z_train, z_test, scaling, x, y, z = read_in_dataset(
+    N, scaling, noise
+)
+
+# calculate beta values
+for i in range(len(lambdas)):
+    plt.subplot(321 + i)
+    betas, z_preds_train, z_preds_test, _ = linreg_to_N(
+        X,
+        X_train,
+        X_test,
+        z_train,
+        z_test,
+        N,
+        scaling=scaling,
+        model=ridge,
+        lam=lambdas[i],
     )
 
-    # calculate beta values
-    for i in range(len(lambdas)):
-        plt.subplot(321 + i)
-        betas, z_preds_train, z_preds_test, _ = linreg_to_N(
-            X,
-            X_train,
-            X_test,
-            z_train,
-            z_test,
-            N,
-            model=ridge,
-            lam=lambdas[i],
-        )
-
     # plot beta values
[…]

### 2022-10-06 src/g_task_e_msecomparison.py

@@ -1,107 +1,104 @@
 """
-task g: compares resampling methods (bootstrap, cross validation and scikit cross validation for ridge model
+task g: compares resampling methods (bootstrap, cross validation and scikit cross validation) for ridge model
 """
 from sklearn.linear_model import Ridge
-from sklearn.model_selection import train_test_split, cross_validate, KFold
-from imageio import imread
+from sklearn.model_selection import cross_validate, KFold
 
 # Our own library of functions
 from utils import *
 
 np.random.seed(42069)
 
-# read in data
-z_terrain1 = imread("../data/tiny_SRTM_data_Norway_1.tif")
-x_terrain1 = np.arange(z_terrain1.shape[0])
-y_terrain1 = np.arange(z_terrain1.shape[1])
-x1, y1 = np.meshgrid(x_terrain1, y_terrain1, indexing="ij")
-
-z_terrain2 = imread("../data/SRTM_data_Norway_2.tif")
-x_terrain2 = np.arange(z_terrain2.shape[0])
-y_terrain2 = np.arange(z_terrain2.shape[1])
-x2, y2 = np.meshgrid(x_terrain2, y_terrain2, indexing="ij")
-
-# highest order polynomial we fit with
-N = 10
-# amount of folds
-K = 10
+# Parameters
+N = 20
+K = 20
 bootstraps = 100
+noise = 0.05
+scaling = False
+kfolds = KFold(n_splits=K)
 
-# todo change plot to only gridsearched lambda
-def ridge_mse_compare_resampling(x, y, z, N, K, bootstraps):
-    kfolds = KFold(n_splits=K)
+# if true, plot only lambda that gives lowest MSE
+# if false, plot range of different lambdas
+plot_only_best_lambda = True
+lambdas = np.logspace(-12, -4, 6)
 
-    # split into training and test
-    X, X_train, X_test, z_train, z_test = preprocess(x, y, z, N, 0.2)
+# read in data
+X, X_train, X_test, z, z_train, z_test, scaling, x, y, z = read_in_dataset(
+    N, scaling, noise
+)
+z = z.ravel()
 
-    # normalize data based on train
-    X, X_train, X_test, z, z_train, z_test = minmax_dataset(
-        X, X_train, X_test, z, z_train, z_test
-    )
+# plot only the gridsearched lambda
+if plot_only_best_lambda:
+    ridge = Ridge(fit_intercept=scaling)
[…]

### 2022-10-06 src/g_task_f_biasvariance.py

@@ -3,84 +3,79 @@ task g: performs bias-variance tradeoff for lasso regression using the scikit le
         lasso model
 """
 from sklearn.linear_model import Lasso
-from imageio import imread
-
 
 # Our own library of functions
 from utils import *
 
 np.random.seed(42069)
 
-# read in data
-z_terrain1 = imread("../data/tiny_SRTM_data_Norway_1.tif")
-x_terrain1 = np.arange(z_terrain1.shape[0])
-y_terrain1 = np.arange(z_terrain1.shape[1])
-x1, y1 = np.meshgrid(x_terrain1, y_terrain1, indexing="ij")
+# parameters
+N = 20
+K = 20
+bootstraps = 100
+plot_only_best_lambda = True
+lambdas = np.logspace(-12, -4, 6)
+# synthetic parameters
+noise = 0.05
+scaling = False
 
-z_terrain2 = imread("../data/SRTM_data_Norway_2.tif")
-x_terrain2 = np.arange(z_terrain2.shape[0])
-y_terrain2 = np.arange(z_terrain2.shape[1])
-x2, y2 = np.meshgrid(x_terrain2, y_terrain2, indexing="ij")
+# read in and get data
+X, X_train, X_test, z, z_train, z_test, scaling, x, y, z = read_in_dataset(N, scaling, noise)
+z = z.ravel()
 
-# Highest order polynomial we fit with
-N = 8
-bootstraps = 10
-lambdas = np.logspace(-6, -1, 6)
+# plot only the gridsearched lambda
+if plot_only_best_lambda:
+    lasso = Lasso(fit_intercept=scaling)
+    lam, _, _ = find_best_lambda(X, z, lasso, lambdas, N, K)
+    lambdas = [lam]
 
-# todo use only gridsearched lambda
-def lasso_bias_variance(x, y, z, N, bootstraps, lambdas):
-    X, X_train, X_test, z_train, z_test = preprocess(x, y, z, N, 0.2)
+# loop through different lambda values
+for i in range(len(lambdas)):
+    if not plot_only_best_lambda:
+           plt.subplot(321 + i)
+           plt.suptitle(f"Bias variance tradeoff for lasso regression")
 
-    X, X_train, X_test, z, z_train, z_test = minmax_dataset(
-         X, X_train, X_test, z, z_train, z_test
-     )
+    # model under testing
+    model_Lasso = Lasso(lambdas[i], max_iter=200, fit_intercept=False)
 
[…]

### 2022-10-06 src/task_b.py

@@ -21,7 +21,7 @@ zs = SkrankeFunction(x, y)
 
 # parameters
 betas_to_plot = 9
-noise = 0.1
+noise = 0.05
 N = 20
 scaling = False
 

### 2022-10-06 src/task_f_gridsearch.py

@@ -1,85 +1,34 @@
-from mpl_toolkits.mplot3d import Axes3D
-import matplotlib.pyplot as plt
-from matplotlib import cm
-from matplotlib.ticker import LinearLocator, FormatStrFormatter
-import numpy as np
-from random import random, seed
-from sklearn.model_selection import train_test_split
-from sklearn.linear_model import Lasso, Ridge
-
-from sklearn.model_selection import GridSearchCV
-from sklearn.preprocessing import PolynomialFeatures
-from sklearn.linear_model import LinearRegression
-from sklearn.metrics import mean_squared_error
-from sklearn.pipeline import Pipeline
-
+"""
+USELESS, PLEASE DELETE
+"""
+"""
+task f: evaluates lasso and ridge to find the lambda value yielding the best MSE score.
+"""
+from sklearn.linear_model import Ridge, Lasso
 
 # Our own library of functions
 from utils import *
 
 np.random.seed(42069)
 # Make data.
-x = np.arange(0, 1, 0.15)
-y = np.arange(0, 1, 0.15)
-# x = np.arange(0, 1, 0.05)
-# y = np.arange(0, 1, 0.05)
+x = np.arange(0, 1, 0.05)
+y = np.arange(0, 1, 0.05)
 x, y = np.meshgrid(x, y)
 z = FrankeFunction(x, y)
-# z = SkrankeFunction(x, y)
 
 # Highest order polynomial we fit with
-N = 15
-K = 10
+N = 20
+K = 20
+noise = 0.05
+scaling = True
 
-# Do the linear_regression
-z += 0.15 * np.random.standard_normal(z.shape)
-X, X_train, X_test, z_train, z_test = preprocess(x, y, z, N, 0.2)
+X, X_train, X_test, z, z_train, z_test, scaling, x, y, z = read_in_dataset(N, scaling, noise)
 z = z.ravel()
 
 lambdas = np.logspace(-12, 0, 25)
-ridge = Ridge(fit_intercept=True)
-# ridge = LinearRegression()
-# ridge = Lasso(fit_intercept=False)
+lambdas[-1] = 0
+ridge = Ridge(fit_intercept=scaling)
+lasso = Lasso(fit_intercept=scaling)
 
-ridge_reg = GridSearchCV(
-    ridge,
-    param_grid={"alpha": list(lambdas)},
-    # param_grid={"fit_intercept": [False, True]},
[…]

### 2022-10-06 src/task_f_msecomparison.py

@@ -1,46 +1,40 @@
-from mpl_toolkits.mplot3d import Axes3D
-import matplotlib.pyplot as plt
-from matplotlib import cm
-from matplotlib.ticker import LinearLocator, FormatStrFormatter
-import numpy as np
-from random import random, seed
-from sklearn.linear_model import Ridge, Lasso, LinearRegression
-from sklearn.model_selection import train_test_split, cross_validate, KFold
+"""
+task f: Plots MSE comparison between OLS, ridge and lasso with cross validation for real and synthetic data.
+"""
+from sklearn.linear_model import Ridge, Lasso
+from sklearn.model_selection import cross_validate, KFold
+from imageio import imread
 
 # Our own library of functions
 from utils import *
 
-np.random.seed(42069)
-# Make data.
-x = np.arange(0, 1, 0.05)
-y = np.arange(0, 1, 0.05)
-x, y = np.meshgrid(x, y)
-z = FrankeFunction(x, y)
-# z = SkrankeFunction(x, y)
-scaling = True
-
-# Highest order polynomial we fit with
-N = 8
+# parameters
+N = 20
 K = 10
+noise = 0.05
 kfolds = KFold(n_splits=K)
-
-# Do the linear_regression
-z += 0.05 * np.random.standard_normal(z.shape)
-X, X_train, X_test, z_train, z_test = preprocess(x, y, z, N, 0.2)
-z = z.ravel()
-
+scaling = False
 lambdas = np.logspace(-12, -4, 6)
+
+errors_OLS = np.zeros(N)
+errors_Ridge = np.zeros(N)
+errors_Lasso = np.zeros(N)
+
+# read in data
+X, X_train, X_test, z, z_train, z_test, scaling, x, y, z = read_in_dataset(N, scaling, noise)
+
+# test different values of lambda
 for i in range(len(lambdas)):
+    z = z.ravel()
     plt.subplot(321 + i)
     plt.suptitle(f"MSE by polynomial degree for OLS, Ridge and Lasso regression")
 
-    errors_OLS = np.zeros(N)
-    errors_Ridge = np.zeros(N)
-    errors_Lasso = np.zeros(N)
-
+    # models under testing
     Ridge_model = Ridge(lambdas[i], fit_intercept=scaling)
-    Lasso_model = Lasso(lambdas[i], fit_intercept=scaling, max_iter=2000)
[…]

### 2022-10-06 src/utils.py

@@ -4,11 +4,14 @@ from matplotlib import cm
 from matplotlib.ticker import LinearLocator, FormatStrFormatter
 import numpy as np
 from random import random, seed
-from sklearn.model_selection import train_test_split
+from sklearn.model_selection import train_test_split, GridSearchCV
 from sklearn.linear_model import LinearRegression
 from sklearn.preprocessing import MinMaxScaler
 from sklearn.utils import resample
 from typing import Tuple, Callable
+from imageio import imread
+import sys
+import argparse
 
 
 def FrankeFunction(x, y):
@@ -144,7 +147,9 @@ def crossval(
 def bias_variance(z_test: np.ndarray, z_preds_test: np.ndarray):
     MSEs, _ = scores(z_test, z_preds_test)
     error = np.mean(MSEs)
-    bias = np.mean((z_test - np.mean(z_preds_test, axis=1, keepdims=True).flatten()) ** 2)
+    bias = np.mean(
+        (z_test - np.mean(z_preds_test, axis=1, keepdims=True).flatten()) ** 2
+    )
     variance = np.mean(np.var(z_preds_test, axis=1, keepdims=True))
 
     return error, bias, variance
@@ -311,3 +316,67 @@ def scores(z, z_preds):
         R2s[n] = R2(z, z_preds[:, n])
 
     return MSEs, R2s
+
+
+def find_best_lambda(X, z, model, lambdas, N, K):
+    model = GridSearchCV(
+        estimator=model,
+        param_grid={"alpha": list(lambdas)},
+        scoring="neg_mean_squared_error",
+        cv=K,
+    )
+    best_polynomial = 0
+    best_lambda = 0
+    best_MSE = 10**10
+
+    for n in range(N):
+        print(n)
+        l = int((n + 1) * (n + 2) / 2)  # Number of elements in beta
+        model.fit(X[:, :l], z)
+
+        if -model.best_score_ < best_MSE:
+            best_MSE = -model.best_score_
+            best_lambda = model.best_params_["alpha"]
+            best_polynomial = n
+
+    return best_lambda, best_MSE, best_polynomial
+
+
+def read_in_dataset(N, scaling, noise):
+    argv = sys.argv[1:]
+
+    parser = argparse.ArgumentParser(description="Compute task g.b")
+
+    # filename is optional
+    parser.add_argument("-f", "--file", help="The filename to apply filter to")
+
+    # parse arguments and call run_filter
[…]

### 2022-10-06 doc/paper/paper.tex

@@
 Using Ridge regression, we see from Fig.~\ref{frankiedifflambdas} that larger and larger values of $\lambda$ efficiently prevents overfitting and reduces the variance between predictions.
-Using Scikit's gridsearch, we find the optimal $\lambda$ to be NUMBER and the optimal polynomial degree to be DEGREE, with an average MSE of NUMBER, which is lower than OLS for the Franke function with an added stochastic noise $\epsilon \sim N(0,0.05)$.
-Ridge regression successfully reduces the model complexity and prevents overfitting for higher polynomial degrees.
+Using Scikit's gridsearch, we find the optimal $\lambda$ to be $10^{-8}$ and the optimal polynomial degree to be 7, with an average MSE of $0.0037999$, which is very slightly lower than OLS for the Franke function with an added stochastic noise $\epsilon \sim N(0,0.05)$.
+We see from this that Ridge regression has not really had much of an effect in decreasing the MSE.
+A possible reason for this is again the form of our synthetic data.
 

### 2022-10-07 doc/paper/paper.tex
More paper

@@
 \usepackage{listings}
+\usepackage{float}
 \usepackage{xcolor}
@@
 Thirdly, we repeat for a given large number of iterations.
-Since we use all the test predictions to calculate the MSE, bias and variance, we will then according to the central limit theorem <reference?> obtain a normal distribution of these variables.
+Since we use all the test predictions to calculate the MSE, bias and variance, we will then according to the central limit theorem (REFERENCE!) obtain a normal distribution of these variables.
 The law of large numbers <reference?> then tells us that the mean sample values of the MSE, bias and variance approach the true value of MSE, bias and variance.
@@
 We find that OLS creates a good fit for the Franke Function.
-We can increase our polynomial degree to around 20 before we see signs of overfitting.
-\footnote{Here we also see something a bit surprising: We see that our own implementation of OLS follows Scikit's implementation up to around polynomial degree 10, at which point scikit's MSE increases rapidly.
+We can increase our polynomial degree to around 20 before we see signs of overfitting.\footnote{Here we also see something a bit surprising: We see that our own implementation of OLS follows Scikit's implementation up to around polynomial degree 10, at which point scikit's MSE increases rapidly.
 We chalk this up to numerical errors and differing implementation of the normal equations, as we simply use the normal equation exactly, while scikit uses an implementation which is probably computationally more efficient but gives slight differences when comparing to the normal equation.
 These differences seem to create a large error after a certain size of the design matrix is reached, which we can not adequately explain.} This is most likely because the Franke function is a sum of exponents, which can be expanded into polynomials.
-We would expect to see higher polynomial degrees give us better and better prediction, up to a certain point.
-When we reach degrees higher than around 20, we see that our test error increases slightly, a sign of overfitting.
+We would expect to see higher polynomial degrees give us better and better prediction, up to a certain point.\footnote{And indeed, this seems to be the case, as if we increase the availiable datapoints considerably, as seen in Fig.~\ref{frankiehigherdatapoints} in the appendix, we can reach much higher polynomial degrees without overfitting} When we reach degrees higher than around 20, we see that our test error increases slightly, a sign of overfitting.
 This is also to be expected, as even if the continual Franke function could be expanded into a series of polynomials, we are working with a limited dataset of only 400 samples (and only 320 of those in our training set).
@@
 The MSE for test in these plots is the average MSE for 100 bootstraps.
+For a full list of all parameters used for every plot shown, see Table.~\ref{allparamstable} in the appendix.
 
@@
 This is not surprising, as at $N = 10$ we have $66$ features, which is a substantial amount compared to our measly 320 training datapoints.
-We could of course arbitrarily increase our dataset to ludicrous levels, but doing linear regression with this amount of data gives valuable insight into how our performance is dependent on our availiable data.
+We could of course arbitrarily increase our dataset to ludicrous levels, but doing linear regression with this dataset gives valuable insight into how our performance is dependent on our availiable data.
 
@@
 We see from this that Ridge regression has not really had much of an effect in decreasing the MSE.
-A possible reason for this is again the form of our synthetic data.
+We must now remember what Ridge aims to do.
+With the regularization parameter, we can prevent overfitting if our model is too complex for our data.
+But we should recall that we decide the complexity of our model, by deciding which polynomial degree we use when creating the design matrix.
+In fact, by choosing polynomial degree 7 for our OLS method to get the best MSE for noisy data, we are in fact doing a primitive form of best subset selection, as described in Hastie et Al.
+p 57 \cite{hastie}.
+Almost by definition, the polynomial degree with the best MSE is not overfitted, because if it was, it would not be the model with the lowest MSE among all polynomial degrees.
+Therefore it is not so unexpected that Ridge regression has a relatively low impact on the MSE.
+We could expect that Ridge would allow us to reach higher polynomial degrees without overfitting, by shrinking features of the higher polynomial degrees that are not supported but allowing us to keep other high degree terms (for example, allowing us to use $x^5y^4$, but discard $x^9$, if only $x^5y^4$ was supported by the data).
+But we can see from the Table.
+~\ref{ols_vs_ridge_table} that we are not able to push into higher degrees.
+A possible explanation for this is that every feature is useful, due to the polynomial expansion of the Franke function.
+
+\begin{table}[t]
+\caption{Best MSE for Ridge regression and OLS with 400 datapoints}
+\begin{center}
+\label{ols_vs_ridge_table}
+\begin{tabular}{c | l l l}
+& & \\ % put some space after the caption
+\hline
+Regression & Std of noise & Best MSE & Best degree \\
+\hline
+OLS & 0.05 & 0.0038302 & 7 \\
+Ridge & 0.05 & 0.0037999 & 7 \\
+\hline
+\end{tabular}
+\end{center}
+\end{table}
 
@@
 \appendix       %%% starting appendix
-\section*{Appendix A: Plots}
+\section*{Appendix A: Plots and tables}
 
-\begin{figure} 
+The next pages contain plots and tables that are of interest.
+
+\begin{table*}
+\caption{This table shows every applicable parameter for every figure, so that they may be easily reproduced. Every plot uses the random seed 42069}
+\begin{center}
+\label{allparamstable}
+\begin{tabular}{c | l l l l l l l}
+& & \\ % put some space after the caption
+\hline
+Figure & Highest $N$ & Std of noise & Number of datapoints & Bootstraps & K-folds & $\lambda$ & Scaling \\
+\hline
+\ref{frankenonoise} & 10 & 0 & 400 & - & - & - & None\\
+\ref{frankenonoisemse} & 25 & 0 & 400 & 100 & - & - & None\\
+\ref{frankenoisemse} & 25 & 0.05 & 400 & 100 & - & - & None\\
+\ref{frankiebiasvariance} & 25 & 0.05 & 100 & 100 & - & - & None\\
+\ref{frankiedifflambdas} & 25 & 0.05 & 100 & 100 & - & variable & Centering\\
+\ref{frankienoisecomparison} & 25 & variable & 400 & 100 & - & - & None\\
+\ref{frankiehigherdatapoints} & 30 & 0 & 10000 & 100 & - & - & None\\
+\hline
+\end{tabular}
+\end{center}
+\end{table*}
+
+\begin{figure*}
+% \begin{figure}[t]
 \centerline{\includegraphics[width=3.25in]{figure/frankienoisecomparison.png}}
@@
 \label{frankienoisecomparison}
-\end{figure}
+\end{figure*}
+
+\begin{figure*}
+% \begin{figure}[t]
+\centerline{\includegraphics[width=3.25in]{figure/frankiehigherdatapoints.png}}
+\caption{Franke function fitted up to polynomial degree 30, with 10000 datapoints instead of 400.
+As we can see, we do not overfit even at this polynomial degree when the availiable data is plentiful.
+The highest MSE was reached at $N=29$, and was $0.00003511$, one tenth of our best prediction with only 400 datapoints}
+\label{frankiehigherdatapoints}
+\end{figure*}
 

### 2022-10-07 src/g_task_d.py
large commit, refactor and code cleanup, added docs

@@ -1,113 +1,94 @@
-from mpl_toolkits.mplot3d import Axes3D
-import matplotlib.pyplot as plt
-from matplotlib import cm
-from matplotlib.ticker import LinearLocator, FormatStrFormatter
-import numpy as np
-from random import random, seed
-from sklearn.model_selection import train_test_split, cross_validate, KFold
-from sklearn.linear_model import LinearRegression
-from imageio import imread
+"""
+task d (and task g): compare resampling method MSE for OLS over N polynomial degrees
+"""
+from sklearn.model_selection import cross_validate, KFold
 
 # Our own library of functions
 from utils import *
-# Load the terrain
-z_terrain1 = imread("../data/tiny_SRTM_data_Norway_1.tif")
-x_terrain1 = np.arange(z_terrain1.shape[0])
-y_terrain1 = np.arange(z_terrain1.shape[1])
-x1, y1 = np.meshgrid(x_terrain1, y_terrain1, indexing="ij")
-
-z_terrain2 = imread("../data/SRTM_data_Norway_2.tif")
-x_terrain2 = np.arange(z_terrain2.shape[0])
-y_terrain2 = np.arange(z_terrain2.shape[1])
-x2, y2 = np.meshgrid(x_terrain2, y_terrain2, indexing="ij")
-
-# Show the terrain
-plt.plot()
-plt.subplot(121)
-plt.title("Terrain over Norway 1")
-plt.imshow(z_terrain1)
-plt.xlabel("Y")
-plt.ylabel("X")
-
-plt.subplot(122)
-plt.title("Terrain over Norway 2")
-plt.imshow(z_terrain2)
-plt.xlabel("Y")
-plt.ylabel("X")
-plt.show()
 
 np.random.seed(42069)
-N = 15
+
+# parameters
+N = 10
 K = 10
 bootstraps = 100
-scaling = True
+# parameters for synthetic data
+noise = 0.05
+scaling = False
 
-np.random.seed(42069)
+# read in data
+X, X_train, X_test, z, z_train, z_test, scaling, x, y, z = read_in_dataset(
+    N, scaling, noise
+)
+z = z.ravel()
 
-def task_d(x, y, z, N, K, bootstraps, scaling):
-    X, X_train, X_test, z_train, z_test = preprocess(x, y, z, N, 0.2)
+# model under testing
+OLS_model = OLS
 
-    X, X_train, X_test, z, z_train, z_test = minmax_dataset(
-        X, X_train, X_test, z, z_train, z_test
+# scikit comparison model
[…]

### 2022-10-07 src/g_task_e_msecomparison.py

@@ -14,7 +14,7 @@ N = 20
 K = 20
 bootstraps = 100
 noise = 0.05
-scaling = False
+scaling = True
 kfolds = KFold(n_splits=K)
 
 # if true, plot only lambda that gives lowest MSE
@@ -30,7 +30,7 @@ z = z.ravel()
 
 # plot only the gridsearched lambda
 if plot_only_best_lambda:
-    ridge = Ridge(fit_intercept=scaling)
+    ridge = Ridge(fit_intercept=False)
     lam, _, _ = find_best_lambda(X, z, ridge, lambdas, N, K)
     lambdas = [lam]
 
@@ -46,7 +46,7 @@ for i in range(len(lambdas)):
     errors_boot = np.zeros(N)
 
     # scikit model under testing
-    Ridge_model = Ridge(lambdas[i], fit_intercept=scaling)
+    Ridge_model = Ridge(lambdas[i], fit_intercept=False)
 
     # implemented model under testing
     model = ridge

### 2022-10-07 src/g_task_f_biasvariance.py

@@ -3,7 +3,6 @@ task g: performs bias-variance tradeoff for lasso regression using the scikit le
         lasso model
 """
 from sklearn.linear_model import Lasso
-
 # Our own library of functions
 from utils import *
 
@@ -13,7 +12,7 @@ np.random.seed(42069)
 N = 20
 K = 20
 bootstraps = 100
-plot_only_best_lambda = True
+plot_only_best_lambda = False
 lambdas = np.logspace(-12, -4, 6)
 # synthetic parameters
 noise = 0.05
@@ -25,7 +24,7 @@ z = z.ravel()
 
 # plot only the gridsearched lambda
 if plot_only_best_lambda:
-    lasso = Lasso(fit_intercept=scaling)
+    lasso = Lasso(fit_intercept=False)
     lam, _, _ = find_best_lambda(X, z, lasso, lambdas, N, K)
     lambdas = [lam]
 

### 2022-10-07 src/noise_plot.py

@@ -24,7 +24,6 @@ noises = np.linspace(0, 0.2, 5)
 scaling = False
 bootstraps = 100
 
-
 def noise_plot(x, y, z, N, noise, scaling, bootstraps):
     # add noise
     z += noise * np.random.standard_normal(z.shape)

### 2022-10-07 src/task_e_biasvariance.py

@@ -1,41 +1,46 @@
-from mpl_toolkits.mplot3d import Axes3D
-import matplotlib.pyplot as plt
-from matplotlib import cm
-from matplotlib.ticker import LinearLocator, FormatStrFormatter
-import numpy as np
-from random import random, seed
-from sklearn.model_selection import train_test_split
-from sklearn.linear_model import LinearRegression, Ridge
-
+"""
+task e (and task g): plots bias-variance trade-off for ridge regression
+"""
+from sklearn.linear_model import Ridge
 # Our own library of functions
 from utils import *
 
 np.random.seed(42069)
-# Make data.
-x = np.arange(0, 1, 0.1)
-y = np.arange(0, 1, 0.1)
-x, y = np.meshgrid(x, y)
-z = FrankeFunction(x, y)
-# z = SkrankeFunction(x, y)
 
-# Highest order polynomial we fit with
+# parameters
+K = 20
 N = 10
 bootstraps = 100
-
-# Do the linear_regression
-z += 0.05 * np.random.standard_normal(z.shape)
-X, X_train, X_test, z_train, z_test = preprocess(x, y, z, N, 0.2)
-
+plot_only_best_lambda = True
 lambdas = np.logspace(-10, 0, 6)
-# lambdas[-1] = 1.83 * (10 ** (-5))
-for i in range(len(lambdas)):
-    plt.subplot(321 + i)
-    plt.suptitle("Bias-variance tradeoff for ridge regression")
+# parameter synthetic data
+noise = 0.05
+scaling = False
 
+# read in data
+X, X_train, X_test, z, z_train, z_test, scaling, x, y, z = read_in_dataset(
+    N, scaling, noise
+)
+z = z.ravel()
+
+# plot only the gridsearched lambda
+if plot_only_best_lambda:
+    ridge = Ridge(fit_intercept=False)
+    lam, _, _ = find_best_lambda(X, z, ridge, lambdas, N, K)
+    lambdas = [lam]
+
+# for lambdas
+for i in range(len(lambdas)):
+    if not plot_only_best_lambda:
+       plt.subplot(321 + i)
+       plt.suptitle(f"Bias variance tradeoff for ridge regression")
+
+    # results
     errors = np.zeros(N)
     biases = np.zeros(N)
     variances = np.zeros(N)
 
+    # for polynomial degree
     for n in range(N):
         print(n)
         l = int((n + 1) * (n + 2) / 2)  # Number of elements in beta
[…]

### 2022-10-07 src/task_b.py
remove old versions

@@ -1,132 +1,0 @@
-"""
-task b: plots 3D surface plot of Franke Function next to our approximated polynomial fit of the Franke Function, beta
-        progression over polynomial degree, and MSE and R2 scores of our implemented OLS model compared with the scikit
-        OLS model. The parameter betas_to_plot changes how many betas are plotted, N how many polynomial degrees we test
-        for, and scaling for if we center our data or not. The noise parameter controls how much noise we add to our
-        Franke Function output. Note that these plots are without resampling as specified in task b.
-"""
-import matplotlib.pyplot as plt
-
-# Our own library of functions
-from utils import *
-
-np.random.seed(42069)
-
-# make data
-x = np.arange(0, 1, 0.05)
-y = np.arange(0, 1, 0.05)
-x, y = np.meshgrid(x, y)
-z = FrankeFunction(x, y)
-zs = SkrankeFunction(x, y)
-
-# parameters
-betas_to_plot = 9
-noise = 0.05
-N = 20
-scaling = False
-
-# add noise
-z += noise * np.random.standard_normal(z.shape)
-X, X_train, X_test, z_train, z_test = preprocess(x, y, z, N, 0.2)
-
-# model under testing
-OLS_model = OLS
-
-# perform linear regression
-betas, z_preds_train, z_preds_test, z_preds = linreg_to_N(
-    X, X_train, X_test, z_train, z_test, N, scaling=scaling, model=OLS_model
-)
-
-# calculate OLS scores without resampling
-MSE_train, R2_train = scores(z_train, z_preds_train)
-MSE_test, R2_test = scores(z_test, z_preds_test)
-
-# scikitLearn OLS for comparison
-OLS_scikit = LinearRegression(
-    fit_intercept=False
-)  # false because we use our own scaling
-
-# perform linear regression scikit
-_, z_preds_train_sk, z_preds_test_sk, _ = linreg_to_N(
-    X, X_train, X_test, z_train, z_test, N, scaling=scaling, model=OLS_scikit
-)
-
-# calculate OLS scikit scores without resampling
-MSE_train_sk, R2_train_sk = scores(z_train, z_preds_train_sk)
-MSE_test_sk, R2_test_sk = scores(z_test, z_preds_test_sk)
-
[…]

### 2022-10-07 src/task_c.py

@@ -1,124 +1,0 @@
-"""
-task c: bootstrapped bias-variance trade-off plot for OLS on synthetic data
-"""
-# Our own library of functions
-from utils import *
-import matplotlib
-
-np.random.seed(42069)
-
-# make data
-x = np.arange(0, 1, 0.05)
-y = np.arange(0, 1, 0.05)
-x, y = np.meshgrid(x, y)
-z = FrankeFunction(x, y)
-zs = SkrankeFunction(x, y)
-
-# parameters
-N = 25
-noise = 0.05
-scaling = True
-bootstraps = 100
-
-# model under testing
-model = OLS
-
-# add noise
-# z += noise * np.random.standard_normal(z.shape)
-X, X_train, X_test, z_train, z_test = preprocess(x, y, z, N, 0.2)
-
-# results
-MSEs_test = np.zeros(N)
-MSEs_train = np.zeros(N)
-MSEs_test_sk = np.zeros(N)
-MSEs_train_sk = np.zeros(N)
-
-biases = np.zeros(N)
-variances = np.zeros(N)
-
-ols_scikit = LinearRegression(fit_intercept=False)
-# bootstrap
-for n in range(N):
-    print(n)
-    l = int((n + 1) * (n + 2) / 2)  # Number of elements in beta
-    z_preds_test = bootstrap(
-        X[:, :l],
-        X_train[:, :l],
-        X_test[:, :l],
-        z_train,
-        z_test,
-        bootstraps,
-        scaling=scaling,
-        model=model,
-    )
-
-    z_preds_test_sk = bootstrap(
-        X[:, :l],
-        X_train[:, :l],
-        X_test[:, :l],
-        z_train,
-        z_test,
-        bootstraps,
-        scaling=scaling,
-        model=ols_scikit,
-    )
-
-    _, z_pred_train, _, _ = evaluate_model(
-        X[:, :l], X_train[:, :l], X_test[:, :l], z_train, model
-    )
-
-    _, z_pred_train_sk, _, _ = evaluate_model(
-        X[:, :l], X_train[:, :l], X_test[:, :l], z_train, ols_scikit
-    )
-
-    MSE_test, bias, variance = bias_variance(z_test, z_preds_test)
-    MSE_train = MSE(z_pred_train, z_train)
-
-    MSE_test_sk, _, _ = bias_variance(z_test, z_preds_test_sk)
-    MSE_train_sk = MSE(z_pred_train_sk, z_train)
-
[…]

### 2022-10-07 src/task_d.py

@@ -1,98 +1,0 @@
-import matplotlib.pyplot as plt
-from sklearn.model_selection import cross_validate, KFold
-
-# Our own library of functions
-from utils import *
-
-np.random.seed(42069)
-
-# make data
-x = np.arange(0, 1, 0.05)
-y = np.arange(0, 1, 0.05)
-x, y = np.meshgrid(x, y)
-z = FrankeFunction(x, y)
-zs = SkrankeFunction(x, y)
-
-# parameters
-N = 20
-noise = 0.05
-scaling = False
-K = 10
-bootstraps = 100
-
-# add noise
-z += noise * np.random.standard_normal(z.shape)
-
-# split datasets
-X, X_train, X_test, z_train, z_test = preprocess(x, y, z, N, 0.2)
-
-z = z.ravel()
-
-# model under testing
-OLS_model = OLS
-
-# scikit model under testing
-OLS_model_scikit = LinearRegression(fit_intercept=scaling)
-kfolds = KFold(n_splits=K)
-
-# results cross validation
-errors_cv = np.zeros(N)
-errors_cv_scikit = np.zeros(N)
-
-# results bootstrap
-errors_boot = np.zeros(N)
-biases_boot = np.zeros(N)
-variances_boot = np.zeros(N)
-
-# cross validation
-print("Cross")
-for n in range(N):
-    print(n) # progress counter
-    l = int((n + 1) * (n + 2) / 2)  # Number of elements in beta
-
-    # own implementation
-    errors_cv[n] = crossval(X[:, :l], z, K, scaling=scaling, model=OLS_model)
-
-    # scikit
-    error_scikit = cross_validate(
-        estimator=OLS_model_scikit,
-        X=X[:, :l],
-        y=z,
-        scoring="neg_mean_squared_error",
-        cv=kfolds,
-    )
-    errors_cv_scikit[n] = np.mean(-error_scikit["test_score"])
-
-# bootstrap
-print("Boot")
-for n in range(N):
-    print(n)
-    l = int((n + 1) * (n + 2) / 2)  # Number of elements in beta
-    z_preds_test = bootstrap(
-        X[:, :l],
-        X_train[:, :l],
-        X_test[:, :l],
-        z_train,
-        z_test,
-        bootstraps,
-        scaling=scaling,
-        model=OLS_model,
-    )
-
-    error, bias, variance = bias_variance(z_test, z_preds_test)
[…]

### 2022-10-07 src/task_e_betas.py

@@ -1,53 +1,0 @@
-from mpl_toolkits.mplot3d import Axes3D
-import matplotlib.pyplot as plt
-from matplotlib import cm
-from matplotlib.ticker import LinearLocator, FormatStrFormatter
-import numpy as np
-from random import random, seed
-from sklearn.model_selection import train_test_split
-from sklearn.linear_model import LinearRegression
-
-# Our own library of functions
-from utils import *
-
-np.random.seed(42069)
-# Make data.
-x = np.arange(0, 1, 0.05)
-y = np.arange(0, 1, 0.05)
-x, y = np.meshgrid(x, y)
-z = FrankeFunction(x, y)
-# z = SkrankeFunction(x, y)
-
-# Highest order polynomial we fit with
-N = 30
-lam = 0.001
-
-z += 0.05 * np.random.standard_normal(z.shape)
-X, X_train, X_test, z_train, z_test = preprocess(x, y, z, N, 0.2)
-
-lambdas = np.logspace(-4, 4, 6)
-for i in range(len(lambdas)):
-    plt.subplot(321 + i)
-    betas, z_preds_train, z_preds_test, _ = linreg_to_N(
-        X,
-        X_train,
-        X_test,
-        z_train,
-        z_test,
-        N,
-        scaling=True,
-        model=ridge,
-        lam=lambdas[i],
-    )
-    plt.plot(betas[0, :], label="beta0")
-    plt.plot(betas[1, :], label="beta1")
-    plt.plot(betas[2, :], label="beta2")
-    plt.plot(betas[3, :], label="beta3")
-    plt.plot(betas[4, :], label="beta4")
-    plt.plot(betas[5, :], label="beta5")
-    plt.title(f"Beta progression for lambda = {lambdas[i]:.5}")
-    plt.legend()
-
-plt.show()
-
-# linear_regression(x,y,z)

### 2022-10-07 src/task_e_msecomparison.py

@@ -1,104 +1,0 @@
-from mpl_toolkits.mplot3d import Axes3D
-import matplotlib.pyplot as plt
-from matplotlib import cm
-from matplotlib.ticker import LinearLocator, FormatStrFormatter
-import numpy as np
-from random import random, seed
-from sklearn.linear_model import Ridge
-from sklearn.model_selection import train_test_split, cross_validate, KFold
-
-# Our own library of functions
-from utils import *
-
-np.random.seed(42069)
-# Make data.
-x = np.arange(0, 1, 0.05)
-y = np.arange(0, 1, 0.05)
-x, y = np.meshgrid(x, y)
-z = FrankeFunction(x, y)
-# z = SkrankeFunction(x, y)
-scaling = False
-
-# Highest order polynomial we fit with
-N = 10
-bootstraps = 100
-K = 10
-kfolds = KFold(n_splits=K)
-
-# Do the linear_regression
-z += 0.05 * np.random.standard_normal(z.shape)
-X, X_train, X_test, z_train, z_test = preprocess(x, y, z, N, 0.2)
-z = z.ravel()
-
-lambdas = np.logspace(-12, -4, 6)
-for i in range(len(lambdas)):
-    plt.subplot(321 + i)
-    plt.suptitle(f"MSE by polynomial degree for different values of lambda")
-
-    errors_cv = np.zeros(N)
-    errors_cv_scikit = np.zeros(N)
-    errors_boot = np.zeros(N)
-
-    Ridge_model = Ridge(lambdas[i], fit_intercept=scaling)
-    print("Cross")
-    for n in range(N):
-        print(n)
-        l = int((n + 1) * (n + 2) / 2)  # Number of elements in beta
-        errors_cv[n] = crossval(X[:, :l], z, K, scaling=scaling)
-
-        error_scikit = cross_validate(
-            estimator=Ridge_model,
-            X=X[:, :l],
-            y=z,
-            scoring="neg_mean_squared_error",
-            cv=kfolds,
-        )
-        errors_cv_scikit[n] = np.mean(-error_scikit["test_score"])
-
-    for n in range(N):
-        print(n)
-        l = int((n + 1) * (n + 2) / 2)  # Number of elements in beta
-        z_preds = bootstrap(
-            X[:, :l],
-            X_train[:, :l],
-            X_test[:, :l],
-            z_train,
-            z_test,
-            bootstraps,
-            scaling=True,
[…]

### 2022-10-07 src/task_f_biasvariance.py

@@ -1,117 +1,0 @@
-from mpl_toolkits.mplot3d import Axes3D
-import matplotlib.pyplot as plt
-from matplotlib import cm
-from matplotlib.ticker import LinearLocator, FormatStrFormatter
-import numpy as np
-from random import random, seed
-from sklearn.model_selection import train_test_split
-from sklearn.linear_model import Lasso, Ridge
-
-
-# Our own library of functions
-from utils import *
-
-np.random.seed(42069)
-# Make data.
-x = np.arange(0, 1, 0.15)
-y = np.arange(0, 1, 0.15)
-# x = np.arange(0, 1, 0.05)
-# y = np.arange(0, 1, 0.05)
-x, y = np.meshgrid(x, y)
-z = FrankeFunction(x, y)
-# z = SkrankeFunction(x, y)
-
-# Highest order polynomial we fit with
-N = 12
-bootstraps = 100
-
-# Do the linear_regression
-z += 0.15 * np.random.standard_normal(z.shape)
-X, X_train, X_test, z_train, z_test = preprocess(x, y, z, N, 0.2)
-# X, X_train, X_test, z, z_train, z_test = normalize_task_g(
-#     X, X_train, X_test, z, z_train, z_test
-# )
-
-lambdas = np.logspace(-6, -1, 6)
-lambdas[-1] = 0.0006951
-for i in range(len(lambdas)):
-    plt.subplot(321 + i)
-    plt.suptitle(f"Bias variance tradeoff for different values of lambda")
-    model_Lasso = Lasso(lambdas[i], max_iter=500, fit_intercept=False)
-
-    errors = np.zeros(N)
-    biases = np.zeros(N)
-    variances = np.zeros(N)
-
-    for n in range(N):
-        print(n)
-        l = int((n + 1) * (n + 2) / 2)  # Number of elements in beta
-        z_preds = bootstrap(
-            X[:, :l],
-            X_train[:, :l],
-            X_test[:, :l],
-            z_train,
-            z_test,
-            bootstraps,
-            model=model_Lasso,
-            lam=lambdas[i],
-            scaling=True,
-        )
-
-        error, bias, variance = bias_variance(z_test, z_preds)
-        errors[n] = error
-        biases[n] = bias
-        variances[n] = variance
-
-    # variances *= 10
[…]

### 2022-10-07 doc/paper.tex
Remove paper directory

--- 2022-10-07 doc/paper/paper.tex
@@
 
-We use the Franke Function with an added stochastic noise of to test our models. We explore how higher and lower levels of noise affect the model.
+We use the Franke Function with an added stochastic noise of to test our models. We explore how higher and lower levels of noise affect the model. Below is the Franke function:
+
+\begin{multline}
+f(x,y) = \frac{3}{4}\exp{\left(-\frac{(9x-2)^2}{4}-\frac{(9y-2)^2}{4}\right)} + \\ \frac{3}{4}\exp{\left(-\frac{(9x+1)^2}{49}-\frac{(9y+1)}{10}\right)} + \\
+\frac{1}{2}\exp{\left(-\frac{(9x-7)^2}{4}-\frac{(9y-3)^2}{4}\right)} - \\
+\frac{1}{5}\exp{\left(-(9x-4)^2-(9y-7)^2\right)}
+\end{multline}
+
+As we can see, it is simply a weighted sum of exponentials. As such it has an infinite Taylor expansion into polynomials.\footnote{The proof for this is left as an exercise for the reader}
 
@@
 We find that OLS creates a good fit for the Franke Function.
-We can increase our polynomial degree to around 20 before we see signs of overfitting.\footnote{Here we also see something a bit surprising: We see that our own implementation of OLS follows Scikit's implementation up to around polynomial degree 10, at which point scikit's MSE increases rapidly.
+With 400 datapoints, we can increase our polynomial degree to around 20 before we see signs of overfitting.\footnote{Here we also see something a bit surprising: We see that our own implementation of OLS follows Scikit's implementation up to around polynomial degree 10, at which point scikit's MSE increases rapidly.
 We chalk this up to numerical errors and differing implementation of the normal equations, as we simply use the normal equation exactly, while scikit uses an implementation which is probably computationally more efficient but gives slight differences when comparing to the normal equation.
-These differences seem to create a large error after a certain size of the design matrix is reached, which we can not adequately explain.} This is most likely because the Franke function is a sum of exponents, which can be expanded into polynomials.
+These differences seem to create a large error after a certain size of the design matrix is reached, which we can not adequately explain.} This is most likely because the Franke function can be expanded into a sum of polynomials.
 We would expect to see higher polynomial degrees give us better and better prediction, up to a certain point.\footnote{And indeed, this seems to be the case, as if we increase the availiable datapoints considerably, as seen in Fig.~\ref{frankiehigherdatapoints} in the appendix, we can reach much higher polynomial degrees without overfitting} When we reach degrees higher than around 20, we see that our test error increases slightly, a sign of overfitting.
@@
 Clearly, we need to introduce regularization if we want a model that does not overfit.
+Fig.
+~\ref{frankenoisybeta} in the appendix shows the beta progression for selected betas, clearly showing that we are overfitting.
 
@@
 \begin{tabular}{c l l}
-& & \\ % put some space after the caption
-\hline
+% & & \\ % put some space after the caption
+% \hline
 Std of noise & Best MSE & Best degree \\
@@
 Using Ridge regression, we see from Fig.~\ref{frankiedifflambdas} that larger and larger values of $\lambda$ efficiently prevents overfitting and reduces the variance between predictions.
+Looking at the beta progression with for the same values of $\lambda$ (Fig.~\ref{frankeridgebetas}), we see that the regularization parameter is successfully shrinking the beta values.
 Using Scikit's gridsearch, we find the optimal $\lambda$ to be $10^{-8}$ and the optimal polynomial degree to be 7, with an average MSE of $0.0037999$, which is very slightly lower than OLS for the Franke function with an added stochastic noise $\epsilon \sim N(0,0.05)$.
@@
 
+\subsubsection{Lasso}
+
+Looking at Lasso regression, we find that it has the same capability to reduce overfitting as Ridge, as can be seen in Fig.~\ref{frankelassodifflambdas}.
+
 \begin{table}[t]
@@
 \begin{tabular}{c | l l l}
-& & \\ % put some space after the caption
-\hline
+% & & \\ % put some space after the caption
+% \hline
 Regression & Std of noise & Best MSE & Best degree \\
@@
 
-% All figures should be positioned at the top of the page where possible.
-All figures should be numbered consecutively and centered under the figure as shown in Fig.~\ref{figure_ASME}.
-All text within the figure should be no smaller than 7~pt.
-There should be a minimum two line spaces between figures and text.
-The number of a referenced figure or table in the text should be preceded by Fig.\ or Tab.\ respectively unless the reference starts a sentence in which case Fig.\ or Tab.\ should be expanded to Figure or Table.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \begin{tabular}{c | l l l l l l l}
-& & \\ % put some space after the caption
-\hline
+% & & \\ % put some space after the caption
+% \hline
 Figure & Highest $N$ & Std of noise & Number of datapoints & Bootstraps & K-folds & $\lambda$ & Scaling \\

### 2022-10-07 src/task_b.py
fix plots

@@ -56,7 +56,7 @@ ax.zaxis.set_major_locator(LinearLocator(10))
 ax.zaxis.set_major_formatter(FormatStrFormatter("%.02f"))
 ax.set_title("Scaled terrain")
 # Add a color bar which maps values to colors.
-fig.colorbar(surf, shrink=0.5, aspect=5)
+#fig.colorbar(surf_real, shrink=0.5, aspect=5)
 
 # Subplot for the prediction
 # Plot the surface.

### 2022-10-07 src/task_e_betas.py

@@ -38,8 +38,14 @@ for i in range(len(lambdas)):
 
     # plot beta values
     if betas_to_plot <= betas.shape[0]:
-        for col in range(betas_to_plot):
-            plt.plot(betas[col, :], label=f"beta{col}", marker="o", markersize=3)
+        for beta in range(betas_to_plot):
+            data = betas[beta, :]
+            data[data == 0] = np.nan
+            plt.plot(data, label=f"beta{beta}", marker="o", markersize=3)
+            plt.xlabel("Polynomial degree (N)")
+            plt.ylabel("Beta value")
+            plt.title("Beta progression")
+            plt.legend()
 
     plt.title(f"Beta progression for lambda = {lambdas[i]:.5}")
     plt.legend()

### 2022-10-07 doc/paper.tex

@@
 
-\lstinputlisting[language=Python, firstline=27, lastline=42]{../../src/utils.py}
+\lstinputlisting[language=Python, firstline=27, lastline=42]{../src/utils.py}
 
@@
 
-\lstinputlisting[language=Python, firstline=238, lastline=245]{../../src/utils.py}
+\lstinputlisting[language=Python, firstline=238, lastline=245]{../src/utils.py}
 
@@
 
-\lstinputlisting[language=Python, firstline=53, lastline=64]{../../src/utils.py}
+\lstinputlisting[language=Python, firstline=53, lastline=64]{../src/utils.py}
 

### 2022-10-07 src/task_b.py
fix data, add figs

@@ -8,8 +8,8 @@ from utils import *
 np.random.seed(42069)
 
 # parameters
-N = 20
-betas_to_plot = 9
+N = 25
+betas_to_plot = 10
 # parameters for synthetic data
 noise = 0.05
 scaling = False
@@ -93,9 +93,9 @@ if betas_to_plot <= betas.shape[0]:
         data = betas[beta, :]
         data[data == 0] = np.nan
         plt.plot(data, label=f"beta{beta}", marker="o", markersize=3)
-        plt.xlabel("Polynomial degree (N)")
-        plt.ylabel("Beta value")
-        plt.title("Beta progression")
+        plt.xlabel("Polynomial degree (N)", size=15)
+        plt.ylabel("Beta value", size=15)
+        plt.title("Beta progression", size=22)
         plt.legend()
     plt.show()
 

### 2022-10-07 src/task_e_betas.py

@@ -7,9 +7,8 @@ from utils import *
 np.random.seed(42069)
 
 # Parameters
-N = 20
-lambdas = np.logspace(-15, -9, 6)
-lambdas[-1] = 10000000000
+N = 10
+lambdas = np.logspace(-10, 0, 4)
 betas_to_plot = 9
 
 # Parameters for synthetic data
@@ -18,12 +17,13 @@ scaling = False
 
 # get data
 X, X_train, X_test, z, z_train, z_test, scaling, x, y, z = read_in_dataset(
-    N, scaling, noise
+    N, scaling=scaling, noise=noise, step=0.1,
 )
 
 # calculate beta values
 for i in range(len(lambdas)):
-    plt.subplot(321 + i)
+    plt.subplot(411 + i)
+    plt.suptitle(f"Beta progression for different lambdas")
     betas, z_preds_train, z_preds_test, _ = linreg_to_N(
         X,
         X_train,
@@ -47,7 +47,7 @@ for i in range(len(lambdas)):
             plt.title("Beta progression")
             plt.legend()
 
-    plt.title(f"Beta progression for lambda = {lambdas[i]:.5}")
+    plt.title(f"lambda = {lambdas[i]:.5}")
     plt.legend()
 
 plt.show()

### 2022-10-07 src/task_e_biasvariance.py

@@ -11,15 +11,15 @@ np.random.seed(42069)
 K = 20
 N = 10
 bootstraps = 100
-plot_only_best_lambda = True
-lambdas = np.logspace(-10, 0, 6)
+plot_only_best_lambda = False
+lambdas = np.logspace(-10, 0, 4)
 # parameter synthetic data
 noise = 0.05
 scaling = False
 
 # read in data
 X, X_train, X_test, z, z_train, z_test, scaling, x, y, z = read_in_dataset(
-    N, scaling, noise
+    N, scaling=scaling, noise=noise, step=0.1,
 )
 z = z.ravel()
 
@@ -32,7 +32,7 @@ if plot_only_best_lambda:
 # for lambdas
 for i in range(len(lambdas)):
     if not plot_only_best_lambda:
-       plt.subplot(321 + i)
+       plt.subplot(411 + i)
        plt.suptitle(f"Bias variance tradeoff for ridge regression")
 
     # results
@@ -66,6 +66,7 @@ for i in range(len(lambdas)):
     plt.plot(errors, label="MSE test")
     plt.plot(biases, label="bias")
     plt.plot(variances, label="variance")
+    plt.ylim(0,0.12)
     plt.xlabel("Polynomial degree (N)")
     plt.tight_layout(h_pad=0.001)
     if plot_only_best_lambda:

### 2022-10-07 src/utils.py

@@ -342,7 +342,7 @@ def find_best_lambda(X, z, model, lambdas, N, K):
     return best_lambda, best_MSE, best_polynomial
 
 
-def read_in_dataset(N, scaling, noise):
+def read_in_dataset(N, *, scaling=False, noise=0.05, step=0.05):
     argv = sys.argv[1:]
 
     parser = argparse.ArgumentParser(description="Compute task g.b")
@@ -370,8 +370,8 @@ def read_in_dataset(N, scaling, noise):
         )
     else:
         # create synthetic data
-        x = np.arange(0, 1, 0.05)
-        y = np.arange(0, 1, 0.05)
+        x = np.arange(0, 1, step)
+        y = np.arange(0, 1, step)
         x, y = np.meshgrid(x, y)
         z = FrankeFunction(x, y)
 

### 2022-10-07 src/task_e_betas.py
figures

@@ -13,7 +13,7 @@ betas_to_plot = 9
 
 # Parameters for synthetic data
 noise = 0.05
-scaling = False
+scaling = True
 
 # get data
 X, X_train, X_test, z, z_train, z_test, scaling, x, y, z = read_in_dataset(
@@ -42,12 +42,10 @@ for i in range(len(lambdas)):
             data = betas[beta, :]
             data[data == 0] = np.nan
             plt.plot(data, label=f"beta{beta}", marker="o", markersize=3)
-            plt.xlabel("Polynomial degree (N)")
-            plt.ylabel("Beta value")
-            plt.title("Beta progression")
-            plt.legend()
-
+    plt.xlabel("Polynomial degree (N)")
+    plt.ylabel("Beta value")
+    plt.title("Beta progression")
     plt.title(f"lambda = {lambdas[i]:.5}")
-    plt.legend()
+    plt.legend(loc='center left', bbox_to_anchor=(1, 0.5))
 
 plt.show()

### 2022-10-07 src/task_e_biasvariance.py

@@ -15,7 +15,7 @@ plot_only_best_lambda = False
 lambdas = np.logspace(-10, 0, 4)
 # parameter synthetic data
 noise = 0.05
-scaling = False
+scaling = True
 
 # read in data
 X, X_train, X_test, z, z_train, z_test, scaling, x, y, z = read_in_dataset(

### 2022-10-07 src/task_f_biasvariance.py

@@ -12,20 +12,20 @@ np.random.seed(42069)
 N = 20
 K = 20
 bootstraps = 100
-plot_only_best_lambda = False
+plot_only_best_lambda = True
 lambdas = np.logspace(-12, -4, 6)
 # synthetic parameters
 noise = 0.05
-scaling = False
+scaling = True
 
 # read in and get data
-X, X_train, X_test, z, z_train, z_test, scaling, x, y, z = read_in_dataset(N, scaling, noise)
+X, X_train, X_test, z, z_train, z_test, scaling, x, y, z = read_in_dataset(N, scaling=scaling, noise=noise)
 z = z.ravel()
 
 # plot only the gridsearched lambda
 if plot_only_best_lambda:
     lasso = Lasso(fit_intercept=False)
-    lam, _, _ = find_best_lambda(X, z, lasso, lambdas, N, K)
+    lam, best_MSE, best_poly = find_best_lambda(X, z, lasso, lambdas, N, K)
     lambdas = [lam]
 
 # loop through different lambda values
@@ -35,7 +35,7 @@ for i in range(len(lambdas)):
            plt.suptitle(f"Bias variance tradeoff for lasso regression")
 
     # model under testing
-    model_Lasso = Lasso(lambdas[i], max_iter=200, fit_intercept=False)
+    model_Lasso = Lasso(lambdas[i], max_iter=1000, fit_intercept=False)
 
     # arrays for bias-variance
     errors = np.zeros(N)
@@ -71,10 +71,11 @@ for i in range(len(lambdas)):
     plt.xlabel("Polynomial degree (N)")
     plt.tight_layout(h_pad=0.001)
     if plot_only_best_lambda:
-        plt.title(f"Bias variance tradeoff for lasso regression for optimal lambda = {lambdas[i]}")
+        plt.title(f"Bias variance tradeoff for lasso regression \n for optimal lambda = {lambdas[i]}")
     else:
         plt.title(f"lambda = {lambdas[i]:.5}")
     plt.legend()
 
 
+print(f"Optimal lambda = {lam}, best MSE = {best_MSE}, best polynomial = {best_poly}")
 plt.show()

### 2022-10-07 src/task_f_msecomparison.py

@@ -3,7 +3,6 @@ task f: Plots MSE comparison between OLS, ridge and lasso with cross validation 
 """
 from sklearn.linear_model import Ridge, Lasso
 from sklearn.model_selection import cross_validate, KFold
-from imageio import imread
 
 # Our own library of functions
 from utils import *

### 2022-10-07 doc/paper.tex
More paper

@@
 {\it We use Ordinary Least Squares, Ridge and Lasso regression on a synthetic dataset generated by the Franke Function, and on a real dataset of elevation data from a terrain in Norway.
-We explore different regularization values and polynomial degrees and find that that MODEL performs best, with a MSE of NUMBER for polynomial degree 6.
+We explore different regularization values and polynomial degrees and find that that all methods perform similarly on both sets of data.
 We find that linear regression performs well on the synthetic data, but falls short when approximating elevation data with large variances in height.
@@
 
+% \begin{figure} 
+% \centerline{\includegraphics[width=3.25in]{figure/frankenoisybeta.png}}
+% \caption{Bias, variance and test MSE plotted over polynomial degree for our OLS prediction on the Franke function with an added stochastic noise $\epsilon \sim N(0,0.05)$}
+% \label{frankenoisybeta}
+% \end{figure}
+
 Table.
@@
 
-\subsubsection{Lasso}
-
-Looking at Lasso regression, we find that it has the same capability to reduce overfitting as Ridge, as can be seen in Fig.~\ref{frankelassodifflambdas}.
-
 \begin{table}[t]
@@
 \label{ols_vs_ridge_table}
-\begin{tabular}{c | l l l}
+\begin{tabular}{c | l l l l}
 % & & \\ % put some space after the caption
 % \hline
-Regression & Std of noise & Best MSE & Best degree \\
+Regression & Std of noise & MSE & Degree & $\lambda$ \\
 \hline
-OLS & 0.05 & 0.0038302 & 7 \\
-Ridge & 0.05 & 0.0037999 & 7 \\
+OLS & 0.05 & 0.0038302 & 7 & - \\
+Ridge & 0.05 & 0.0037999 & 7 & $10^{-8}$ \\
+\hline
+\end{tabular}
+\end{center}
+\end{table}
+
+\subsubsection{Lasso}
+
+Looking at Lasso regression, we find that it has the same capability to reduce overfitting as Ridge, as can be seen in Fig.~\ref{frankelassodifflambdas}.
+However, looking at Table.~\ref{ols_vs_ridge_vs_lasso_table}, we see that Lasso falls short when finding a good MSE.
+Intuitively, it does not make sense for Lasso to be worse than OLS, because Lasso with $\lambda = 0$ is equivalent to OLS.
+But we cannot avoid considering the fact that Lasso does not have an analytical solution, and as such we must use numerical methods to find the optimal parameters.
+Using scikit's implementation, we have not been able to get the numerical method to converge, which means that we have not found a global minimum.\footnote{Scikit explicitly mentions that using $\lambda = 0$ does not converge well with their implementation} For this reason, Lasso regression has not given satisfying results.
+
+\begin{table}[t]
+\caption{Best MSE for OLS, Ridge and Lasso with 400 datapoints}
+\begin{center}
+\label{ols_vs_ridge_vs_lasso_table}
+\begin{tabular}{c | l l l l}
+% & & \\ % put some space after the caption
+% \hline
+Regression & Std of noise & MSE & Degree & $\lambda$ \\
+\hline
+OLS & 0.05 & 0.0038302 & 7 & - \\
+Ridge & 0.05 & 0.0037999 & 7 & $10^{-8}$ \\
+Lasso & 0.05 & 0.0095846 & 10 & $2.5118 \cdot 10^{-6}$ \\
 \hline
@@
 
-\subsubsection{Lasso}
-
-Turning our attention to Lasso regression, we also find that overfitting can be avoided with a good choice of $\lambda$.
-Comparing the three methods, we see that for noisy target data, Ridge and Lasso are superior to OLS, as seen in Fig.
-NUMBER.
+\begin{figure} 
+\centerline{\includegraphics[width=3.25in]{figure/frankeridgebetas.png}}
+\caption{Beta progression for different values of lambda in Ridge regression. Note the different scales on each of the plots, showing how increasing the regularization parameter shrinks the betas}
+\label{frankeridgebetas}
+\end{figure}
 
@@
 \ref{frankiedifflambdas} & 25 & 0.05 & 100 & 100 & - & variable & Centering\\
+\ref{frankeridgebetas} & 25 & 0.05 & 100 & - & - & variable & Centering\\
+\ref{frankenoisybeta} & 25 & 0.05 & 400 & - & - & - & Centering\\
 \ref{frankienoisecomparison} & 25 & variable & 400 & 100 & - & - & None\\
@@
 
+\begin{figure*}
+% \begin{figure}[t]
+\centerline{\includegraphics[width=3.25in]{figure/frankenoisybeta.png}}
+\caption{Beta progression for OLS}
+\label{frankenoisybeta}
+\end{figure*}
+
 \end{document}

### 2022-10-07 src/task_f_biasvariance.py
add lasso lambda fig

@@ -9,11 +9,11 @@ from utils import *
 np.random.seed(42069)
 
 # parameters
-N = 20
 K = 20
+N = 10
 bootstraps = 100
-plot_only_best_lambda = True
-lambdas = np.logspace(-12, -4, 6)
+plot_only_best_lambda = False
+lambdas = np.logspace(-10, 0, 4)
 # synthetic parameters
 noise = 0.05
 scaling = True
@@ -31,7 +31,7 @@ if plot_only_best_lambda:
 # loop through different lambda values
 for i in range(len(lambdas)):
     if not plot_only_best_lambda:
-           plt.subplot(321 + i)
+           plt.subplot(411 + i)
            plt.suptitle(f"Bias variance tradeoff for lasso regression")
 
     # model under testing
@@ -71,11 +71,11 @@ for i in range(len(lambdas)):
     plt.xlabel("Polynomial degree (N)")
     plt.tight_layout(h_pad=0.001)
     if plot_only_best_lambda:
+        print(f"Optimal lambda = {lam}, best MSE = {best_MSE}, best polynomial = {best_poly}")
         plt.title(f"Bias variance tradeoff for lasso regression \n for optimal lambda = {lambdas[i]}")
     else:
         plt.title(f"lambda = {lambdas[i]:.5}")
     plt.legend()
 
 
-print(f"Optimal lambda = {lam}, best MSE = {best_MSE}, best polynomial = {best_poly}")
 plt.show()

### 2022-10-10 doc/paper.tex
More paper stuff

@@
 
+The correctness of our linear regression can be confirmed by running our debug function (the so called SkrankeFunction)\footnote{this can be done by running \texttt{\$ python3 task\_b.py --debug -n 2}}, which is a second degree polynomial where $b_0 = 0, \; b_1 = 1, \; b_2 = 2$ and so on.
+
 \subsubsection{Implementation of resampling methods}
@@
 But we should recall that we decide the complexity of our model, by deciding which polynomial degree we use when creating the design matrix.
-In fact, by choosing polynomial degree 7 for our OLS method to get the best MSE for noisy data, we are in fact doing a primitive form of best subset selection, as described in Hastie et Al.
-p 57 \cite{hastie}.
+In fact, by choosing polynomial degree 7 for our OLS method to get the best MSE for noisy data, we are in fact doing a primitive form of best subset selection, as described in Hastie et Al.\cite[57]{hastie}.
 Almost by definition, the polynomial degree with the best MSE is not overfitted, because if it was, it would not be the model with the lowest MSE among all polynomial degrees.
@@
 
-\bibliographystyle{alpha}
+\bibliographystyle{apalike}
 \bibliography{bibliography}

### 2022-10-10 src/noise_plot.py
Add cli parser, change scaling to centering

@@ -21,10 +21,10 @@ zs = SkrankeFunction(x, y)
 # parameters
 N = 10
 noises = np.linspace(0, 0.2, 5)
-scaling = False
+centering = False
 bootstraps = 100
 
-def noise_plot(x, y, z, N, noise, scaling, bootstraps):
+def noise_plot(x, y, z, N, noise, centering, bootstraps):
     # add noise
     z += noise * np.random.standard_normal(z.shape)
     X, X_train, X_test, z_train, z_test = preprocess(x, y, z, N, 0.2)
@@ -43,7 +43,7 @@ def noise_plot(x, y, z, N, noise, scaling, bootstraps):
             z_train,
             z_test,
             bootstraps,
-            scaling=scaling,
+            centering=centering,
         )
 
         MSE_test, _ = scores(z_test, z_preds_test)
@@ -53,7 +53,7 @@ def noise_plot(x, y, z, N, noise, scaling, bootstraps):
 
 
 for noise in noises:
-    test_errors = noise_plot(x, y, z, N, noise, scaling, bootstraps)
+    test_errors = noise_plot(x, y, z, N, noise, centering, bootstraps)
     # plot
     plt.plot(
         test_errors,

### 2022-10-10 src/task_b.py

@@ -7,17 +7,21 @@ from utils import *
 
 np.random.seed(42069)
 
-# parameters
-N = 25
-betas_to_plot = 10
-# parameters for synthetic data
-noise = 0.05
-scaling = False
-
 # get data
-X, X_train, X_test, z, z_train, z_test, scaling, x, y, z = read_in_dataset(
-    N, scaling=scaling, noise=noise
-)
+(
+    betas_to_plot,
+    N,
+    X,
+    X_train,
+    X_test,
+    z,
+    z_train,
+    z_test,
+    centering,
+    x,
+    y,
+    z,
+) = read_from_cmdline()
 
 # implemented model under testing
 OLS_model = OLS
@@ -26,12 +30,12 @@ OLS_scikit = LinearRegression(fit_intercept=False)
 
 # perform linear regression
 betas, z_preds_train, z_preds_test, z_preds = linreg_to_N(
-    X, X_train, X_test, z_train, z_test, N, scaling=scaling, model=OLS_model
+    X, X_train, X_test, z_train, z_test, N, centering=centering, model=OLS_model
 )
 
 # perform linear regression scikit
 _, z_preds_train_sk, z_preds_test_sk, _ = linreg_to_N(
-    X, X_train, X_test, z_train, z_test, N, scaling=scaling, model=OLS_scikit
+    X, X_train, X_test, z_train, z_test, N, centering=centering, model=OLS_scikit
 )
 
 # Calculate OLS scores
@@ -56,7 +60,7 @@ ax.zaxis.set_major_locator(LinearLocator(10))
 ax.zaxis.set_major_formatter(FormatStrFormatter("%.02f"))
 ax.set_title("Scaled terrain")
 # Add a color bar which maps values to colors.
-#fig.colorbar(surf_real, shrink=0.5, aspect=5)
+# fig.colorbar(surf_real, shrink=0.5, aspect=5)
 
 # Subplot for the prediction
 # Plot the surface.
@@ -88,16 +92,15 @@ plt.imshow(pred_map)
 plt.colorbar()
 plt.show()
 
-if betas_to_plot <= betas.shape[0]:
-    for beta in range(betas_to_plot):
-        data = betas[beta, :]
-        data[data == 0] = np.nan
-        plt.plot(data, label=f"beta{beta}", marker="o", markersize=3)
-        plt.xlabel("Polynomial degree (N)", size=15)
-        plt.ylabel("Beta value", size=15)
-        plt.title("Beta progression", size=22)
[…]

### 2022-10-10 src/task_c.py

@@ -11,10 +11,10 @@ N = 20
 bootstraps = 100
 # parameters for synthetic data
 noise = 0.05
-scaling = False
+centering = False
 
 # get data
-X, X_train, X_test, z, z_train, z_test, scaling, x, y, z = read_in_dataset(N, scaling, noise)
+X, X_train, X_test, z, z_train, z_test, centering, x, y, z = read_in_dataset(N, centering, noise)
 
 # result arrays
 errors = np.zeros(N)
@@ -37,7 +37,7 @@ for n in range(N):
         z_train,
         z_test,
         bootstraps,
-        scaling=scaling,
+        centering=centering,
         model=OLS_model,
     )
 

### 2022-10-10 src/task_d.py

@@ -14,11 +14,11 @@ K = 10
 bootstraps = 100
 # parameters for synthetic data
 noise = 0.05
-scaling = False
+centering = False
 
 # read in data
-X, X_train, X_test, z, z_train, z_test, scaling, x, y, z = read_in_dataset(
-    N, scaling, noise
+X, X_train, X_test, z, z_train, z_test, centering, x, y, z = read_in_dataset(
+    N, centering, noise
 )
 z = z.ravel()
 
@@ -45,7 +45,7 @@ for n in range(N):
     l = int((n + 1) * (n + 2) / 2)  # Number of elements in beta
 
     # own implementation
-    errors_cv[n] = crossval(X[:, :l], z, K, scaling=scaling, model=OLS_model)
+    errors_cv[n] = crossval(X[:, :l], z, K, centering=centering, model=OLS_model)
 
     # scikit
     error_scikit = cross_validate(
@@ -70,7 +70,7 @@ for n in range(N):
         z_train,
         z_test,
         bootstraps,
-        scaling=scaling,
+        centering=centering,
         model=OLS_model,
     )
 
@@ -86,9 +86,7 @@ plt.plot(errors_cv, label="cross validation implementation")
 plt.plot(errors_cv_scikit, label="cross validation scikit learn")
 plt.ylabel("MSE score")
 plt.xlabel("Polynomial degree (N)")
-plt.title(
-    f"MSE by Resampling Method"
-)
+plt.title(f"MSE by Resampling Method")
 plt.legend()
 
 plt.show()

### 2022-10-10 src/task_e_betas.py

@@ -13,11 +13,14 @@ betas_to_plot = 9
 
 # Parameters for synthetic data
 noise = 0.05
-scaling = True
+centering = True
 
 # get data
-X, X_train, X_test, z, z_train, z_test, scaling, x, y, z = read_in_dataset(
-    N, scaling=scaling, noise=noise, step=0.1,
+X, X_train, X_test, z, z_train, z_test, centering, x, y, z = read_in_dataset(
+    N,
+    centering=centering,
+    noise=noise,
+    step=0.1,
 )
 
 # calculate beta values
@@ -31,7 +34,7 @@ for i in range(len(lambdas)):
         z_train,
         z_test,
         N,
-        scaling=scaling,
+        centering=centering,
         model=ridge,
         lam=lambdas[i],
     )
@@ -46,6 +49,6 @@ for i in range(len(lambdas)):
     plt.ylabel("Beta value")
     plt.title("Beta progression")
     plt.title(f"lambda = {lambdas[i]:.5}")
-    plt.legend(loc='center left', bbox_to_anchor=(1, 0.5))
+    plt.legend(loc="center left", bbox_to_anchor=(1, 0.5))
 
 plt.show()

### 2022-10-10 src/task_e_biasvariance.py

@@ -2,6 +2,7 @@
 task e (and task g): plots bias-variance trade-off for ridge regression
 """
 from sklearn.linear_model import Ridge
+
 # Our own library of functions
 from utils import *
 
@@ -15,11 +16,14 @@ plot_only_best_lambda = False
 lambdas = np.logspace(-10, 0, 4)
 # parameter synthetic data
 noise = 0.05
-scaling = True
+centering = True
 
 # read in data
-X, X_train, X_test, z, z_train, z_test, scaling, x, y, z = read_in_dataset(
-    N, scaling=scaling, noise=noise, step=0.1,
+X, X_train, X_test, z, z_train, z_test, centering, x, y, z = read_in_dataset(
+    N,
+    centering=centering,
+    noise=noise,
+    step=0.1,
 )
 z = z.ravel()
 
@@ -32,8 +36,8 @@ if plot_only_best_lambda:
 # for lambdas
 for i in range(len(lambdas)):
     if not plot_only_best_lambda:
-       plt.subplot(411 + i)
-       plt.suptitle(f"Bias variance tradeoff for ridge regression")
+        plt.subplot(411 + i)
+        plt.suptitle(f"Bias variance tradeoff for ridge regression")
 
     # results
     errors = np.zeros(N)
@@ -51,7 +55,7 @@ for i in range(len(lambdas)):
             z_train,
             z_test,
             bootstraps,
-            scaling=scaling,
+            centering=centering,
             model=ridge,
             lam=lambdas[i],
         )
@@ -66,13 +70,16 @@ for i in range(len(lambdas)):
     plt.plot(errors, label="MSE test")
     plt.plot(biases, label="bias")
     plt.plot(variances, label="variance")
-    plt.ylim(0,0.12)
+    plt.ylim(0, 0.12)
     plt.xlabel("Polynomial degree (N)")
     plt.tight_layout(h_pad=0.001)
     if plot_only_best_lambda:
-        plt.title(f"Bias variance tradeoff for ridge regression for optimal lambda = {lambdas[i]}")
+        plt.title(
+            f"Bias variance tradeoff for ridge regression for optimal lambda = {lambdas[i]}"
+        )
     else:
         plt.title(f"lambda = {lambdas[i]:.5}")
     plt.legend()
 
 plt.show()
+

### 2022-10-10 src/task_e_manualsearch.py

@@ -43,7 +43,7 @@ for i in range(num_lambdas):
         z_train,
         z_test,
         bootstraps,
-        scaling=True,
+        centering=True,
         model=ridge,
         lam=lambdas[i],
     )

### 2022-10-10 src/task_e_msecomparison.py

@@ -14,7 +14,7 @@ N = 20
 K = 20
 bootstraps = 100
 noise = 0.05
-scaling = True
+centering = True
 kfolds = KFold(n_splits=K)
 
 # if true, plot only lambda that gives lowest MSE
@@ -23,8 +23,8 @@ plot_only_best_lambda = True
 lambdas = np.logspace(-12, -4, 6)
 
 # read in data
-X, X_train, X_test, z, z_train, z_test, scaling, x, y, z = read_in_dataset(
-    N, scaling, noise
+X, X_train, X_test, z, z_train, z_test, centering, x, y, z = read_in_dataset(
+    N, centering, noise
 )
 z = z.ravel()
 
@@ -79,7 +79,7 @@ for i in range(len(lambdas)):
             z_train,
             z_test,
             bootstraps,
-            scaling=scaling,
+            centering=centering,
             model=model,
             lam=lambdas[i],
         )

### 2022-10-10 src/task_f_biasvariance.py

@@ -16,10 +16,10 @@ plot_only_best_lambda = False
 lambdas = np.logspace(-10, 0, 4)
 # synthetic parameters
 noise = 0.05
-scaling = True
+centering = True
 
 # read in and get data
-X, X_train, X_test, z, z_train, z_test, scaling, x, y, z = read_in_dataset(N, scaling=scaling, noise=noise)
+X, X_train, X_test, z, z_train, z_test, centering, x, y, z = read_in_dataset(N, centering=centering, noise=noise)
 z = z.ravel()
 
 # plot only the gridsearched lambda
@@ -53,7 +53,7 @@ for i in range(len(lambdas)):
             z_train,
             z_test,
             bootstraps,
-            scaling=scaling,
+            centering=centering,
             model=model_Lasso,
             lam=lambdas[i],
         )

### 2022-10-10 src/task_f_msecomparison.py

@@ -11,7 +11,7 @@ from utils import *
 N = 20
 K = 10
 noise = 0.05
-scaling = False
+centering = False
 lambdas = np.logspace(-12, -4, 6)
 
 kfolds = KFold(n_splits=K)
@@ -21,7 +21,9 @@ errors_Ridge = np.zeros(N)
 errors_Lasso = np.zeros(N)
 
 # read in data
-X, X_train, X_test, z, z_train, z_test, scaling, x, y, z = read_in_dataset(N, scaling, noise)
+X, X_train, X_test, z, z_train, z_test, centering, x, y, z = read_in_dataset(
+    N, centering, noise
+)
 
 # test different values of lambda
 for i in range(len(lambdas)):
@@ -68,9 +70,9 @@ for i in range(len(lambdas)):
         errors_Lasso[n] = np.mean(-scores_Lasso["test_score"])
 
     # plot
-    plt.plot(errors_OLS, 'r--', label="OLS")
-    plt.plot(errors_Ridge, 'b--', label="Ridge")
-    plt.plot(errors_Lasso, 'g--', label="Lasso")
+    plt.plot(errors_OLS, "r--", label="OLS")
+    plt.plot(errors_Ridge, "b--", label="Ridge")
+    plt.plot(errors_Lasso, "g--", label="Lasso")
     plt.ylabel("MSE score")
     plt.xlabel("Polynomial degree (N)")
     plt.title(f"lambda = {lambdas[i]:.5}")
@@ -78,3 +80,4 @@ for i in range(len(lambdas)):
     plt.legend()
 
 plt.show()
+

### 2022-10-10 src/utils.py

@@ -24,7 +24,7 @@ def FrankeFunction(x, y):
 
 # debug function
 def SkrankeFunction(x, y):
-    return x**2 + y**2
+    return 1*x + 2*y + 3*x**2 + 4*x*y + 5*y**2
 
 
 def create_X(x, y, n):
@@ -62,7 +62,7 @@ def OLS(
     return beta
 
 
-def ridge(X_train, z_train, lam, *, scaling=False):
+def ridge(X_train, z_train, lam, *, centering=False):
     L = X_train.shape[1]
 
     beta = np.linalg.pinv(X_train.T @ X_train + lam * np.eye(L)) @ X_train.T @ z_train
@@ -78,7 +78,7 @@ def bootstrap(
     z_test: np.ndarray,
     bootstraps: int,
     *,
-    scaling: bool = False,
+    centering: bool = False,
     model: Callable = OLS,
     lam: float = 0,
 ):
@@ -88,7 +88,7 @@ def bootstrap(
     for i in range(bootstraps):
         X_, z_ = resample(X_train, z_train)
         _, z_pred_train, z_pred_test, _ = evaluate_model(
-            X, X_, X_test, z_, model, lam=lam, scaling=scaling
+            X, X_, X_test, z_, model, lam=lam, centering=centering
         )
         z_preds_train[:, i] = z_pred_train
         z_preds_test[:, i] = z_pred_test
@@ -101,7 +101,7 @@ def crossval(
     z: np.ndarray,
     K: int,
     *,
-    scaling: bool = False,
+    centering: bool = False,
     model=OLS,
     lam: float = 0,
 ):
@@ -137,7 +137,7 @@ def crossval(
             z_train,
             model,
             lam=lam,
-            scaling=scaling,
+            centering=centering,
         )
         errors[k] = MSE(z_test, z_pred_test)
 
@@ -172,11 +172,11 @@ def evaluate_model(
     model,
     *,
     lam: float = 0,
-    scaling: bool = False,
+    centering: bool = False,
 ):
     if isinstance(model, Callable):
         intercept = 0
-        if scaling:
+        if centering:
             X_train = X_train[:, 1:]
             X_test = X_test[:, 1:]
             X = X[:, 1:]
@@ -201,7 +201,7 @@ def evaluate_model(
                     z_train,
                     lam,
                 )
-        # intercept is zero if no scaling
+        # intercept is zero if no centering
         z_pred_train = X_train @ beta + intercept
[…]

### 2022-10-10 src/task_c.py
Update all files to take from commandline

@@ -7,14 +7,22 @@ from utils import *
 np.random.seed(42069)
 
 # parameters
-N = 20
 bootstraps = 100
-# parameters for synthetic data
-noise = 0.05
-centering = False
 
-# get data
-X, X_train, X_test, z, z_train, z_test, centering, x, y, z = read_in_dataset(N, centering, noise)
+(
+    betas_to_plot,
+    N,
+    X,
+    X_train,
+    X_test,
+    z,
+    z_train,
+    z_test,
+    centering,
+    x,
+    y,
+    z,
+) = read_from_cmdline()
 
 # result arrays
 errors = np.zeros(N)
@@ -48,7 +56,7 @@ for n in range(N):
     variances[n] = variance
 
 # plot
-plt.plot(errors, 'g--', label="MSE test")
+plt.plot(errors, "g--", label="MSE test")
 plt.plot(biases, label="biases")
 plt.plot(variances, label="variances")
 plt.xlabel("Polynomial degree (N)")

### 2022-10-10 src/task_d.py

@@ -9,17 +9,24 @@ from utils import *
 np.random.seed(42069)
 
 # parameters
-N = 10
 K = 10
 bootstraps = 100
-# parameters for synthetic data
-noise = 0.05
-centering = False
 
-# read in data
-X, X_train, X_test, z, z_train, z_test, centering, x, y, z = read_in_dataset(
-    N, centering, noise
-)
+(
+    betas_to_plot,
+    N,
+    X,
+    X_train,
+    X_test,
+    z,
+    z_train,
+    z_test,
+    centering,
+    x,
+    y,
+    z,
+) = read_from_cmdline()
+
 z = z.ravel()
 
 # model under testing

### 2022-10-10 src/task_e_betas.py

@@ -7,21 +7,25 @@ from utils import *
 np.random.seed(42069)
 
 # Parameters
-N = 10
 lambdas = np.logspace(-10, 0, 4)
 betas_to_plot = 9
 
 # Parameters for synthetic data
-noise = 0.05
-centering = True
 
-# get data
-X, X_train, X_test, z, z_train, z_test, centering, x, y, z = read_in_dataset(
+(
+    betas_to_plot,
     N,
-    centering=centering,
-    noise=noise,
-    step=0.1,
-)
+    X,
+    X_train,
+    X_test,
+    z,
+    z_train,
+    z_test,
+    centering,
+    x,
+    y,
+    z,
+) = read_from_cmdline()
 
 # calculate beta values
 for i in range(len(lambdas)):

### 2022-10-10 src/task_e_biasvariance.py

@@ -10,21 +10,26 @@ np.random.seed(42069)
 
 # parameters
 K = 20
-N = 10
 bootstraps = 100
 plot_only_best_lambda = False
 lambdas = np.logspace(-10, 0, 4)
 # parameter synthetic data
-noise = 0.05
-centering = True
 
-# read in data
-X, X_train, X_test, z, z_train, z_test, centering, x, y, z = read_in_dataset(
+(
+    betas_to_plot,
     N,
-    centering=centering,
-    noise=noise,
-    step=0.1,
-)
+    X,
+    X_train,
+    X_test,
+    z,
+    z_train,
+    z_test,
+    centering,
+    x,
+    y,
+    z,
+) = read_from_cmdline()
+
 z = z.ravel()
 
 # plot only the gridsearched lambda
@@ -82,4 +87,3 @@ for i in range(len(lambdas)):
     plt.legend()
 
 plt.show()
-

### 2022-10-10 src/task_e_msecomparison.py

@@ -10,11 +10,8 @@ from utils import *
 np.random.seed(42069)
 
 # Parameters
-N = 20
 K = 20
 bootstraps = 100
-noise = 0.05
-centering = True
 kfolds = KFold(n_splits=K)
 
 # if true, plot only lambda that gives lowest MSE
@@ -22,10 +19,20 @@ kfolds = KFold(n_splits=K)
 plot_only_best_lambda = True
 lambdas = np.logspace(-12, -4, 6)
 
-# read in data
-X, X_train, X_test, z, z_train, z_test, centering, x, y, z = read_in_dataset(
-    N, centering, noise
-)
+(
+    betas_to_plot,
+    N,
+    X,
+    X_train,
+    X_test,
+    z,
+    z_train,
+    z_test,
+    centering,
+    x,
+    y,
+    z,
+) = read_from_cmdline()
 z = z.ravel()
 
 # plot only the gridsearched lambda

### 2022-10-10 src/task_f_biasvariance.py

@@ -3,6 +3,7 @@ task g: performs bias-variance tradeoff for lasso regression using the scikit le
         lasso model
 """
 from sklearn.linear_model import Lasso
+
 # Our own library of functions
 from utils import *
 
@@ -10,16 +11,23 @@ np.random.seed(42069)
 
 # parameters
 K = 20
-N = 10
 bootstraps = 100
 plot_only_best_lambda = False
 lambdas = np.logspace(-10, 0, 4)
-# synthetic parameters
-noise = 0.05
-centering = True
-
-# read in and get data
-X, X_train, X_test, z, z_train, z_test, centering, x, y, z = read_in_dataset(N, centering=centering, noise=noise)
+(
+    betas_to_plot,
+    N,
+    X,
+    X_train,
+    X_test,
+    z,
+    z_train,
+    z_test,
+    centering,
+    x,
+    y,
+    z,
+) = read_from_cmdline()
 z = z.ravel()
 
 # plot only the gridsearched lambda
@@ -31,8 +39,8 @@ if plot_only_best_lambda:
 # loop through different lambda values
 for i in range(len(lambdas)):
     if not plot_only_best_lambda:
-           plt.subplot(411 + i)
-           plt.suptitle(f"Bias variance tradeoff for lasso regression")
+        plt.subplot(411 + i)
+        plt.suptitle(f"Bias variance tradeoff for lasso regression")
 
     # model under testing
     model_Lasso = Lasso(lambdas[i], max_iter=1000, fit_intercept=False)
@@ -71,8 +79,12 @@ for i in range(len(lambdas)):
     plt.xlabel("Polynomial degree (N)")
     plt.tight_layout(h_pad=0.001)
     if plot_only_best_lambda:
-        print(f"Optimal lambda = {lam}, best MSE = {best_MSE}, best polynomial = {best_poly}")
-        plt.title(f"Bias variance tradeoff for lasso regression \n for optimal lambda = {lambdas[i]}")
+        print(
+            f"Optimal lambda = {lam}, best MSE = {best_MSE}, best polynomial = {best_poly}"
+        )
+        plt.title(
+            f"Bias variance tradeoff for lasso regression \n for optimal lambda = {lambdas[i]}"
+        )
     else:
         plt.title(f"lambda = {lambdas[i]:.5}")
     plt.legend()

### 2022-10-10 src/task_f_msecomparison.py

@@ -8,22 +8,29 @@ from sklearn.model_selection import cross_validate, KFold
 from utils import *
 
 # parameters
-N = 20
 K = 10
-noise = 0.05
-centering = False
 lambdas = np.logspace(-12, -4, 6)
 
 kfolds = KFold(n_splits=K)
 
+(
+    betas_to_plot,
+    N,
+    X,
+    X_train,
+    X_test,
+    z,
+    z_train,
+    z_test,
+    centering,
+    x,
+    y,
+    z,
+) = read_from_cmdline()
+
 errors_OLS = np.zeros(N)
 errors_Ridge = np.zeros(N)
 errors_Lasso = np.zeros(N)
-
-# read in data
-X, X_train, X_test, z, z_train, z_test, centering, x, y, z = read_in_dataset(
-    N, centering, noise
-)
 
 # test different values of lambda
 for i in range(len(lambdas)):
@@ -80,4 +87,3 @@ for i in range(len(lambdas)):
     plt.legend()
 
 plt.show()
-

### 2022-10-10 src/utils.py
Small fix

@@ -367,7 +367,7 @@ def read_from_cmdline():
         raise ValueError(f"Noise value out of range [0,1]: {args.noise}")
 
     if args.step < 0.01 or args.step > 0.4:
-        raise ValueError(f"Noise value out of range [0,1]: {args.noise}")
+        raise ValueError(f"Step value out of range [0,1]: {args.noise}")
 
     if args.n <= 0:
         raise ValueError(f"Polynomial degree must be positive: {args.N}")

### 2022-10-10 doc/paper.tex
Update paper

@@
-%%% use twocolumn and 10pt options with the asme2ej format
 \documentclass[twocolumn,10pt,cleanfoot]{asme2ej}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{abstract}
-{\it We use Ordinary Least Squares, Ridge and Lasso regression on a synthetic dataset generated by the Franke Function, and on a real dataset of elevation data from a terrain in Norway.
-We explore different regularization values and polynomial degrees and find that that all methods perform similarly on both sets of data.
-We find that linear regression performs well on the synthetic data, but falls short when approximating elevation data with large variances in height.
-}
-\end{abstract}
+\section{Abstract}
+We used linear regression to create a fit of a synthetic dataset generated by the Franke Function, and for real datasets of elevation data from terrain in Norway.
+More specifically, we used Ordinary Least Squares (OLS), Ridge and Lasso regression to explore how the different methods perform on different datasets.
+Comparing the Mean Squared Error (MSE) for the different methods, we found that Ridge and Lasso outperform OLS for small datasets and high polynomial degrees, but that OLS, given the optimal polynomial degree for the dataset, yields the lowest MSE.
+Overall we found that the different forms of linear regression fit the synthetic data well.
+However, when looking at the real datasets, we found that all forms of linear regression performed relatively poorly.
+We concluded that linear regression perhaps is not the best way of predicting terrain data with rapid and irregular changes in elevation.
 
@@
 
-Linear regression is a simple method in machine learning, used to create a fit of a dataset of which it can produce continuous predictions.
+% One of the core challenges of machine learning is model selection, because we cannot make a priori assessments about how well the model will fit our dataset.
+The model is at the heart of machine learning, its like the CPU of the "machine" in machine learning.
+It really puts the "machine" in machine learning, so to speak.
+Learning is like the RAM.
+We are the keyboard and you are the screen.
+Machine learning is well suited for finding complex relationships that are impossible to describe as a series of logical statements, or program by hand.
+Machine learning methods can look at millions of datapoints and parameters at once, and thus find links in the data which could never be found by manual inspection.
+But to do this, we require a model that is suited to our dataset.
+However, in order to create an optimal model, we require optimal parameters which cannot be known a priori.
+Thus, we require extensive testing and tuning of parameters to inform our selection.
+In this paper, we explore
+
+% One of the core challenges of machine learning is model selection, because we cannot make a priori assessments about how well the model will fit our dataset.
+If our model is not complex enough for our problem, we are unlikely to reach a good prediction.
+Likewise, if our model is too complex, we run the risk of overfitting, where our model adapts to our training data perfectly and thus loses the ability to make predictions on unseen data.
+Given the enormous datasets that
+
+% Linear regression is a simple method in machine learning, used to create a fit of a dataset of which it can produce continuous predictions.
 This is achieved by assuming that we can model the output values as a linear combination of continuous functions of the input values.
@@
 
-However, in order to create an optimal model, we require optimal parameters which cannot be known a priori.
-Thus, we require extensive testing and tuning of parameters to inform our selection.
-This presents a major obstacle, as we must for example find the optimal polynomial degree $N$ for which our model best approximates the observed values at the same time as we tune the hyperparameter $\lambda$ responsible for an adequate regularisation.
-We must find these parameters which lie at an intersection, where all parameters yield the minimal error in unity.
-
-Furthermore, we will compare different models of linear regression, namely Ordinary Least Squares (OLS), Ridge and Lasso regression, and employ resampling methods, namely bootstrap and cross-validation to ensure that we draw conclusive results as these help us reduce the randomness in our parameters based off how we split our data.
+% Furthermore, we will compare different models of linear regression, namely Ordinary Least Squares (OLS), Ridge and Lasso regression, and employ resampling methods, namely bootstrap and cross-validation to ensure that we draw conclusive results as these help us reduce the randomness in our parameters based off how we split our data.
 <source?>
@@
 
-This sections covers the theoretic background which underlies the methods used in this paper.
+This sections covers the theoretic background which underlies the methods used in this paper. We use the following shorthands for our mathematical notation.
 
@@
 
-The correctness of our linear regression can be confirmed by running our debug function (the so called SkrankeFunction)\footnote{this can be done by running \texttt{\$ python3 task\_b.py --debug -n 2}}, which is a second degree polynomial where $b_0 = 0, \; b_1 = 1, \; b_2 = 2$ and so on.
+The correctness of our linear regression can be confirmed by running our debug function (the so called SkrankeFunction)\footnote{this can be done by running \texttt{\$ python3 task\_b.py --debug -n 2 --noscale}}, which is a second degree polynomial where $b_0 = 0, \; b_1 = 1, \; b_2 = 2$ and so on.
+The SkrankeFunction (taken from \texttt{src/utils.py}) looks like this
+
+\lstinputlisting[language=Python, firstline=238, lastline=245]{../src/utils.py}
 
@@
 With the regularization parameter, we can prevent overfitting if our model is too complex for our data.
+Looking at Fig.~\ref{frankeridgenoisecomparison} in comparison to Fig.~\ref{frankienoisecomparison} (both in the appendix), we can see that Ridge really has achieved this, reducing overfitting as our data becomes more and more noisy.
 But we should recall that we decide the complexity of our model, by deciding which polynomial degree we use when creating the design matrix.
@@
 
-\subsection{Elevation data}
+\subsection{Real elevation data}
 
@@
 Here we find some limitations of linear regression.
-We see from Fig.
-NUMBER that the we are able to fit the terrain up to a certain point, gaining an MSE of NUMBER with OLS with polynomial degree DEGREE.
+We see from Fig.~\ref{realmse} that the we are able to fit the terrain up to a certain point, gaining an MSE of NUMBER with OLS with polynomial degree DEGREE.
 However, we see that beyond this, we have little to gain.
@@
 Clearly, linear regression does not offer a very good tradeoff between computational cost and performance for this data.
+
+% \begin{figure} 
+% \centerline{\includegraphics[width=3.25in]{figure/realmse.png}}
+% \caption{Test and train MSE }
+% \label{frankiedifflambdas}
+% \end{figure}
 
@@
 \centerline{\includegraphics[width=3.25in]{figure/frankienoisecomparison.png}}
-\caption{Test and train MSE for the Franke function with an increasing amount of noise}
+\caption{Test and train MSE for the OLS prediction of Franke function with an increasing amount of noise}
 \label{frankienoisecomparison}
+\end{figure*}
+
+\begin{figure*}
+% \begin{figure}[t]
+\centerline{\includegraphics[width=3.25in]{figure/frankeridgenoisecomparison.png}}
+\caption{Test and train MSE for the Ridge prediction with $\lambda = 10^{-8}$ of the Franke function with an increasing amount of noise}
+\label{frankeridgenoisecomparison}
 \end{figure*}

### 2022-10-10 doc/paper.tex
More paper

@@
 
-\lstinputlisting[language=Python, firstline=53, lastline=64]{../src/utils.py}
+\lstinputlisting[language=Python, firstline=56, lastline=64]{../src/utils.py}
 
@@
 
-The correctness of our linear regression can be confirmed by running our debug function (the so called SkrankeFunction)\footnote{this can be done by running \texttt{\$ python3 task\_b.py --debug -n 2 --noscale}}, which is a second degree polynomial where $b_0 = 0, \; b_1 = 1, \; b_2 = 2$ and so on.
+The correctness of our Ordinary Least Squares regression can be confirmed by running our debug function (the so called SkrankeFunction)\footnote{this can be done by running \texttt{\$ python3 task\_b.py --debug --noscale -n 2}}, which is a second degree polynomial where $b_0 = 0, \; b_1 = 1, \; b_2 = 2$ and so on.
 The SkrankeFunction (taken from \texttt{src/utils.py}) looks like this
 
-\lstinputlisting[language=Python, firstline=238, lastline=245]{../src/utils.py}
+\lstinputlisting[language=Python, firstline=26, lastline=27]{../src/utils.py}
+
+and the result of our regression is correctly given as $\bm{\beta} = (0 \; 1 \; 2 \; 3 \; 4 \; 5\;)^T$.
 

### 2022-10-10 doc/paper.tex
fix introduction

@@
 
-% One of the core challenges of machine learning is model selection, because we cannot make a priori assessments about how well the model will fit our dataset.
-The model is at the heart of machine learning, its like the CPU of the "machine" in machine learning.
-It really puts the "machine" in machine learning, so to speak.
-Learning is like the RAM.
-We are the keyboard and you are the screen.
-Machine learning is well suited for finding complex relationships that are impossible to describe as a series of logical statements, or program by hand.
-Machine learning methods can look at millions of datapoints and parameters at once, and thus find links in the data which could never be found by manual inspection.
-But to do this, we require a model that is suited to our dataset.
-However, in order to create an optimal model, we require optimal parameters which cannot be known a priori.
-Thus, we require extensive testing and tuning of parameters to inform our selection.
-In this paper, we explore
+One of the core challenges in machine learning is model selection, because we cannot make a priori assessments about how well the model will fit our dataset.
+A multitude of factors determine the efficacy of our model, the machine learning method we use, the hyperparameters of said method, and the model complexity we select.
+An uninformed decision in model selection could result in an under- or overfitting to the data, yielding poor results.
+Thus, it is important that we rigorously test our tuning such that these factors coincide at an intersection, providing the minimal error in unity.
 
-% One of the core challenges of machine learning is model selection, because we cannot make a priori assessments about how well the model will fit our dataset.
-If our model is not complex enough for our problem, we are unlikely to reach a good prediction.
-Likewise, if our model is too complex, we run the risk of overfitting, where our model adapts to our training data perfectly and thus loses the ability to make predictions on unseen data.
-Given the enormous datasets that
+In this paper, we will provide a detailed inspection of the Ordinary Least Squares (OLS), Ridge and Lasso linear regression methods, it's hyperparameters and varying model complexity with the aim of making an informed decision about model selection when fitting to different datasets.
+The datasets consist of synthetic data created by the Franke function, and real datasets of terrain in Norway.
+We inform our decisions by plotting the mean squared error (MSE), our cost function, over different polynomial degrees describing model complexity, over different resampling techniques and over different values of lambda, which is the hyperparameter responsible for a suitable regularization.
+We will perform a bias-variance decomposition for the different linear regression methods to check for under- and overfitting and plot our implementations of the methods up against popular implementations such as scikit learn to ensure that the behaviour of our models is expected.
 
-% Linear regression is a simple method in machine learning, used to create a fit of a dataset of which it can produce continuous predictions.
-This is achieved by assuming that we can model the output values as a linear combination of continuous functions of the input values.
-For example, we can assume that the observed values in our dataset can be approximated by different polynomial degrees of the input values and find the linear coefficients $\beta$ which minimize the difference between our observed and predicted values.
-This general idea that we can create a model which approximates our observed data with a minimal difference between observed and predicted values is a problem at the very heart of machine learning.
-
-% Furthermore, we will compare different models of linear regression, namely Ordinary Least Squares (OLS), Ridge and Lasso regression, and employ resampling methods, namely bootstrap and cross-validation to ensure that we draw conclusive results as these help us reduce the randomness in our parameters based off how we split our data.
-<source?>
+First, we will go more into depth about the methods and the theory behind them, along with a brief description of our implementation.
+We will further continue to the results, followed up by a critical discussion of the results.
+Finally, we will give a conclusion on the optimal model selection for the different datasets.
 

### 2022-10-10 src/task_d.py

@@ -91,9 +91,9 @@ for n in range(N):
 plt.plot(errors_boot, label="bootstrap")
 plt.plot(errors_cv, label="cross validation implementation")
 plt.plot(errors_cv_scikit, label="cross validation scikit learn")
-plt.ylabel("MSE score")
-plt.xlabel("Polynomial degree (N)")
-plt.title(f"MSE by Resampling Method")
-plt.legend()
+plt.ylabel("MSE score", size=15)
+plt.xlabel("Polynomial degree (N)", size=15)
+plt.title(f"MSE by Resampling Method", size=18)
+plt.legend(prop={'size': 10})
 
 plt.show()

### 2022-10-10 src/noise_plot.py
Make bootstrap mse test in task c and update files

@@ -24,6 +24,7 @@ noises = np.linspace(0, 0.2, 5)
 centering = False
 bootstraps = 100
 
+
 def noise_plot(x, y, z, N, noise, centering, bootstraps):
     # add noise
     z += noise * np.random.standard_normal(z.shape)
@@ -36,7 +37,7 @@ def noise_plot(x, y, z, N, noise, centering, bootstraps):
     for n in range(N):
         print(n)
         l = int((n + 1) * (n + 2) / 2)  # Number of elements in beta
-        z_preds_test = bootstrap(
+        z_preds_test, _ = bootstrap(
             X[:, :l],
             X_train[:, :l],
             X_test[:, :l],
@@ -44,6 +45,8 @@ def noise_plot(x, y, z, N, noise, centering, bootstraps):
             z_test,
             bootstraps,
             centering=centering,
+            model=ridge,
+            lam=10 ** (-8),
         )
 
         MSE_test, _ = scores(z_test, z_preds_test)

### 2022-10-10 src/task_d.py

@@ -70,7 +70,7 @@ for n in range(N):
     print(n)
     l = int((n + 1) * (n + 2) / 2)  # Number of elements in beta
 
-    z_preds = bootstrap(
+    z_preds, _ = bootstrap(
         X[:, :l],
         X_train[:, :l],
         X_test[:, :l],
@@ -94,6 +94,6 @@ plt.plot(errors_cv_scikit, label="cross validation scikit learn")
 plt.ylabel("MSE score", size=15)
 plt.xlabel("Polynomial degree (N)", size=15)
 plt.title(f"MSE by Resampling Method", size=18)
-plt.legend(prop={'size': 10})
+plt.legend(prop={"size": 10})
 
 plt.show()

### 2022-10-10 src/task_e_biasvariance.py

@@ -53,7 +53,7 @@ for i in range(len(lambdas)):
     for n in range(N):
         print(n)
         l = int((n + 1) * (n + 2) / 2)  # Number of elements in beta
-        z_preds = bootstrap(
+        z_preds, _ = bootstrap(
             X[:, :l],
             X_train[:, :l],
             X_test[:, :l],

### 2022-10-10 src/task_e_manualsearch.py

@@ -36,7 +36,7 @@ xi = [i for i in range(0, num_lambdas, 2)]
 for i in range(num_lambdas):
     print(i)
 
-    z_preds = bootstrap(
+    z_preds, _ = bootstrap(
         X,
         X_train,
         X_test,

### 2022-10-10 doc/paper.tex
More paper

@@
 Our assumption is that there is some relationship $y_i = f(x_i) + \epsilon_i$ where $\epsilon_i \sim N(0,\sigma^2)$, and we wish to create a model which approximates this relationship.
-More specifically, we wish to find a vector $\beta$ which gives us the "best" approximation $\bm{y_{pred}}$ of $\bm{y}$, by using the parameters in the following equation:
+More specifically, we wish to find a vector $\beta$ which gives us the "best" approximation $\bm{\tilde{y}}$ of $\bm{y}$, by using the parameters in the following equation:
 
@@
 X \bm{\beta} = \bm{\tilde{y}}
+\label{first}
 \end{equation}
 
-$X$ is called the design matrix, and is an $n$ by $p$ matrix created by our input data and our choice of basis functions, where $n$ is the amount of different input states we have (for our problem, $n$ is the number of distinct $x$ values times the number of distinct $y$ values) and $p$ is the number of features we use for our fit (in our case, $p$ is based on the polynomial degree we use).
-The element $X_{ij}$ is calculated by evaluating the basis function corresponding the coloumn $j$ for the input data corresponding to the row $j$.
-As we can see, each row of this matrix is scalar multiplied by the vector $\beta$, giving us a linear combination of the values of $X_{i*}$ for each $y_i$ in the target vector $\bm{y}$.
-This vector is called the parameter vector, and finding the optimal values $\hat{\bm{\beta}}$ is the problem we wish to solve.
+$X$ is called the design matrix, and is an $n$ by $p$ matrix created by our input data and our choice of basis functions, where $n$ is the amount of different input states we have and $p$ is the number of features we use for our fit.
+For our problem, $n$ is the number of distinct $x$ values times the number of distinct $y$ values, and $p$ is based on the polynomial degree we use.
+The basis functions are the summands of a two variable polynomial of degree $N$.
+The element $x_{ij}$ is calculated by evaluating the basis function corresponding the coloumn $j$ for the input data corresponding to the row $i$.
+As we can see from Eq.~\ref{first}, each row of this matrix is scalar multiplied by the vector $\beta$, giving us a linear combination of the values of $X_{i*}$ for each $y_i$ in the vector $\bm{\tilde{y}}$.
+The vector $\bm{\beta}$ is called the parameter vector, and finding the optimal values $\hat{\bm{\beta}}$ is the problem we wish to solve.
 But for this, we need to define what an "optimal" prediction $\bm{\tilde{y}}$ is.
@@
 
-The define the optimal prediction, we use a cost function, and the choice of cost function greatly influences what optimal parameters we get.
-\footnote{It is here important to keep in mind that the cost function used when fitting the model (training) rarely is the same as the one used for testing our prediction, as described in <somewhere>}.
-An intuitive cost function is the mean squared error (MSE).
+To define the optimal prediction, we use a cost function, and the choice of cost function greatly influences what optimal parameters we get.\footnote{It is here important to keep in mind that the cost function used when fitting the model (training) rarely is the same as the one used for testing our prediction, as described in <somewhere>} An intuitive cost function is the mean squared error (MSE).
 This function simply computes the average distance between our prediction and the target value, squared:
@@
 We see that we have a function of our parameters $\bm{\beta}$, which describe how much our prediction deviates from our target values.
-Our problem now is to optimize this function, that is, find the values of $\beta$ which give MSE the lowest value.
+Our problem now is to optimize this function, that is, find the values of $\beta$ which gives the MSE the lowest value.
 Luckily for us, it can be proven that the second derivative of this function is always positive (WHERE?), which means that we can find the global minimum by solving the equation
@@
 
-The process we have just described, with MSE as the cost function, is the what is known as Ordinary Least Squares (OLS) regression.
+The process we have just described, with MSE as the cost function, is what is known as Ordinary Least Squares (OLS) regression.
 Using different cost functions when fitting the model leads to different equations for $\bm{\hat{\beta}}$.
@@
 ~\ref{ols_mse_table} shows the best polynomial degree and MSE for regular and noisy data.
-Here we see that the best degree without noise is 10, which shows that even without noise, the low amount of data leads to a slight overfit even at very small polynomial degrees.
+Here we see that the best degree without noise is 9, which shows that even without noise, the low amount of data leads to a slight overfit even at very small polynomial degrees.
 This is not surprising, as at $N = 10$ we have $66$ features, which is a substantial amount compared to our measly 320 training datapoints.
@@
 \hline
-0 & 0.0004335 & 10\\
+0 & 0.00051085 & 9\\
 0.05 & 0.0038302 & 7 \\
@@
 % \hline
-Figure & Highest $N$ & Std of noise & Number of datapoints & Bootstraps & K-folds & $\lambda$ & Scaling \\
+Figure & Highest $N$ & Std of noise & Step (\# of points) & Bootstraps & K-folds & $\lambda$ & Scaling \\
 \hline
-\ref{frankenonoise} & 10 & 0 & 400 & - & - & - & None\\
-\ref{frankenonoisemse} & 25 & 0 & 400 & 100 & - & - & None\\
-\ref{frankenoisemse} & 25 & 0.05 & 400 & 100 & - & - & None\\
-\ref{frankiebiasvariance} & 25 & 0.05 & 100 & 100 & - & - & None\\
-\ref{frankiedifflambdas} & 25 & 0.05 & 100 & 100 & - & variable & Centering\\
-\ref{frankeridgebetas} & 25 & 0.05 & 100 & - & - & variable & Centering\\
-\ref{frankenoisybeta} & 25 & 0.05 & 400 & - & - & - & Centering\\
-\ref{frankienoisecomparison} & 25 & variable & 400 & 100 & - & - & None\\
-\ref{frankiehigherdatapoints} & 30 & 0 & 10000 & 100 & - & - & None\\
+\ref{frankiebiasvariance} & 25 & 0.05 & 0.1 (100) & 100 & - & - & None\\
+\ref{frankiedifflambdas} & 25 & 0.05 & 0.1 (100) & 100 & - & variable & Centering\\
+\ref{frankeridgebetas} & 25 & 0.05 & 0.1 (100) & - & - & variable & Centering\\
+\ref{frankenoisybeta} & 25 & 0.05 & 0.05 (400) & - & - & - & Centering\\
+\ref{frankienoisecomparison} & 25 & variable & 0.05 (400) & 100 & - & - & None\\
+\ref{frankiehigherdatapoints} & 30 & 0 & 0.01 (10000) & 100 & - & - & None\\
+\hline
+\end{tabular}
+\end{center}
+\end{table*}
+
+\begin{table*}
+\caption{This table shows the command to execute, as well as the advanced parameters to change, to reproduce every figure in the report.
+(More info about the scripts and their parameters can be found in the README)}
+\begin{center}
+\label{allparamstable}
+\begin{tabular}{c | l l l l}
+% & & \\ % put some space after the caption
+% \hline
+Figure & Shell command (leading \texttt{python3} omitted) & Bootstraps & K-folds & $\lambda$ \\
+\hline
+\ref{frankenonoise} & \texttt{task\_b.py --noise 0 -n 25} & 100 & - & - \\
+\ref{frankenonoisemse} & \texttt{task\_c.py --noise 0 -n 25}  & 100 & - & - \\
+\ref{frankenoisemse} & \texttt{task\_c.py -n 25}  & 100 & - & - \\
+\ref{frankienoisecomparison} & \texttt{noise\_plot.py} & 100 & - & - \\
+\ref{frankiehigherdatapoints} & \texttt{task\_c.py --noise 0 -n 30 --step 0.01} & 100 & - & - \\
 \hline

### 2022-10-10 src/task_c.py
Add minimal error

@@ -79,12 +79,10 @@ for n in range(N):
     MSE_test_sk = np.mean(mse_scores_sk)
     MSE_train_sk = MSE(z_pred_train_sk, z_train)
 
-    print(MSE_train_sk)
     MSEs_test_sk[n] = MSE_test_sk
     MSEs_train_sk[n] = MSE_train_sk
 
-    # bias-variance trade-off
-
+print(f"Minimal MSE_test value = {np.min(MSEs_test)} for N = {np.argmin(MSEs_test)}")
 # plot mse test and train
 plt.plot(MSEs_train, label="train implementation", marker="o", markersize=3)
 plt.plot(MSEs_test, label="test implementation", marker="o", markersize=3)

### 2022-10-11 doc/paper.tex
More paper

@@
 The MSE for test in these plots is the average MSE for 100 bootstraps.
-For a full list of all parameters used for every plot shown, see Table.~\ref{allparamstable} in the appendix.
+For a comprehensive list of the shell command required to reproduce this and every other plot, see Table.~\ref{allparamstable} in the appendix.
 
@@
 \hline
-0 & 0.00051085 & 9\\
+0 & 0.0005109 & 9\\
 0.05 & 0.0038302 & 7 \\
@@
 But we can see from the Table.
-~\ref{ols_vs_ridge_table} that we are not able to push into higher degrees.
+~\ref{ols_vs_ridge_vs_lasso_table} that we are not able to push into higher degrees.
 A possible explanation for this is that every feature is useful, due to the polynomial expansion of the Franke function.
 
-\begin{table}[t]
-\caption{Best MSE for Ridge regression and OLS with 400 datapoints}
-\begin{center}
-\label{ols_vs_ridge_table}
-\begin{tabular}{c | l l l l}
-% & & \\ % put some space after the caption
+% \begin{table}[t]
+% \caption{Best MSE for Ridge regression and OLS with 400 datapoints}
+% \begin{center}
+% \label{ols_vs_ridge_table}
+% \begin{tabular}{c | l l l l}
+% % & & \\ % put some space after the caption
+% % \hline
+% Regression & Std of noise & MSE & Degree & $\lambda$ \\
 % \hline
-Regression & Std of noise & MSE & Degree & $\lambda$ \\
-\hline
-OLS & 0.05 & 0.0038302 & 7 & - \\
-Ridge & 0.05 & 0.0037999 & 7 & $10^{-8}$ \\
-\hline
-\end{tabular}
-\end{center}
-\end{table}
+% OLS & 0.05 & 0.0038302 & 7 & - \\
+% Ridge & 0.05 & 0.0037999 & 7 & $10^{-8}$ \\
+% \hline
+% \end{tabular}
+% \end{center}
+% \end{table}
 

### 2022-10-11 src/task_c.py
Fix fit_intercept

@@ -39,7 +39,7 @@ MSEs_train_sk = np.zeros(N)
 
 # model under testing
 OLS_model = OLS
-OLS_scikit = LinearRegression()
+OLS_scikit = LinearRegression(fit_intercept=False)
 
 # for polynomial degree
 for n in range(N):

### 2022-10-11 src/utils.py
Remove prints

@@ -408,8 +408,6 @@ def read_from_cmdline():
             # add noise
             z += args.noise * np.random.standard_normal(z.shape)
         centering = not args.noscale
-        print(f"{args.noscale=}")
-        print(f"{centering=}")
 
         X, X_train, X_test, z_train, z_test = preprocess(x, y, z, args.n, 0.2)
 

### 2022-10-11 doc/paper.tex
More paper

@@
 
-\begin{table}[t]
+\begin{table*}[t]
 \caption{Best MSE for OLS, Ridge and Lasso with 400 datapoints}
@@
 \end{center}
-\end{table}
+\end{table*}
 
@@
 
+\begin{figure*} 
+\centerline{\includegraphics[width=6.85in]{figure/real1surfacecomp.png}}
+\caption{A comparison of our best OLS prediction and actual terrain data (terrain 1), at polynomial degree 19}
+\label{real1surfacecomp}
+\end{figure*}
+
 \subsection{Real elevation data}
@@
 Here we find some limitations of linear regression.
-We see from Fig.~\ref{realmse} that the we are able to fit the terrain up to a certain point, gaining an MSE of NUMBER with OLS with polynomial degree DEGREE.
-However, we see that beyond this, we have little to gain.
-This can be seen when we increase the polynomial degree to very large amounts, without seeing any decrease in MSE (Fig.
+We see from Fig.~\ref{real1msetraintest} that the we are able to fit the terrain up to a certain point, gaining an MSE of $0.0142683$ with OLS with polynomial degree 19.
+This is about two orders of magnitude higher than our lowest MSE for the Franke function without noise, which had values on a similar scale.\footnote{The Franke function without noise has a range of about $(0,1.2)$ and our MinMaxed terrain data has been scaled to a range of $(0,1)$.
+While not exactly the same, these are on the same order of magnitude, and we believe that comparing the MSE scores is reasonable} In fact, our MSE is comparable to the Franke function with a stochastic noise $\epsilon \sim N(0,0.1)$, which implies that our terrain is very irregular and difficult to approximate.
+Looking at the plot for polynomial degree 19 (Fig.~\ref{real1surfacecomp}), we see that our poor MSE is reflected visually in our fit as well, as the surfaces do not look particularly similar.
+Comparing the best polynomial degree of the best fit with the synthetic data, we see that our best fit is reached much later.
+This implies that we are not overfitting yet, and indeed the bias variance decomposition confirms this, as we see that variance is essentially inconsequential (Fig.~\ref{realbiasvarianceOLS}).
+%This can be seen when we increase the polynomial degree to very large amounts, without seeing any decrease in MSE (Fig.
 NUMBER).
@@
 NUMBER), we see that the parameters stabilize, which implies that we are not overfitting, but simply reaching our best possible prediction for polynomial regression.
-The fact that we are not overfitting is further reinforced by looking at the bias-variance plot in Fig.
-NUMBER, which shows no increase in variance.
+The fact that we are not overfitting is further reinforced by looking at the bias-variance plot in Fig.~\ref{realbiasvarianceOLS}, which shows no increase in variance.
 It is reasonable to expect that our model will not overfit, due to the large increase in data we now use for the fit.
@@
 This means that for a 50th order polynomial, which will have $1326$ features, our number of features are less than $1\%$ of our datapoints.
-If we decrease the number of datapoints used for training considerably, we can see overfitting start to occur (see Appendix 1).
+If we decrease the number of datapoints used for training considerably, we can see overfitting start to occur (Fig.~\ref{real1mselessdata} in the appendix).
 So we are not overfitting, but we are not decreasing our MSE either.
@@
 Clearly, linear regression does not offer a very good tradeoff between computational cost and performance for this data.
+
+\begin{figure} 
+\centerline{\includegraphics[width=3.25in]{figure/realbiasvarianceOLS.png}}
+\caption{Bias, variance and test MSE plotted over polynomial degree for our OLS prediction of Terrain 1}
+\label{realbiasvarianceOLS}
+\end{figure}
+
+\begin{figure} 
+\centerline{\includegraphics[width=3.25in]{figure/real1msetraintest.png}}
+\caption{Test and train MSE for OLS on Terrain 1}
+\label{real1msetraintest}
+\end{figure}
 
@@
 \hline
-\ref{frankiebiasvariance} & 25 & 0.05 & 0.1 (100) & 100 & - & - & None\\
 \ref{frankiedifflambdas} & 25 & 0.05 & 0.1 (100) & 100 & - & variable & Centering\\
 \ref{frankeridgebetas} & 25 & 0.05 & 0.1 (100) & - & - & variable & Centering\\
-\ref{frankenoisybeta} & 25 & 0.05 & 0.05 (400) & - & - & - & Centering\\
-\ref{frankienoisecomparison} & 25 & variable & 0.05 (400) & 100 & - & - & None\\
-\ref{frankiehigherdatapoints} & 30 & 0 & 0.01 (10000) & 100 & - & - & None\\
 \hline
@@
 \hline
-\ref{frankenonoise} & \texttt{task\_b.py --noise 0 -n 25} & 100 & - & - \\
+\ref{frankenonoise} & \texttt{task\_b.py --noise 0 -n 25} & - & - & - \\
 \ref{frankenonoisemse} & \texttt{task\_c.py --noise 0 -n 25}  & 100 & - & - \\
 \ref{frankenoisemse} & \texttt{task\_c.py -n 25}  & 100 & - & - \\
+\ref{frankiebiasvariance} & \texttt{task\_c.py -n 25 --step 0.1}  & 100 & - & - \\
+\ref{frankenoisybeta} & \texttt{task\_c.py -n 25}  & 100 & - & - \\
 \ref{frankienoisecomparison} & \texttt{noise\_plot.py} & 100 & - & - \\
 \ref{frankiehigherdatapoints} & \texttt{task\_c.py --noise 0 -n 30 --step 0.01} & 100 & - & - \\
+\ref{real1surfacecomp} & \texttt{task\_b.py --file ../data/small\_SRTM\_data\_Norway\_1.tif -n 19} & - & - & - \\
+\ref{real1mselessdata} & \texttt{task\_c.py -f ../data/tiny\_SRTM\_data\_Norway\_1.tif -n 20} & 100 & - & - \\
 \hline
@@
 \label{frankienoisecomparison}
+\end{figure*}
+
+\begin{figure*}
+% \begin{figure}[t]
+\centerline{\includegraphics[width=3.25in]{figure/real1mselessdata.png}}
+\caption{Test and train MSE for the OLS prediction of terrain 1, this time downscaled to only $50x100$. Here the optimal degree was 11, showing that with lower data we start to overfit more rapidly.}
+\label{real1mselessdata}
 \end{figure*}

### 2022-10-11 doc/paper.tex
More paper

@@
 \usepackage{listings}
+\usepackage{tablefootnote}
 \usepackage{float}
@@
 	Bachelor Student, Robotics and Intelligent Systems\\
+	Department of Informatics\\
 	The faculty of Mathmatics and Natural Sciences\\
-	Department of Informatics\\
     Email: jonatahh@ifi.uio.no
@@
 	Bachelor Student, Robotics and Intelligent Systems\\
+	Department of Informatics\\
 	The faculty of Mathmatics and Natural Sciences\\
-	Department of Informatics\\
     Email: ericer@ifi.uio.no
@@
 
-Using Ridge regression, we see from Fig.~\ref{frankiedifflambdas} that larger and larger values of $\lambda$ efficiently prevents overfitting and reduces the variance between predictions.
-Looking at the beta progression with for the same values of $\lambda$ (Fig.~\ref{frankeridgebetas}), we see that the regularization parameter is successfully shrinking the beta values.
+Using Ridge regression, we see from Fig.~\ref{frankiedifflambdas} in the appendix that larger and larger values of $\lambda$ efficiently prevents overfitting and reduces the variance between predictions.
+Looking at the beta progression with for the same values of $\lambda$ (Fig.~\ref{frankeridgebetas} in the appendix), we see that the regularization parameter is successfully shrinking the beta values.
 Using Scikit's gridsearch, we find the optimal $\lambda$ to be $10^{-8}$ and the optimal polynomial degree to be 7, with an average MSE of $0.0037999$, which is very slightly lower than OLS for the Franke function with an added stochastic noise $\epsilon \sim N(0,0.05)$.
@@
 A possible explanation for this is that every feature is useful, due to the polynomial expansion of the Franke function.
-
-% \begin{table}[t]
-% \caption{Best MSE for Ridge regression and OLS with 400 datapoints}
-% \begin{center}
-% \label{ols_vs_ridge_table}
-% \begin{tabular}{c | l l l l}
-% % & & \\ % put some space after the caption
-% % \hline
-% Regression & Std of noise & MSE & Degree & $\lambda$ \\
-% \hline
-% OLS & 0.05 & 0.0038302 & 7 & - \\
-% Ridge & 0.05 & 0.0037999 & 7 & $10^{-8}$ \\
-% \hline
-% \end{tabular}
-% \end{center}
-% \end{table}
 
@@
 
-\begin{figure} 
-\centerline{\includegraphics[width=3.25in]{figure/frankiedifflambdas.png}}
-\caption{Bias, variance and test MSE plotted over polynomial degree for our Ridge prediction on the Franke function with an added stochastic noise $\epsilon \sim N(0,0.05)$, using lambdas of increasing size.
-Here we have reduced the number of datapoints to only 100 to force overfitting earlier.}
-\label{frankiedifflambdas}
-\end{figure}
-
-\begin{figure} 
-\centerline{\includegraphics[width=3.25in]{figure/frankeridgebetas.png}}
-\caption{Beta progression for different values of lambda in Ridge regression. Note the different scales on each of the plots, showing how increasing the regularization parameter shrinks the betas}
-\label{frankeridgebetas}
-\end{figure}
-
-\begin{figure*} 
-\centerline{\includegraphics[width=6.85in]{figure/real1surfacecomp.png}}
-\caption{A comparison of our best OLS prediction and actual terrain data (terrain 1), at polynomial degree 19}
-\label{real1surfacecomp}
-\end{figure*}
-
 \subsection{Real elevation data}
@@
 This is about two orders of magnitude higher than our lowest MSE for the Franke function without noise, which had values on a similar scale.\footnote{The Franke function without noise has a range of about $(0,1.2)$ and our MinMaxed terrain data has been scaled to a range of $(0,1)$.
-While not exactly the same, these are on the same order of magnitude, and we believe that comparing the MSE scores is reasonable} In fact, our MSE is comparable to the Franke function with a stochastic noise $\epsilon \sim N(0,0.1)$, which implies that our terrain is very irregular and difficult to approximate.
+While not exactly the same, these are on the same order of magnitude, and we believe that comparing the MSE scores is reasonable} In fact, our MSE is comparable to the Franke function with a stochastic noise $\epsilon \sim N(0,0.1)$ (Table.~\ref{ols_mse_table_dif_data}), which implies that our terrain is very irregular and difficult to approximate.
 Looking at the plot for polynomial degree 19 (Fig.~\ref{real1surfacecomp}), we see that our poor MSE is reflected visually in our fit as well, as the surfaces do not look particularly similar.
@@
 Our synthetic data used $20 x 20 = 400$ datapoints, while we now have $320 x 640 = 204 800$ datapoints.
-This means that for a 50th order polynomial, which will have $1326$ features, our number of features are less than $1\%$ of our datapoints.
-If we decrease the number of datapoints used for training considerably, we can see overfitting start to occur (Fig.~\ref{real1mselessdata} in the appendix).
+This means that for a 20th order polynomial, which will have $231$ features, we will have around a thousand datapoints per feature.
+If we decrease the number of datapoints availiable considerably, we can see overfitting start to occur (Fig.~\ref{real1mselessdata} in the appendix).
 So we are not overfitting, but we are not decreasing our MSE either.
@@
 Clearly, linear regression does not offer a very good tradeoff between computational cost and performance for this data.
+
+\begin{table}[t]
+\caption{Best MSE and polynomial degree for OLS on synthetic and real data}
+\begin{center}
+\label{ols_mse_table_dif_data}
+\begin{tabular}{c l l l}
+Dataset & Std of noise & Best MSE & Best degree \\
+\hline
+Franke & 0 & 0.0005109 & 9\\
+Franke & 0.05 & 0.0038302 & 7 \\
+Franke & 0.10 & 0.0139833 & 7 \\
+Terrain 1 & - & 0.0142683 & 19 \\
+\hline
+\end{tabular}
+\end{center}
+\end{table}
 
@@
 \end{figure}
-
-% \begin{figure} 
-% \centerline{\includegraphics[width=3.25in]{figure/realmse.png}}
-% \caption{Test and train MSE }
-% \label{frankiedifflambdas}
-% \end{figure}
+\begin{figure} 
+\centerline{\includegraphics[width=3.25in]{figure/real1msetraintest.png}}
+\caption{Test and train MSE for OLS on Terrain 1}
+\label{real1msetraintest}
+\end{figure}
 
@@
 As our model is not overfitting, we do not expect Ridge to offer much improvement.
+This proves to be the case, as we see from Fig.~\ref{real1ridgedifflambdas}.
+Here we see that for low values of lambda there is no improvement, as there are no parameters that need regularization, and no overfitting to supress.
+For larger values, we see that we are shrinking the parameters too much, and we see an increase in MSE.
+We could expect to see that Ridge shows improvement at higher polynomial degrees, when OLS starts to overfit, but due to computational constraints we have not been able to push much further than 20 degrees.
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%% begin figure %%%%%%%%%%%%%%%%%%%
-% \begin{figure}[t]
-% \begin{center}
-% \setlength{\unitlength}{0.012500in}%
-% \begin{picture}(115,35)(255,545)
-% \thicklines
-% \put(255,545){\framebox(115,35){}}
-% \put(275,560){Beautiful Figure}
-% \end{picture}
-% \end{center}
-% \caption{The caption of a single sentence does not have period at the end}
-% \label{figure_ASME} 
-% \end{figure}
-%%%%%%%%%%%%%%%% end figure %%%%%%%%%%%%%%%%%%% 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsubsection{Lasso Regression}
 
-%%%%%%%%%%%%%%%% begin figure %%%%%%%%%%%%%%%%%%%
-% \begin{figure} 
-% \centerline{\includegraphics[width=3.34in]{figure/FMANU_MD_05_1272_5.jpg}}
-% \caption{While this figures is easily readable at a double column width of 6.5in, when it is shrunk to 3.25in column width the text is unreadable. This paper was held from production.}
-% \label{fig_example2.jpg}
-% \end{figure}
-%%%%%%%%%%%%%%%% end figure %%%%%%%%%%%%%%%%%%%
+With lasso regression, we see similar results as with Ridge.
+Comparing all the three method, we see again that OLS comes out on top.
+This time, it is not because we pick the polynomial degree that does not overfit, but because we are not overfitting at the highest degree we are using for our prediction.
+See Table.~\ref{ols_vs_ridge_vs_lasso_table_real}.
 
-%%%%%%%%%%%%%%%% begin figure %%%%%%%%%%%%%%%%%%%
-%%% the maximum width in double column is 6.85in
-% \begin{figure*} 
-% \centerline{\includegraphics[width=6.85in]{figure/FMANU_MD_04_1274_13.jpg}}
-% \caption{A figure expanded to double column width the text from Figure~\ref{fig_example3.jpg}}
-% \label{fig_example4.jpg}
-% \end{figure*}
-%%%%%%%%%%%%%%%% end figure %%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{figure} 
+\centerline{\includegraphics[width=3.25in]{figure/real1ridgedifflambdas.png}}
+\caption{Test and train MSE for OLS on Terrain 1}
+\label{real1ridgedifflambdas}
+\end{figure}
+
+\begin{table*}[t]
+\caption{Best MSE for OLS, Ridge and Lasso regression on real elevation data}
+\begin{center}
+\label{ols_vs_ridge_vs_lasso_table_real}
+\begin{tabular}{c | l l l}
+Regression & MSE & Degree & $\lambda$ \\
+\hline
+OLS & 0.0142683 & 19 & - \\
+Ridge & 0.017034 & 7 & $10^{-8}$ \\
+Lasso & 0.0095846 & 10 & $2.5118 \cdot 10^{-6}$ \\
+\hline
+\end{tabular}
+\end{center}
+\end{table*}
+
+\section{Conclusion}
+
+In this paper we have compared Ordinary Least Squares, Ridge and Lasso regression for both synthetic and real data.
+We have found that linear regression was well suited for approximating the synthetic Franke function, but less suited for the real elevation data.
+We have seen that Ridge and Lasso gives us the ability to prevent overfitting for more complex models, but that they have little to offer in this case, due to our ability to easily reduce or increase the complexity of our model, and due to the nature of our data.
+We have also seen the value of using regression methods such as bootstrap and cross validation to learn valuable insights about the quality of our predictions.
 
@@
 \begin{table*}
-\caption{This table shows every applicable parameter for every figure, so that they may be easily reproduced. Every plot uses the random seed 42069}
-\begin{center}
-\label{allparamstable}
-\begin{tabular}{c | l l l l l l l}
-% & & \\ % put some space after the caption
-% \hline
-Figure & Highest $N$ & Std of noise & Step (\# of points) & Bootstraps & K-folds & $\lambda$ & Scaling \\
-\hline
-\ref{frankiedifflambdas} & 25 & 0.05 & 0.1 (100) & 100 & - & variable & Centering\\
-\ref{frankeridgebetas} & 25 & 0.05 & 0.1 (100) & - & - & variable & Centering\\
-\hline
-\end{tabular}
-\end{center}
-\end{table*}
-
-\begin{table*}
 \caption{This table shows the command to execute, as well as the advanced parameters to change, to reproduce every figure in the report.
@@
 \begin{tabular}{c | l l l l}
-% & & \\ % put some space after the caption
-% \hline
 Figure & Shell command (leading \texttt{python3} omitted) & Bootstraps & K-folds & $\lambda$ \\
@@
 \ref{frankiebiasvariance} & \texttt{task\_c.py -n 25 --step 0.1}  & 100 & - & - \\
+\ref{frankiedifflambdas} & \texttt{task\_e\_biasvariance.py -n 25}  & 100 & - & - \\
+\ref{frankeridgebetas} & \texttt{task\_e\_betas.py -n 25}  & 100 & - & - \\
 \ref{frankenoisybeta} & \texttt{task\_c.py -n 25}  & 100 & - & - \\
 \ref{frankienoisecomparison} & \texttt{noise\_plot.py} & 100 & - & - \\
+\ref{frankeridgenoisecomparison} & \texttt{noise\_plot.py} \footnotesize{(\it Model has been changed to ridge with lambda=$10^{-8}$)} & 100 & - & - \\
 \ref{frankiehigherdatapoints} & \texttt{task\_c.py --noise 0 -n 30 --step 0.01} & 100 & - & - \\
@@
 \begin{figure*} 
+\centerline{\includegraphics[width=3.25in]{figure/frankeridgebetas.png}}
+\caption{Beta progression for different values of lambda in Ridge regression. Note the different scales on each of the plots, showing how increasing the regularization parameter shrinks the betas}
+\label{frankeridgebetas}
+\end{figure*}
+
+\begin{figure*} 
+\centerline{\includegraphics[width=3.25in]{figure/frankiedifflambdas.png}}
+\caption{Bias, variance and test MSE plotted over polynomial degree for our Ridge prediction on the Franke function with an added stochastic noise $\epsilon \sim N(0,0.05)$, using lambdas of increasing size.
+Here we have reduced the number of datapoints to only 100 to force overfitting earlier.}
+\label{frankiedifflambdas}
+\end{figure*}
+
+\begin{figure*}
 % \begin{figure}[t]
@@
 \begin{figure*}
-% \begin{figure}[t]
+\centerline{\includegraphics[width=3.25in]{figure/frankeridgenoisecomparison.png}}
+\caption{Test and train MSE for the Ridge prediction with $\lambda = 10^{-8}$ of the Franke function with an increasing amount of noise}
+\label{frankeridgenoisecomparison}
+\end{figure*}
+
+\begin{figure*}
 \centerline{\includegraphics[width=3.25in]{figure/real1mselessdata.png}}
@@
 \begin{figure*}
-% \begin{figure}[t]
-\centerline{\includegraphics[width=3.25in]{figure/frankeridgenoisecomparison.png}}
-\caption{Test and train MSE for the Ridge prediction with $\lambda = 10^{-8}$ of the Franke function with an increasing amount of noise}
-\label{frankeridgenoisecomparison}
-\end{figure*}
-
-\begin{figure*}
-% \begin{figure}[t]
 \centerline{\includegraphics[width=3.25in]{figure/frankiehigherdatapoints.png}}
@@
 \begin{figure*}
-% \begin{figure}[t]
 \centerline{\includegraphics[width=3.25in]{figure/frankenoisybeta.png}}
@@
 
+\begin{figure*} 
+\centerline{\includegraphics[width=6.85in]{figure/real1surfacecomp.png}}
+\caption{A comparison of our best OLS prediction and actual terrain data (terrain 1), at polynomial degree 19}
+\label{real1surfacecomp}
+\end{figure*}
+
 \end{document}

### 2022-10-11 src/task_b.py

@@ -23,6 +23,9 @@ np.random.seed(42069)
     z,
 ) = read_from_cmdline()
 
+print(np.min(z))
+print(np.max(z))
+
 # implemented model under testing
 OLS_model = OLS
 # scikit model under testing
@@ -69,14 +72,14 @@ ax = fig.add_subplot(122, projection="3d")
 surf = ax.plot_surface(
     x,
     y,
-    np.reshape(z_preds[:, N], z.shape),
+    np.reshape(z_preds[:, np.argmin(MSE_test)], z.shape),
     cmap=cm.coolwarm,
     linewidth=0,
     antialiased=False,
 )
 ax.zaxis.set_major_locator(LinearLocator(10))
 ax.zaxis.set_major_formatter(FormatStrFormatter("%.02f"))
-ax.set_title("Polynomial fit of scaled terrain")
+ax.set_title(f"Polynomial fit of scaled terrain, N = {np.argmin(MSE_test)}")
 fig.colorbar(surf, shrink=0.5, aspect=5)
 
 plt.show()
@@ -108,9 +111,9 @@ plt.plot(MSE_train, label="train implementation", marker="o", markersize=3)
 plt.plot(MSE_test, label="test implementation", marker="o", markersize=3)
 plt.plot(MSE_train_sk, "r--", label="train ScikitLearn", marker="o", markersize=3)
 plt.plot(MSE_test_sk, "g--", label="test ScikitLearn", marker="o", markersize=3)
-plt.ylabel("MSE score")
-plt.xlabel("Polynomial degree (N)")
-plt.title("MSE scores over model complexity")
+plt.ylabel("MSE score", size=15)
+plt.xlabel("Polynomial degree (N)", size=15)
+plt.title("MSE scores over model complexity", size=22)
 plt.legend()
 plt.show()
 

### 2022-10-11 src/task_c.py
Fix dumbass issue

@@ -42,7 +42,7 @@ OLS_model = OLS
 OLS_scikit = LinearRegression(fit_intercept=False)
 
 # for polynomial degree
-for n in range(N):
+for n in range(N+1):
     print(n)
     l = int((n + 1) * (n + 2) / 2)  # Number of elements in beta
 

### 2022-10-11 src/task_d.py

@@ -66,7 +66,7 @@ for n in range(N):
 
 # run bootstrap
 print("Boot")
-for n in range(N):
+for n in range(N+1):
     print(n)
     l = int((n + 1) * (n + 2) / 2)  # Number of elements in beta
 

### 2022-10-11 src/task_e_biasvariance.py

@@ -50,7 +50,7 @@ for i in range(len(lambdas)):
     variances = np.zeros(N)
 
     # for polynomial degree
-    for n in range(N):
+    for n in range(N + 1):
         print(n)
         l = int((n + 1) * (n + 2) / 2)  # Number of elements in beta
         z_preds, _ = bootstrap(
@@ -79,7 +79,8 @@ for i in range(len(lambdas)):
     plt.tight_layout(h_pad=0.001)
     if plot_only_best_lambda:
         plt.title(
-            f"Bias variance tradeoff for ridge regression for optimal lambda = {lambdas[i]}", size=18
+            f"Bias variance tradeoff for ridge regression for optimal lambda = {lambdas[i]}",
+            size=18,
         )
     else:
         plt.title(f"lambda = {lambdas[i]:.5}", size=15)

### 2022-10-11 src/task_e_msecomparison.py

@@ -76,7 +76,7 @@ for i in range(len(lambdas)):
 
     # calculate bootstrap error
     print("Running Bootstrap")
-    for n in range(N):
+    for n in range(N+1):
         print(n)
         l = int((n + 1) * (n + 2) / 2)  # number of elements in beta
         z_preds_test, _ = bootstrap(

### 2022-10-11 src/task_f_biasvariance.py

@@ -51,7 +51,7 @@ for i in range(len(lambdas)):
     variances = np.zeros(N)
 
     # for polynomial degree
-    for n in range(N):
+    for n in range(N + 1):
         print(n)
         l = int((n + 1) * (n + 2) / 2)  # Number of elements in beta
         z_preds_test, _ = bootstrap(
@@ -83,7 +83,8 @@ for i in range(len(lambdas)):
             f"Optimal lambda = {lam}, best MSE = {best_MSE}, best polynomial = {best_poly}"
         )
         plt.title(
-            f"Bias variance tradeoff for lasso regression \n for optimal lambda = {lambdas[i]}", size=18
+            f"Bias variance tradeoff for lasso regression \n for optimal lambda = {lambdas[i]}",
+            size=18,
         )
     else:
         plt.title(f"lambda = {lambdas[i]:.5}", size=15)

### 2022-10-11 src/utils.py

@@ -188,7 +188,7 @@ def evaluate_model(
             elif model.__name__ == "ridge":
                 beta = model((X_train - X_train_mean), (z_train - z_train_mean), lam)
 
-            intercept = np.mean(z_train_mean - X_train_mean @ beta)
+            intercept = z_train_mean - X_train_mean @ beta
 
         else:
             if model.__name__ == "OLS":

### 2022-10-11 src/utils.py
plot changes

@@ -327,7 +327,7 @@ def find_best_lambda(X, z, model, lambdas, N, K):
     best_lambda = 0
     best_MSE = 10**10
 
-    for n in range(N):
+    for n in range(N-1,N):
         print(n)
         l = int((n + 1) * (n + 2) / 2)  # Number of elements in beta
         model.fit(X[:, :l], z)

### 2022-10-11 doc/paper.tex
MORE ASS PAPER

@@
 To counteract this, we can add to the cost function a term which increases with the values in $\bm{\beta}$.
-Optimizing the cost function would now not only mean finding the smallest difference between target and predicted values, but also keeping the parameters small.
+Optimizing the cost function would now not only mean finding the smallest difference between target and predicted values, but also minimizing the absolute value of the parameters.
 We now define a new cost function
@@
 
-	We see that this equation, with a $\lambda > 0$ avoids the problem of an inversion of a non-singular matrix, and gives us an analytical solution for $\bm{\hat{\beta}}$ yet again.
-\footnote{We should cherish this analytical solution, as it may well be the last time we see a cost function that can be optimized analytically.} By using ridge regresison, we can control the complexity of our model and effectively hinder overfitting.
+We see that this equation, with a $\lambda > 0$ avoids the problem of an inversion of a non-singular matrix, and gives us an analytical solution for $\bm{\hat{\beta}}$ yet again.\footnote{We should cherish this analytical solution, as it may well be the last time we see a cost function that can be optimized analytically.} By using ridge regresison, we can control the complexity of our model and effectively hinder overfitting.
 
@@
 
-However, the derivative of the absolute value is not continous, so there is sadly no analytical solution for the optimal parameters with this form of linear regression.
+However, the derivative of the absolute value is not continous, so there is no analytical solution for the optimal parameters with this form of linear regression.
 We must then solve the problem numerically, which can be done by using Gradient Descent.
-Like Ridge regression, this form of regression controls the complexity of the model, with the added benifit that Lasso can drive parameters to exactly zero.
+Like Ridge regression, this form of regression controls the complexity of the model, with the added benefit that Lasso can drive parameters to exactly zero.
 
@@
 Thirdly, we repeat for a given large number of iterations.
-Since we use all the test predictions to calculate the MSE, bias and variance, we will then according to the central limit theorem (REFERENCE!) obtain a normal distribution of these variables.
+Since we use all the test predictions to calculate the MSE, bias and variance, we will then according to the central limit theorem (REFERENCE!) obtain a normal distribution of these performance metrics.
 The law of large numbers <reference?> then tells us that the mean sample values of the MSE, bias and variance approach the true value of MSE, bias and variance.
@@
 Theoretically, because both two resampling techniques use the central limit theorem and the law of large numbers, they should return the same values for the MSE, bias and variance given the same dataset.
-Though resampling techniques can be computationally expensive, they make up for it by aiding in model evaluation by supplying us with variables obtained from a larger number of samples.
+Though resampling techniques can be computationally expensive, they make up for it by aiding in model evaluation by supplying us with performance metrics obtained from a larger number of samples.
 
@@
 Regardless of the source of the data, we create the design matrix in a similar manner, as the relation between our input data and our target values are the same for both datasets, that is, for each combination of $x$ and $y$, we have a height $z$.
-For the Franke function this height is no more than the height over the $xy$-plane, but for our real data, this value represents the elevation in meters above ocean level for a terrain in Norway.
-As we want to do a polynomial fit of the data, we create our design matrix such that each row $i$ has the following values:
+For our real data, this value represents the elevation in meters above ocean level for a terrain in Norway.
+As we wish to perform a polynomial fit, we create our design matrix such that each row $i$ has the following values:
 
@@
 
-\lstinputlisting[language=Python, firstline=27, lastline=42]{../src/utils.py}
+\lstinputlisting[language=Python, firstline=30, lastline=44]{../src/utils.py}
 
 While the creating of the design matrix is equivalent for the two datasets, the preprocessing of the data afterwards is not.
-For the Franke function, we claim that no normalization is needed, as we generate the data ourselves to be between $0$ and $1$.
+For the Franke function no scaling is needed, as we generate the data ourselves to be between $0$ and $1$.
 Given that we are doing a polynomial fit, every element in our design matrix will be a multiplication of $x^p$ and $y^k$ for different $p$'s and $k$'s, which, given our choice of range for $x$ and $y$, guarantees that every element is between $0$ and $1$.
 However, we have included a centering of the data, which has no effect on the Ordinary Least Squares regression, but which is used when applying Ridge and Lasso regression, to avoid punishing the intercept.
-The centering is done as follows: First, the first coloumn of the design matrix (the intercept coloumn) is discarderd.
+The centering is done as follows: First, the first coloumn of the design matrix (the intercept coloumn) is discarded.
 Then we calculate the mean for the target values and the design matrix (the mean for the design matrix is a vector with mean values for each coloumn).
-We subtract this from the design matrix and target values when performing the fit giving us a vector $\bm{\beta}$.
+We subtract this from the design matrix and the target values when performing the fit giving us a vector $\bm{\beta}$.
 Afterwards we calculate the intercept with the following equation
@@
 \begin{equation}
-	intercept = mean(\bar{y} - \bar{X}_c^T \bm{\beta})
+	intercept = \bar{y} - \bm{\bar{x}_c}^T \bm{\beta}
 \end{equation}
 
-	where $\bar{y}$ is the mean value of $\bm{y}$ and $\bm{\bar{X}_c}$ is a vector of size $p$ with $\bar{X}_i = mean(X_i)$ (the mean of each coloumn in $X$).
+	where $\bar{y}$ is the mean value of $\bm{y}$ and $\bm{\bar{x}_c}$ is a vector of size $p$ with $\bar{X}_i = mean(X_i)$ (the mean of each coloumn in $X$).
 We finally create our prediction by using the $\bm{\beta}$, $X$ and intercept we calculated above:
@@
 
-The code for this, taken from the \texttt{evaluate\_model} function in \texttt{src/utils.py} is as follows: \footnote{Some intermediate parts of our implementation not related to this have been removed for readability}
+The code for this, taken from the \texttt{evaluate\_model} function in \texttt{src/utils.py} is as follows:\footnote{Some intermediate parts of our implementation not related to this have been removed for readability}
 
@@
 
-For the real elevation data, normalization is definitely required, as the values for $x$, $y$ reach several hundred and $z$ reaches over a thousand.
+For the real elevation data, scaling is definitely required, as the values for $x$, $y$ reach several hundred and $z$ reaches over a thousand.
 For polynomial degree $10$ and an $x$ value which reaches $400$, we will have a design matrix with values from $0$ to $400^{10}$.
@@
 We have therefore chosen to scale our design matrix using Scikit's \texttt{MinMaxScaler}, which scales each column to be in the range of $(0,1)$.
-The calculation of mean and variance is done on the training data, and applied to both the training and test data.
+The scaling coefficients are calculated on the training data, and applied to both the training and test data.
 We have done the same with the values for $z$.
@@
 Another form of "preprocessing" which is strictly practical in nature, is that we have reduced the dimensions of the data to 320 by 640.
-This was done by downscaling the image.
-\footnote{The interpolation was done using "nearest neighbor interpolation", which does not introduce new values into the data.
-An equivalent programmatic solution would be to for instance only look at every 4 pixels} This was simply done because the design matrix becomes several gigabytes in size if one uses the whole dataset, and model fitting takes an unreasonable amount of time.
+This was done by downscaling the image.\footnote{The interpolation was done using "nearest neighbor interpolation", which does not introduce new values into the data \cite[77]{dip}.
+An equivalent programmatic solution would be to for instance only look at every 4 pixels} This was simply done because the design matrix becomes several gigabytes in size if one uses the whole dataset, and model fitting then takes an unreasonable amount of time.
 
@@
 This is important because our $x$'s and $y$'s are not normal distributions but consecutive values between $0$ and $1$, and therefore consecutive rows correspond specific areas of our data, which is not wanted.
-We then fit on $K-1$ folds, and test on the last.
+We then fit on $K-1$ folds, and test on the remaining fold.
 We return the average error.
@@
 Without a regularization hyperparameter, parameters fluctuate wildly and our prediction suffers.
-Fig.~\ref{frankenoisemse} shows our MSE up to polynomial degree 25 with an added stochastic noise $\epsilon \sim N(0,0.05)$, where we see that we are overfitting severely.
+Fig.~\ref{frankenoisemse} shows our MSE up to polynomial degree 24 with an added stochastic noise $\epsilon \sim N(0,0.05)$, where we see that we are overfitting severely.
 See Fig.~\ref{frankienoisecomparison} in the appendix for test MSE for more and more noisy target values.
@@
 \end{figure}
-
-% \begin{figure} 
-% \centerline{\includegraphics[width=3.25in]{figure/frankenoisybeta.png}}
-% \caption{Bias, variance and test MSE plotted over polynomial degree for our OLS prediction on the Franke function with an added stochastic noise $\epsilon \sim N(0,0.05)$}
-% \label{frankenoisybeta}
-% \end{figure}
 
@@
 Here we find some limitations of linear regression.
-We see from Fig.~\ref{real1msetraintest} that the we are able to fit the terrain up to a certain point, gaining an MSE of $0.0142683$ with OLS with polynomial degree 19.
+We see from Fig.~\ref{real1msetraintest} that the we are able to fit the terrain up to a certain point, gaining an MSE of $0.0142683$ with OLS with polynomial degree 19, our highest polynomial.
 This is about two orders of magnitude higher than our lowest MSE for the Franke function without noise, which had values on a similar scale.\footnote{The Franke function without noise has a range of about $(0,1.2)$ and our MinMaxed terrain data has been scaled to a range of $(0,1)$.
@@
 
-\begin{figure} 
-\centerline{\includegraphics[width=3.25in]{figure/real1msetraintest.png}}
-\caption{Test and train MSE for OLS on Terrain 1}
-\label{real1msetraintest}
-\end{figure}
+% \begin{figure} 
+% \centerline{\includegraphics[width=3.25in]{figure/real1msetraintest.png}}
+% \caption{Test and train MSE for OLS on Terrain 1}
+% \label{real1msetraintest}
+% \end{figure}
+
 \begin{figure} 
@@
 This proves to be the case, as we see from Fig.~\ref{real1ridgedifflambdas}.
-Here we see that for low values of lambda there is no improvement, as there are no parameters that need regularization, and no overfitting to supress.
-For larger values, we see that we are shrinking the parameters too much, and we see an increase in MSE.
+Here we see that different values of lambda have very little impact, as there are no parameters that need regularization, and no overfitting to supress.
+For very large values, we see that we are shrinking the parameters too much, and we see a slight increase in MSE, but overall we do not see much of an effect.
 We could expect to see that Ridge shows improvement at higher polynomial degrees, when OLS starts to overfit, but due to computational constraints we have not been able to push much further than 20 degrees.
+Using scikit's GridSearch, we see some surprising results that we assume are due to us
 
@@
 \centerline{\includegraphics[width=3.25in]{figure/real1ridgedifflambdas.png}}
-\caption{Test and train MSE for OLS on Terrain 1}
+\caption{Bias variance tradeoff for ridge regression with different lambdas}
 \label{real1ridgedifflambdas}
@@
 OLS & 0.0142683 & 19 & - \\
-Ridge & 0.017034 & 7 & $10^{-8}$ \\
+Ridge & 0.015433 & 19 & $18.32$ \\
 Lasso & 0.0095846 & 10 & $2.5118 \cdot 10^{-6}$ \\
@@
 \ref{frankenonoise} & \texttt{task\_b.py --noise 0 -n 25} & - & - & - \\
-\ref{frankenonoisemse} & \texttt{task\_c.py --noise 0 -n 25}  & 100 & - & - \\
-\ref{frankenoisemse} & \texttt{task\_c.py -n 25}  & 100 & - & - \\
-\ref{frankiebiasvariance} & \texttt{task\_c.py -n 25 --step 0.1}  & 100 & - & - \\
-\ref{frankiedifflambdas} & \texttt{task\_e\_biasvariance.py -n 25}  & 100 & - & - \\
-\ref{frankeridgebetas} & \texttt{task\_e\_betas.py -n 25}  & 100 & - & - \\
-\ref{frankenoisybeta} & \texttt{task\_c.py -n 25}  & 100 & - & - \\
+\ref{frankenonoisemse} & \texttt{task\_c.py --noise 0 -n 24}  & 100 & - & - \\
+\ref{frankenoisemse} & \texttt{task\_c.py -n 24}  & 100 & - & - \\
+\ref{frankiebiasvariance} & \texttt{task\_c.py -n 24 --step 0.1}  & 100 & - & - \\
+\ref{frankiedifflambdas} & \texttt{task\_e\_biasvariance.py -n 24}  & 100 & - & - \\
+\ref{real1ridgedifflambdas} & \texttt{task\_e\_biasvariance.py --file ../data/small\_*\_1.tif -n 24}  & 100 & - & - \\
+\ref{frankeridgebetas} & \texttt{task\_e\_betas.py -n 24}  & 100 & - & - \\
+\ref{frankenoisybeta} & \texttt{task\_c.py -n 24}  & 100 & - & - \\
 \ref{frankienoisecomparison} & \texttt{noise\_plot.py} & 100 & - & - \\
@@
 \ref{frankiehigherdatapoints} & \texttt{task\_c.py --noise 0 -n 30 --step 0.01} & 100 & - & - \\
-\ref{real1surfacecomp} & \texttt{task\_b.py --file ../data/small\_SRTM\_data\_Norway\_1.tif -n 19} & - & - & - \\
-\ref{real1mselessdata} & \texttt{task\_c.py -f ../data/tiny\_SRTM\_data\_Norway\_1.tif -n 20} & 100 & - & - \\
+\ref{real1surfacecomp} & \texttt{task\_b.py --file ../data/small\_*\_1.tif -n 19} & - & - & - \\
+\ref{real1mselessdata} & \texttt{task\_c.py --file ../data/tiny\_SRTM\_data\_Norway\_1.tif -n 19} & 100 & - & - \\
+\ref{real1msetraintest} & \texttt{task\_c.py --file ../data/small\_*\_1.tif -n 19} & 100 & - & - \\
+\ref{realbiasvarianceOLS} & \texttt{task\_c.py --file ../data/small\_*\_1.tif -n 19} & 100 & - & - \\
 \hline

### 2022-10-11 src/utils.py
Snaker

@@ -354,7 +354,7 @@ def read_from_cmdline():
     group.add_argument("-no", "--noise", help="Amount of noise to have. Recommended range [0-0.1]. Default 0.05", type=float, default=0.05)
     parser.add_argument("-st", "--step", help="Step size for linspace function. Range [0.01-0.4]. Default 0.05", type=float, default=0.05)
     parser.add_argument("-b", "--betas", help="Betas to plot, when applicable. Default 10", type=int)
-    parser.add_argument("-n", help="Polynomial degree. Default 10", type=int, default=10)
+    parser.add_argument("-n", help="Polynomial degree. Default 9", type=int, default=9)
     parser.add_argument("-nsc", "--noscale", help="Do not use scaling (centering for synthetic case or MinMaxScaling for organic case)", action="store_true") 
 
     # parse arguments and call run_filter

### 2022-10-11 doc/paper.tex
More paper

@@
 Our problem now is to optimize this function, that is, find the values of $\beta$ which gives the MSE the lowest value.
-Luckily for us, it can be proven that the second derivative of this function is always positive (WHERE?), which means that we can find the global minimum by solving the equation
-
-Hastie 45
+Luckily for us, it can be proven that the second derivative of this function is always positive \cite[106]{linreg}, which means that we can find the global minimum by solving the equation
 
@@
 We find that OLS creates a good fit for the Franke Function.
-With 400 datapoints, we can increase our polynomial degree to around 20 before we see signs of overfitting.\footnote{Here we also see something a bit surprising: We see that our own implementation of OLS follows Scikit's implementation up to around polynomial degree 10, at which point scikit's MSE increases rapidly.
+With 400 datapoints, we can increase our polynomial degree to 20 before we see signs of overfitting.\footnote{Here we also see something a bit surprising: We see that our own implementation of OLS follows Scikit's implementation up to around polynomial degree 10, at which point scikit's MSE increases rapidly.
 We chalk this up to numerical errors and differing implementation of the normal equations, as we simply use the normal equation exactly, while scikit uses an implementation which is probably computationally more efficient but gives slight differences when comparing to the normal equation.
@@
 Our synthetic data used $20 x 20 = 400$ datapoints, while we now have $320 x 640 = 204 800$ datapoints.
-This means that for a 20th order polynomial, which will have $231$ features, we will have around a thousand datapoints per feature.
+This means that for a 19th order polynomial, which will have $210$ features, we will have around a thousand datapoints per feature.
 If we decrease the number of datapoints availiable considerably, we can see overfitting start to occur (Fig.~\ref{real1mselessdata} in the appendix).
 So we are not overfitting, but we are not decreasing our MSE either.
-This is likely because the elevation data has a jagged and complex shape, which cannot be accurately recreated, even with a polynomial of degree $50$.
+This is likely because the elevation data has a jagged and complex shape, which cannot be accurately recreated, even with a polynomial of degree $19$.
 We should remember that the Franke function can be expanded into a series of polynomials, which is why we can expect to get a lower and lower MSE as the polynomial degree increases (so long as we have enough data and it is not noisy).
@@
 For very large values, we see that we are shrinking the parameters too much, and we see a slight increase in MSE, but overall we do not see much of an effect.
-We could expect to see that Ridge shows improvement at higher polynomial degrees, when OLS starts to overfit, but due to computational constraints we have not been able to push much further than 20 degrees.
-Using scikit's GridSearch, we see some surprising results that we assume are due to us
+We could expect to see that Ridge shows improvement at higher polynomial degrees, when OLS starts to overfit, but due to computational constraints we have not been able to push much further than 19 degrees.
+Using scikit's GridSearch, we see some surprising results that we assume are due to the fact that we are not overfitting.
+We see that our best MSE is higher than for OLS, which is surprising given that Ridge with $\lambda = 0$ is equivalent to OLS.
+However, we must here consider that our lowest OLS MSE was found by bootstrap with 100 iterations, and our grid searched best lambda and MSE was found by crossvalidation with 10 folds.
+Given that we do not expect large differences in values with either method because we are not overfitting, it is not so unusual that our Ridge MSE is slightly higher.
 

### 2022-10-11 src/task_b.py
remove prints, add todo info

@@ -23,9 +23,6 @@ np.random.seed(42069)
     z,
 ) = read_from_cmdline()
 
-print(np.min(z))
-print(np.max(z))
-
 # implemented model under testing
 OLS_model = OLS
 # scikit model under testing

### 2022-10-11 doc/paper.tex
Paper finito

@@
 
-To define the optimal prediction, we use a cost function, and the choice of cost function greatly influences what optimal parameters we get.\footnote{It is here important to keep in mind that the cost function used when fitting the model (training) rarely is the same as the one used for testing our prediction, as described in <somewhere>} An intuitive cost function is the mean squared error (MSE).
+To define the optimal prediction, we use a cost function, and the choice of cost function greatly influences what optimal parameters we get.\footnote{It is here important to keep in mind that the cost function used when fitting the model (training) rarely is the same as the one used for testing our prediction} An intuitive cost function is the mean squared error (MSE).
 This function simply computes the average distance between our prediction and the target value, squared:
@@
 Thirdly, we repeat for a given large number of iterations.
-Since we use all the test predictions to calculate the MSE, bias and variance, we will then according to the central limit theorem (REFERENCE!) obtain a normal distribution of these performance metrics.
-The law of large numbers <reference?> then tells us that the mean sample values of the MSE, bias and variance approach the true value of MSE, bias and variance.
+Since we use all the test predictions to calculate the MSE, bias and variance, we will then according to the central limit theorem \cite{CLT} obtain a normal distribution of these performance metrics.
+The law of large numbers \cite{LLN} then tells us that the mean sample values of the MSE, bias and variance approach the true value of MSE, bias and variance.
 This is dependent on our assumption that our data is independently and identically distributed.
@@
 But we cannot avoid considering the fact that Lasso does not have an analytical solution, and as such we must use numerical methods to find the optimal parameters.
-Using scikit's implementation, we have not been able to get the numerical method to converge, which means that we have not found a global minimum.\footnote{Scikit explicitly mentions that using $\lambda = 0$ does not converge well with their implementation} For this reason, Lasso regression has not given satisfying results.
+Using scikit's implementation, we have not been able to get the numerical method to converge, which means that we have not found a global minimum.\footnote{Scikit explicitly mentions that using $\lambda = 0$ does not converge well with their implementation \cite{lasso}} For this reason, Lasso regression has not given satisfying results.
 
@@
 This implies that we are not overfitting yet, and indeed the bias variance decomposition confirms this, as we see that variance is essentially inconsequential (Fig.~\ref{realbiasvarianceOLS}).
-%This can be seen when we increase the polynomial degree to very large amounts, without seeing any decrease in MSE (Fig.
-NUMBER).
-Looking at the beta progression for selected betas (Fig.
-NUMBER), we see that the parameters stabilize, which implies that we are not overfitting, but simply reaching our best possible prediction for polynomial regression.
+The fact that we are not overfitting can be seen when we increase the polynomial degree to very large amounts, seeing only a very modest decrease in MSE to $0.01381$ (Fig.~\ref{realOLSmse40} in the appendix).
+Comparing this to tenfold decrease in MSE when increasing the polynomial degree for the Franke function, we see a clear difference.
+Looking at the beta progression for selected betas (Fig.\ref{realOLSbetaprog40} in the appendix), we see that the parameters stabilize, which again implies that we are not overfitting, but simply reaching our best possible prediction for polynomial regression.
 The fact that we are not overfitting is further reinforced by looking at the bias-variance plot in Fig.~\ref{realbiasvarianceOLS}, which shows no increase in variance.
@@
 As our model is not overfitting, we do not expect Ridge to offer much improvement.
-This proves to be the case, as we see from Fig.~\ref{real1ridgedifflambdas}.
+This proves to be the case, as we see from Fig.~\ref{real1ridgedifflambdas} in the appendix.
 Here we see that different values of lambda have very little impact, as there are no parameters that need regularization, and no overfitting to supress.
@@
 With lasso regression, we see similar results as with Ridge.
-Comparing all the three method, we see again that OLS comes out on top.
-This time, it is not because we pick the polynomial degree that does not overfit, but because we are not overfitting at the highest degree we are using for our prediction.
-See Table.~\ref{ols_vs_ridge_vs_lasso_table_real}.
+Fig.~\ref{real1lassobiasvariance} in the appendix shows little change, except for when our $\lambda$ is too high and we force all parameters to $0$.
+Comparing lasso to the two other methods for the real dataset has proven to be difficult, as the numerical methods for finding the best parameters are very time consuming.
+We have not been able to gridsearch up to polynomial degree 19, but up to degree 9 we found the best lambda to be $2.154^{-7}$ for polynomial degree 9, giving an MSE of $0.01685$.
 
-\begin{figure} 
-\centerline{\includegraphics[width=3.25in]{figure/real1ridgedifflambdas.png}}
-\caption{Bias variance tradeoff for ridge regression with different lambdas}
-\label{real1ridgedifflambdas}
-\end{figure}
+Comparing all three methods up to degree 9 for different lambdas (using the same lambda for Lasso and Ridge, just to get an overview) in Fig.~\ref{real1msecomparison} in the appendix, we see the same trend as we have been describing.
+Regularization offers little reward, which is to be expected when we are not overfitting.
 
 \begin{table*}[t]
-\caption{Best MSE for OLS, Ridge and Lasso regression on real elevation data}
+\caption{Best MSE for OLS and Ridge regression on real elevation data}
 \begin{center}
@@
 Ridge & 0.015433 & 19 & $18.32$ \\
-Lasso & 0.0095846 & 10 & $2.5118 \cdot 10^{-6}$ \\
 \hline
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\appendix       %%% starting appendix
 \section*{Appendix A: Plots and tables}
@@
 \begin{table*}
-\caption{This table shows the command to execute, as well as the advanced parameters to change, to reproduce every figure in the report.
-(More info about the scripts and their parameters can be found in the README)}
+\caption{This table shows the command to execute to reproduce every figure in the report. (More info about the scripts and their parameters can be found in the README)}
 \begin{center}
 \label{allparamstable}
-\begin{tabular}{c | l l l l}
-Figure & Shell command (leading \texttt{python3} omitted) & Bootstraps & K-folds & $\lambda$ \\
+\begin{tabular}{c | l l l}
+Figure & Shell command (leading \texttt{python3} omitted) \\
 \hline
-\ref{frankenonoise} & \texttt{task\_b.py --noise 0 -n 25} & - & - & - \\
-\ref{frankenonoisemse} & \texttt{task\_c.py --noise 0 -n 24}  & 100 & - & - \\
-\ref{frankenoisemse} & \texttt{task\_c.py -n 24}  & 100 & - & - \\
-\ref{frankiebiasvariance} & \texttt{task\_c.py -n 24 --step 0.1}  & 100 & - & - \\
-\ref{frankiedifflambdas} & \texttt{task\_e\_biasvariance.py -n 24}  & 100 & - & - \\
-\ref{real1ridgedifflambdas} & \texttt{task\_e\_biasvariance.py --file ../data/small\_*\_1.tif -n 24}  & 100 & - & - \\
-\ref{frankeridgebetas} & \texttt{task\_e\_betas.py -n 24}  & 100 & - & - \\
-\ref{frankenoisybeta} & \texttt{task\_c.py -n 24}  & 100 & - & - \\
-\ref{frankienoisecomparison} & \texttt{noise\_plot.py} & 100 & - & - \\
-\ref{frankeridgenoisecomparison} & \texttt{noise\_plot.py} \footnotesize{(\it Model has been changed to ridge with lambda=$10^{-8}$)} & 100 & - & - \\
-\ref{frankiehigherdatapoints} & \texttt{task\_c.py --noise 0 -n 30 --step 0.01} & 100 & - & - \\
-\ref{real1surfacecomp} & \texttt{task\_b.py --file ../data/small\_*\_1.tif -n 19} & - & - & - \\
-\ref{real1mselessdata} & \texttt{task\_c.py --file ../data/tiny\_SRTM\_data\_Norway\_1.tif -n 19} & 100 & - & - \\
-\ref{real1msetraintest} & \texttt{task\_c.py --file ../data/small\_*\_1.tif -n 19} & 100 & - & - \\
-\ref{realbiasvarianceOLS} & \texttt{task\_c.py --file ../data/small\_*\_1.tif -n 19} & 100 & - & - \\
+\ref{frankenonoise} & \texttt{task\_b.py --noise 0 -n 25}\\
+\ref{frankenonoisemse} & \texttt{task\_c.py --noise 0 -n 24}\\
+\ref{frankenoisemse} & \texttt{task\_c.py -n 24}\\
+\ref{frankiebiasvariance} & \texttt{task\_c.py -n 24 --step 0.1}\\
+\ref{frankiedifflambdas} & \texttt{task\_e\_biasvariance.py -n 24}\\
+\ref{real1ridgedifflambdas} & \texttt{task\_e\_biasvariance.py --file ../data/small\_SRTM\_data\_Norway\_1.tif -n 24}\\
+\ref{real1msecomparison} & \texttt{task\_f\_biasvariance.py --file ../data/small\_SRTM\_data\_Norway\_1.tif -n 9}\\
+\ref{frankeridgebetas} & \texttt{task\_e\_betas.py -n 24}\\
+\ref{frankenoisybeta} & \texttt{task\_c.py -n 24}\\
+\ref{frankienoisecomparison} & \texttt{noise\_plot.py}\\
+\ref{frankeridgenoisecomparison} & \texttt{noise\_plot.py} \footnotesize{(\it Model has been changed to ridge with lambda=$10^{-8}$)}\\
+\ref{frankiehigherdatapoints} & \texttt{task\_c.py --noise 0 -n 30 --step 0.01}\\
+\ref{real1surfacecomp} & \texttt{task\_b.py --file ../data/small\_SRTM\_data\_Norway\_1.tif -n 19}\\
+\ref{real1mselessdata} & \texttt{task\_c.py --file ../data/tiny\_SRTM\_data\_Norway\_1.tif -n 19}\\
+\ref{real1msetraintest} & \texttt{task\_c.py --file ../data/small\_SRTM\_data\_Norway\_1.tif -n 19}\\
+\ref{realbiasvarianceOLS} & \texttt{task\_c.py --file ../data/small\_SRTM\_data\_Norway\_1.tif -n 19}\\
+\ref{realOLSmse40} & \texttt{task\_b.py --file ../data/small\_SRTM\_data\_Norway\_1.tif -n 40}\\
+\ref{realOLSbetaprog40} & \texttt{task\_b.py --file ../data/small\_SRTM\_data\_Norway\_1.tif -n 40}\\
+\ref{real1msecomparison} & \texttt{task\_f\_msecomparison.py --file ../data/small\_SRTM\_data\_Norway\_1.tif -n 9}\\
 \hline
@@
 \begin{figure*}
+\centerline{\includegraphics[width=3.25in]{figure/realOLSmse40.png}}
+\caption{Test and train MSE on our terrain, this time up to degree 40.
+We see only a slight decrease in MSE, down to $0.01381$, even though we double the amount of degrees.
+This has not been bootstrapped due to computational limitations}
+\label{realOLSmse40}
+\end{figure*}
+
+\begin{figure*}
+\centerline{\includegraphics[width=3.25in]{figure/realOLSbetaprog40.png}}
+\caption{Beta progression up to polynomial degree 40 for real terrain data, showing that the are relatively constant, implying that we are not overfitting}
+\label{realOLSbetaprog40}
+\end{figure*}
+
+\begin{figure*}
 \centerline{\includegraphics[width=3.25in]{figure/frankiehigherdatapoints.png}}
@@
 \begin{figure*}
+\centerline{\includegraphics[width=3.25in]{figure/real1msecomparison.png}}
+\caption{Comparison between OLS, Ridge and Lasso for different values of lambda, for real terrain data}
+\label{real1msecomparison}
+\end{figure*}
+
+\begin{figure*} 
 \centerline{\includegraphics[width=6.85in]{figure/real1surfacecomp.png}}
@@
 
+\begin{figure} 
+\centerline{\includegraphics[width=3.25in]{figure/real1ridgedifflambdas.png}}
+\caption{Bias variance tradeoff for ridge regression with different lambdas, for the real dataset}
+\label{real1ridgedifflambdas}
+\end{figure}
+
+\begin{figure} 
+\centerline{\includegraphics[width=3.25in]{figure/real1lassobiasvariance.png}}
+\caption{Bias variance tradeoff for lasso regression with different lambdas, for the real dataset}
+\label{real1lassobiasvariance}
+\end{figure}
+
 \end{document}

### 2022-10-11 src/task_c.py
fixed indexing to new system

@@ -27,14 +27,14 @@ bootstraps = 100
 ) = read_from_cmdline()
 
 # result arrays
-MSEs_test = np.zeros(N)
-MSEs_train = np.zeros(N)
-biases = np.zeros(N)
-variances = np.zeros(N)
+MSEs_test = np.zeros(N+1)
+MSEs_train = np.zeros(N+1)
+biases = np.zeros(N+1)
+variances = np.zeros(N+1)
 
 # scikit
-MSEs_test_sk = np.zeros(N)
-MSEs_train_sk = np.zeros(N)
+MSEs_test_sk = np.zeros(N+1)
+MSEs_train_sk = np.zeros(N+1)
 
 
 # model under testing

### 2022-10-11 src/task_d.py

@@ -37,13 +37,13 @@ OLS_scikit = LinearRegression(fit_intercept=False)
 kfolds = KFold(n_splits=K)
 
 # cross val results
-errors_cv = np.zeros(N)
-errors_cv_scikit = np.zeros(N)
+errors_cv = np.zeros(N+1)
+errors_cv_scikit = np.zeros(N+1)
 
 # bootstrap results
-errors_boot = np.zeros(N)
-biases_boot = np.zeros(N)
-variances_boot = np.zeros(N)
+errors_boot = np.zeros(N+1)
+biases_boot = np.zeros(N+1)
+variances_boot = np.zeros(N+1)
 
 # run cross val
 print("Cross")

### 2022-10-11 src/task_e_msecomparison.py

@@ -48,9 +48,9 @@ for i in range(len(lambdas)):
         plt.suptitle(f"MSE by polynomial degree for resampling methods Ridge")
 
     # arrays for MSE scores
-    errors_cv = np.zeros(N)
-    errors_cv_scikit = np.zeros(N)
-    errors_boot = np.zeros(N)
+    errors_cv = np.zeros(N+1)
+    errors_cv_scikit = np.zeros(N+1)
+    errors_boot = np.zeros(N+1)
 
     # scikit model under testing
     Ridge_model = Ridge(lambdas[i], fit_intercept=False)

### 2022-10-11 src/task_f_msecomparison.py

@@ -29,9 +29,9 @@ kfolds = KFold(n_splits=K)
     z,
 ) = read_from_cmdline()
 
-errors_OLS = np.zeros(N)
-errors_Ridge = np.zeros(N)
-errors_Lasso = np.zeros(N)
+errors_OLS = np.zeros(N+1)
+errors_Ridge = np.zeros(N+1)
+errors_Lasso = np.zeros(N+1)
 
 min_error_OLS = np.inf
 min_error_Ridge = np.inf

### 2022-10-11 src/utils.py

@@ -327,7 +327,7 @@ def find_best_lambda(X, z, model, lambdas, N, K):
     best_lambda = 0
     best_MSE = 10**10
 
-    for n in range(N-1,N):
+    for n in range(N+1):
         print(n)
         l = int((n + 1) * (n + 2) / 2)  # Number of elements in beta
         model.fit(X[:, :l], z)

### 2022-10-11 doc/paper.tex
More

@@
 
+\subsection{Resampling}
+
+As stated in the method section, we suspect that bootstrap and cross validation both yield approximations to the true performance metrics due to the central limit theorem, and thus should give identical results.
+We can observe from Fig.~\ref{frankemsebyresamplingmethod} that our implementation of bootstrap and cross validation yield almost exactly the same results.
+(Fig.~\ref{realmsebyresamplingmethod} in the appendix shows a similar result for the real data)
+
+\begin{figure}[h]
+\centerline{\includegraphics[width=3.25in]{figure/frankemsebyresamplingmethod.png}}
+\caption{Comparison between approximated MSE for different resampling methods for the Franke function}
+\label{frankemsebyresamplingmethod}
+\end{figure}
+
 \subsection{Franke Function}
@@
 \subsubsection{Ordinary Least Squares}
-
-%%%%%%% FIGURES %%%%%%%%%
-\begin{figure*} 
-\centerline{\includegraphics[width=6.85in]{figure/frankenonoise.png}}
-\caption{A comparison of our best OLS prediction and the Franke function with no noise, at polynomial degree 10}
-\label{frankenonoise}
-\end{figure*}
-
-\begin{figure} 
-\centerline{\includegraphics[width=3.25in]{figure/frankenonoisemse.png}}
-\caption{Test and train MSE for the Franke function with no noise, using OLS}
-\label{frankenonoisemse}
-\end{figure}
 
@@
 
+\begin{figure}[h]
+\centerline{\includegraphics[width=3.25in]{figure/frankenonoise.png}}
+\caption{A comparison of our best OLS prediction and the Franke function with no noise, at polynomial degree 10}
+\label{frankenonoise}
+\end{figure}
+
+\begin{figure}[h]
+\centerline{\includegraphics[width=3.25in]{figure/frankenonoisemse.png}}
+\caption{Test and train MSE for the Franke function with no noise, using OLS}
+\label{frankenonoisemse}
+\end{figure}
+
 However, when we introduce a stochastic noise, we see that overfitting happens rapidly.
@@
 
-\begin{figure} 
+\begin{figure}[h]
 \centerline{\includegraphics[width=3.25in]{figure/frankenoisemse.png}}
@@
 
-\begin{figure} 
+\begin{figure}[h]
 \centerline{\includegraphics[width=3.25in]{figure/frankiebiasvariance.png}}
@@
 
-\begin{table}[t]
+\begin{table}[h]
 \caption{Best MSE and polynomial degree for OLS}
@@
 
-Looking at Lasso regression, we find that it has the same capability to reduce overfitting as Ridge, as can be seen in Fig.~\ref{frankelassodifflambdas}.
+Looking at Lasso regression, we find that it has the same capability to reduce overfitting as Ridge, as can be seen in Fig.~\ref{frankedifflambdaslasso} in the appendix.
 However, looking at Table.~\ref{ols_vs_ridge_vs_lasso_table}, we see that Lasso falls short when finding a good MSE.
@@
 With the confidence that our methods work as expected on synthetic data, we turn our attention to the real elevation data.
+First, let us confirm that it is necessary to scale our values using MinMax-scaling, as we have described in our implementation.
+Fig.~\ref{realunscaled} and Fig.~\ref{realunscaledmse} in the appendix shows our predicted terrain and associated MSE, showing that we reach a very poor prediction without scaling.
 
@@
 While not exactly the same, these are on the same order of magnitude, and we believe that comparing the MSE scores is reasonable} In fact, our MSE is comparable to the Franke function with a stochastic noise $\epsilon \sim N(0,0.1)$ (Table.~\ref{ols_mse_table_dif_data}), which implies that our terrain is very irregular and difficult to approximate.
-Looking at the plot for polynomial degree 19 (Fig.~\ref{real1surfacecomp}), we see that our poor MSE is reflected visually in our fit as well, as the surfaces do not look particularly similar.
+Looking at the plot for polynomial degree 19 (Fig.~\ref{real1surfacecomp} in the appendix), we see that our poor MSE is reflected visually in our fit as well, as the surfaces do not look particularly similar.
 Comparing the best polynomial degree of the best fit with the synthetic data, we see that our best fit is reached much later.
@@
 
-\begin{figure} 
+\begin{figure}[h]
 \centerline{\includegraphics[width=3.25in]{figure/realbiasvarianceOLS.png}}
@@
 
-\begin{figure} 
+\begin{figure}[h]
 \centerline{\includegraphics[width=3.25in]{figure/real1msetraintest.png}}
@@
 \ref{frankenonoise} & \texttt{task\_b.py --noise 0 -n 25}\\
+\ref{realunscaled} & \texttt{task\_b.py --file ../data/small\_SRTM\_data\_Norway\_1.tif -n 19 --noscale}\\
+\ref{realunscaledmse} & \texttt{task\_b.py --file ../data/small\_SRTM\_data\_Norway\_1.tif -n 19 --noscale}\\
 \ref{frankenonoisemse} & \texttt{task\_c.py --noise 0 -n 24}\\
@@
 \ref{frankiebiasvariance} & \texttt{task\_c.py -n 24 --step 0.1}\\
-\ref{frankiedifflambdas} & \texttt{task\_e\_biasvariance.py -n 24}\\
+\ref{frankiedifflambdas} & \texttt{task\_e\_biasvariance.py -n 24 --step 0.1}\\
+\ref{frankedifflambdaslasso} & \texttt{task\_f\_biasvariance.py -n 9}\\
 \ref{real1ridgedifflambdas} & \texttt{task\_e\_biasvariance.py --file ../data/small\_SRTM\_data\_Norway\_1.tif -n 24}\\
@@
 \ref{realOLSbetaprog40} & \texttt{task\_b.py --file ../data/small\_SRTM\_data\_Norway\_1.tif -n 40}\\
+\ref{frankemsebyresamplingmethod} & \texttt{task\_d.py -n 15 --step 0.02}\\
 \ref{real1msecomparison} & \texttt{task\_f\_msecomparison.py --file ../data/small\_SRTM\_data\_Norway\_1.tif -n 9}\\
@@
 
+%%%%%%%%%%% FIGURES %%%%%%%%%%%%%%%%%%%%%%%%%5
+
 \begin{figure*}
-\centerline{\includegraphics[width=3.25in]{figure/frankeridgebetas.png}}
-\caption{Beta progression for different values of lambda in Ridge regression. Note the different scales on each of the plots, showing how increasing the regularization parameter shrinks the betas}
-\label{frankeridgebetas}
+\centerline{\includegraphics[width=3.25in]{figure/realmsebyresamplingmethod.png}}
+\caption{Comparison between approximated MSE for different resampling methods for the real terrain data}
+\label{realmsebyresamplingmethod}
+\end{figure*}
+
+\begin{figure*}
+\centerline{\includegraphics[width=3.25in]{figure/frankiehigherdatapoints.png}}
+\caption{Franke function fitted up to polynomial degree 30, with 10000 datapoints instead of 400.
+As we can see, we do not overfit even at this polynomial degree when the availiable data is plentiful.
+The highest MSE was reached at $N=29$, and was $0.00003511$, one tenth of our best prediction with only 400 datapoints}
+\label{frankiehigherdatapoints}
+\end{figure*}
+
+\begin{figure*}
+% \begin{figure}[t]
+\centerline{\includegraphics[width=3.25in]{figure/frankienoisecomparison.png}}
+\caption{Test and train MSE for the OLS prediction of Franke function with an increasing amount of noise}
+\label{frankienoisecomparison}
+\end{figure*}
+
+\begin{figure*}
+\centerline{\includegraphics[width=3.25in]{figure/frankenoisybeta.png}}
+\caption{Beta progression for OLS}
+\label{frankenoisybeta}
 \end{figure*}
@@
 \begin{figure*} 
-% \begin{figure}[t]
-\centerline{\includegraphics[width=3.25in]{figure/frankienoisecomparison.png}}
-\caption{Test and train MSE for the OLS prediction of Franke function with an increasing amount of noise}
-\label{frankienoisecomparison}
+\centerline{\includegraphics[width=3.25in]{figure/frankeridgebetas.png}}
+\caption{Beta progression for different values of lambda in Ridge regression. Note the different scales on each of the plots, showing how increasing the regularization parameter shrinks the betas}
+\label{frankeridgebetas}
 \end{figure*}
@@
 \begin{figure*}
-\centerline{\includegraphics[width=3.25in]{figure/real1mselessdata.png}}
-\caption{Test and train MSE for the OLS prediction of terrain 1, this time downscaled to only $50x100$. Here the optimal degree was 11, showing that with lower data we start to overfit more rapidly.}
-\label{real1mselessdata}
+\centerline{\includegraphics[width=3.25in]{figure/frankedifflambdaslasso.png}}
+\caption{Bias, variance and test MSE plotted over polynomial degree for our Lasso prediction on the Franke function with an added stochastic noise $\epsilon \sim N(0,0.05)$, using lambdas of increasing size.
+Here we have reduced the number of datapoints to only 100 to force overfitting earlier.}
+\label{frankedifflambdaslasso}
+\end{figure*}
+
+\begin{figure*}
+\centerline{\includegraphics[width=3.25in]{figure/realunscaled.png}}
+\caption{Our prediction for the real terrain using unscaled data, showing that we reach a very poor result}
+\label{realunscaled}
+\end{figure*}
+
+\begin{figure*}
+\centerline{\includegraphics[width=3.25in]{figure/realunscaledmse.png}}
+\caption{Test and train MSE for our prediction for the real terrain using unscaled data, showing that we reach a very poor result}
+\label{realunscaledmse}
+\end{figure*}
+
+\begin{figure*} 
+\centerline{\includegraphics[width=6.85in]{figure/real1surfacecomp.png}}
+\caption{A comparison of our best OLS prediction and actual terrain data (terrain 1), at polynomial degree 19}
+\label{real1surfacecomp}
 \end{figure*}
@@
 \begin{figure*}
-\centerline{\includegraphics[width=3.25in]{figure/frankiehigherdatapoints.png}}
-\caption{Franke function fitted up to polynomial degree 30, with 10000 datapoints instead of 400.
-As we can see, we do not overfit even at this polynomial degree when the availiable data is plentiful.
-The highest MSE was reached at $N=29$, and was $0.00003511$, one tenth of our best prediction with only 400 datapoints}
-\label{frankiehigherdatapoints}
-\end{figure*}
-
-\begin{figure*}
-\centerline{\includegraphics[width=3.25in]{figure/frankenoisybeta.png}}
-\caption{Beta progression for OLS}
-\label{frankenoisybeta}
-\end{figure*}
-
-\begin{figure*}
-\centerline{\includegraphics[width=3.25in]{figure/real1msecomparison.png}}
-\caption{Comparison between OLS, Ridge and Lasso for different values of lambda, for real terrain data}
-\label{real1msecomparison}
-\end{figure*}
-
-\begin{figure*} 
-\centerline{\includegraphics[width=6.85in]{figure/real1surfacecomp.png}}
-\caption{A comparison of our best OLS prediction and actual terrain data (terrain 1), at polynomial degree 19}
-\label{real1surfacecomp}
+\centerline{\includegraphics[width=3.25in]{figure/real1mselessdata.png}}
+\caption{Test and train MSE for the OLS prediction of terrain 1, this time downscaled to only $50x100$. Here the optimal degree was 11, showing that with lower data we start to overfit more rapidly.}
+\label{real1mselessdata}
 \end{figure*}
@@
 
+\begin{figure*}
+\centerline{\includegraphics[width=3.25in]{figure/real1msecomparison.png}}
+\caption{Comparison between OLS, Ridge and Lasso for different values of lambda, for real terrain data}
+\label{real1msecomparison}
+\end{figure*}
+
 \end{document}

### 2022-10-11 src/task_e_msecomparison.py
fixed cross validation error

@@ -12,7 +12,7 @@ np.random.seed(42069)
 # Parameters
 K = 20
 bootstraps = 100
-kfolds = KFold(n_splits=K)
+kfolds = KFold(n_splits=K, shuffle=True)
 
 # if true, plot only lambda that gives lowest MSE
 # if false, plot range of different lambdas
@@ -102,7 +102,7 @@ for i in range(len(lambdas)):
     plt.xlabel("Polynomial degree (N)")
     if plot_only_best_lambda:
         plt.title(
-            f"MSE by polynomial degree for resampling methods for optimal lambda = {lambdas[i]} Ridge"
+            f"MSE by polynomial degree for resampling methods \n for optimal lambda = {lambdas[i]} Ridge"
         )
     else:
         plt.title(f"lambda = {lambdas[i]:.5}")

### 2022-10-11 src/task_f_msecomparison.py

@@ -12,7 +12,7 @@ from utils import *
 K = 10
 lambdas = np.logspace(-10, 0, 4)
 
-kfolds = KFold(n_splits=K)
+kfolds = KFold(n_splits=K, shuffle=True)
 
 (
     betas_to_plot,

### 2022-10-11 src/utils.py

@@ -4,7 +4,7 @@ from matplotlib import cm
 from matplotlib.ticker import LinearLocator, FormatStrFormatter
 import numpy as np
 from random import random, seed
-from sklearn.model_selection import train_test_split, GridSearchCV
+from sklearn.model_selection import train_test_split, GridSearchCV, KFold
 from sklearn.linear_model import LinearRegression
 from sklearn.preprocessing import MinMaxScaler
 from sklearn.utils import resample
@@ -317,11 +317,12 @@ def scores(z, z_preds):
 
 
 def find_best_lambda(X, z, model, lambdas, N, K):
+    kfolds = KFold(n_splits=K, shuffle=True)
     model = GridSearchCV(
         estimator=model,
         param_grid={"alpha": list(lambdas)},
         scoring="neg_mean_squared_error",
-        cv=K,
+        cv=kfolds,
     )
     best_polynomial = 0
     best_lambda = 0

### 2022-10-11 src/task_e_biasvariance.py
fix details

@@ -11,7 +11,7 @@ np.random.seed(42069)
 # parameters
 K = 20
 bootstraps = 100
-plot_only_best_lambda = False
+plot_only_best_lambda = True
 lambdas = np.logspace(-10, 0, 4)
 
 (
@@ -33,12 +33,14 @@ z = z.ravel()
 
 # plot only the gridsearched lambda
 if plot_only_best_lambda:
-    lambdas = np.logspace(-8, 3, 20)
+    lambdas = np.logspace(-8, 3, 15)
     lambdas[0] = 0
     ridge = Ridge(fit_intercept=False)
     lam, best_MSE, best_poly = find_best_lambda(X, z, ridge, lambdas, N, K)
     lambdas = [lam]
+    f"Optimal lambda = {lam}, best MSE = {best_MSE}, best polynomial = {best_poly}"
 
+raise ValueError
 # for lambdas
 for i in range(len(lambdas)):
     if not plot_only_best_lambda:

### 2022-10-11 src/task_e_msecomparison.py

@@ -37,6 +37,8 @@ z = z.ravel()
 
 # plot only the gridsearched lambda
 if plot_only_best_lambda:
+    lambdas = np.logspace(-8, 3, 20)
+    lambdas[0] = 0
     ridge = Ridge(fit_intercept=False)
     lam, _, _ = find_best_lambda(X, z, ridge, lambdas, N, K)
     lambdas = [lam]

### 2022-10-11 src/task_f_msecomparison.py

@@ -9,7 +9,7 @@ from sklearn.model_selection import cross_validate, KFold
 from utils import *
 
 # parameters
-K = 10
+K = 20
 lambdas = np.logspace(-10, 0, 4)
 
 kfolds = KFold(n_splits=K, shuffle=True)
@@ -54,7 +54,7 @@ for i in range(len(lambdas)):
 
     # cross validate, get error
     print("OLS")
-    for n in range(N):
+    for n in range(N+1):
         print(n)
         l = int((n + 1) * (n + 2) / 2)  # Number of elements in beta
 

### 2022-10-11 doc/paper.tex
mroe

@@
 \section{Abstract}
-We used linear regression to create a fit of a synthetic dataset generated by the Franke Function, and for real datasets of elevation data from terrain in Norway.
+We used linear regression to create a fit of a synthetic dataset generated by the Franke Function, and for a real dataset of elevation data from a terrain in Norway.
 More specifically, we used Ordinary Least Squares (OLS), Ridge and Lasso regression to explore how the different methods perform on different datasets.
-Comparing the Mean Squared Error (MSE) for the different methods, we found that Ridge and Lasso outperform OLS for small datasets and high polynomial degrees, but that OLS, given the optimal polynomial degree for the dataset, yields the lowest MSE.
+On the synthetic dataset, we found that Ridge regression allowed us to increase our model complexity and achieve lower MSE scores than OLS, but that Lasso struggled due to computational constraints.
+On the real dataset, we found
+% Comparing the Mean Squared Error (MSE) for the different methods, we found that Ridge and Lasso outperform OLS for small datasets and high polynomial degrees, but that OLS, given the optimal polynomial degree for the dataset, yields the lowest MSE.
 Overall we found that the different forms of linear regression fit the synthetic data well.
@@
 Looking at the beta progression with for the same values of $\lambda$ (Fig.~\ref{frankeridgebetas} in the appendix), we see that the regularization parameter is successfully shrinking the beta values.
-Using Scikit's gridsearch, we find the optimal $\lambda$ to be $10^{-8}$ and the optimal polynomial degree to be 7, with an average MSE of $0.0037999$, which is very slightly lower than OLS for the Franke function with an added stochastic noise $\epsilon \sim N(0,0.05)$.
-We see from this that Ridge regression has not really had much of an effect in decreasing the MSE.
+Using Scikit's gridsearch, we find the optimal $\lambda$ to be $1.43 \cdot 10^{-7}$ and the optimal polynomial degree to be 13, with an average MSE of $0.003027$.
+This is a good improvement of our OLS fit, and shows that we are able to push our model complexity significantly higher and gain a moderate MSE improvement with a correctly chosen regularization parameter.
+Looking at the beta progression for this value of lambda Fig.~\ref{frankebestlambdabetas}, we see that the beta values settle and do not vary greatly between polynomial degrees.
+Fig.~\ref{frankebestlambdamse} shows the MSE score, which shows how we are successfully avoiding overfitting.
+Table.~\ref{ols_vs_ridge_vs_lasso_table} shows the relevant values.
+
+\begin{figure}[h]
+\centerline{\includegraphics[width=3.25in]{figure/frankebestlambdabetas.png}}
+\caption{Beta progression for our Ridge prediction with the optimal lambda}
+\label{frankebestlambdabetas}
+\end{figure}
+
+\begin{figure}[h]
+\centerline{\includegraphics[width=3.25in]{figure/frankebestlambdamse.png}}
+\caption{Test and train MSE for ridge}
+\label{frankebestlambdamse}
+\end{figure}
+
+% We see from this that Ridge regression has not really had much of an effect in decreasing the MSE.
 We must now remember what Ridge aims to do.
@@
 OLS & 0.05 & 0.0038302 & 7 & - \\
-Ridge & 0.05 & 0.0037999 & 7 & $10^{-8}$ \\
-Lasso & 0.05 & 0.0095846 & 10 & $2.5118 \cdot 10^{-6}$ \\
+Ridge & 0.05 & 0.003027 & 13 & $1.43 \cdot 10^{-7}$ \\
+Lasso & 0.05 & 0.008291 & 11 & $7.847 \cdot 10^{-7}$ \\
 \hline
@@
 The fact that we are not overfitting can be seen when we increase the polynomial degree to very large amounts, seeing only a very modest decrease in MSE to $0.01381$ (Fig.~\ref{realOLSmse40} in the appendix).
-Comparing this to tenfold decrease in MSE when increasing the polynomial degree for the Franke function, we see a clear difference.
+Comparing this to the tenfold decrease in MSE when increasing the polynomial degree for the Franke function, we see a clear difference.
 Looking at the beta progression for selected betas (Fig.\ref{realOLSbetaprog40} in the appendix), we see that the parameters stabilize, which again implies that we are not overfitting, but simply reaching our best possible prediction for polynomial regression.
@@
 For very large values, we see that we are shrinking the parameters too much, and we see a slight increase in MSE, but overall we do not see much of an effect.
-We could expect to see that Ridge shows improvement at higher polynomial degrees, when OLS starts to overfit, but due to computational constraints we have not been able to push much further than 19 degrees.
-Using scikit's GridSearch, we see some surprising results that we assume are due to the fact that we are not overfitting.
-We see that our best MSE is higher than for OLS, which is surprising given that Ridge with $\lambda = 0$ is equivalent to OLS.
-However, we must here consider that our lowest OLS MSE was found by bootstrap with 100 iterations, and our grid searched best lambda and MSE was found by crossvalidation with 10 folds.
-Given that we do not expect large differences in values with either method because we are not overfitting, it is not so unusual that our Ridge MSE is slightly higher.
+We could expect to see that Ridge shows improvement at higher polynomial degrees, when OLS starts to overfit, but due to computational constraints we have not been able to push much further than 9 degrees.
+Using scikit's GridSearch, we get that the optimal lambda is 0, which is exactly what we would expect on a model that does not overfit.
+See Table.~\ref{ols_vs_ridge_vs_lasso_table_real}.
 
@@
 \begin{table*}[t]
-\caption{Best MSE for OLS and Ridge regression on real elevation data}
+\caption{Best MSE for OLS, Ridge and Lasso regression on real elevation data. We only fit up to degree 9 due to computational constraints}
 \begin{center}
@@
 \hline
-OLS & 0.0142683 & 19 & - \\
-Ridge & 0.015433 & 19 & $18.32$ \\
+OLS & 0.01538 & 9 & - \\
+Ridge & 0.01538 & 9 & 0 \\
+Lasso & 0.01538 & 9 & 0 \\
 \hline
@@
 We have found that linear regression was well suited for approximating the synthetic Franke function, but less suited for the real elevation data.
-We have seen that Ridge and Lasso gives us the ability to prevent overfitting for more complex models, but that they have little to offer in this case, due to our ability to easily reduce or increase the complexity of our model, and due to the nature of our data.
+We have seen that a correctly chosen regularization parameter can allow us to push our model complexity higher than we can with OLS, and thus give us a better MSE than OLS.
+% but that they have little to offer in this case, due to our ability to easily reduce or increase the complexity of our model, and due to the nature of our data.
 We have also seen the value of using regression methods such as bootstrap and cross validation to learn valuable insights about the quality of our predictions.
@@
 \ref{frankenonoise} & \texttt{task\_b.py --noise 0 -n 25}\\
+\ref{frankebestlambdamse} & \texttt{task\_b.py --noise 0 -n 25} \footnotesize{(\it Model has been changed to ridge with lambda=$1.43 \cdot 10^{-7}$)}\\
+\ref{frankebestlambdabetas} & \texttt{task\_b.py --noise 0 -n 25} \footnotesize{(\it Model has been changed to ridge with lambda=$1.43 \cdot 10^{-7}$)}\\
 \ref{realunscaled} & \texttt{task\_b.py --file ../data/small\_SRTM\_data\_Norway\_1.tif -n 19 --noscale}\\
@@
 \ref{frankienoisecomparison} & \texttt{noise\_plot.py}\\
-\ref{frankeridgenoisecomparison} & \texttt{noise\_plot.py} \footnotesize{(\it Model has been changed to ridge with lambda=$10^{-8}$)}\\
 \ref{frankiehigherdatapoints} & \texttt{task\_c.py --noise 0 -n 30 --step 0.01}\\
@@
 
-\begin{figure*}
-\centerline{\includegraphics[width=3.25in]{figure/frankeridgenoisecomparison.png}}
-\caption{Test and train MSE for the Ridge prediction with $\lambda = 10^{-8}$ of the Franke function with an increasing amount of noise}
-\label{frankeridgenoisecomparison}
-\end{figure*}
+% \begin{figure*}
+% \centerline{\includegraphics[width=3.25in]{figure/frankeridgenoisecomparison.png}}
+% \caption{Test and train MSE for the Ridge prediction with $\lambda = 10^{-8}$ of the Franke function with an increasing amount of noise}
+% \label{frankeridgenoisecomparison}
+% \end{figure*}
 

### 2022-10-11 doc/paper.tex
grammar change
More paper

@@
 First, we will go more into depth about the methods and the theory behind them, along with a brief description of our implementation.
-We will further continue to the results, followed up by a critical discussion of the results.
+We will further continue to the results, alongside a critical discussion of the results.
 Finally, we will give a conclusion on the optimal model selection for the different datasets.
@@
 \ref{frankiedifflambdas} & \texttt{task\_e\_biasvariance.py -n 24 --step 0.1}\\
-\ref{frankedifflambdaslasso} & \texttt{task\_f\_biasvariance.py -n 9}\\
+\ref{frankedifflambdaslasso} & \texttt{task\_f\_biasvariance.py -n 9 --step 0.15}\\
 \ref{real1ridgedifflambdas} & \texttt{task\_e\_biasvariance.py --file ../data/small\_SRTM\_data\_Norway\_1.tif -n 24}\\
-\ref{real1msecomparison} & \texttt{task\_f\_biasvariance.py --file ../data/small\_SRTM\_data\_Norway\_1.tif -n 9}\\
 \ref{frankeridgebetas} & \texttt{task\_e\_betas.py -n 24}\\
@@
 \caption{Bias, variance and test MSE plotted over polynomial degree for our Lasso prediction on the Franke function with an added stochastic noise $\epsilon \sim N(0,0.05)$, using lambdas of increasing size.
-Here we have reduced the number of datapoints to only 100 to force overfitting earlier.}
+Here we have reduced the number of datapoints to less than 100 to force overfitting earlier.}
 \label{frankedifflambdaslasso}

### 2022-10-11 doc/paper.tex
Paper

@@
 \section{Abstract}
-We used linear regression to create a fit of a synthetic dataset generated by the Franke Function, and for a real dataset of elevation data from a terrain in Norway.
+We used linear regression to create a fit of a synthetic dataset generated by the Franke function, and of a real dataset of elevation data from a terrain in Norway.
 More specifically, we used Ordinary Least Squares (OLS), Ridge and Lasso regression to explore how the different methods perform on different datasets.
 On the synthetic dataset, we found that Ridge regression allowed us to increase our model complexity and achieve lower MSE scores than OLS, but that Lasso struggled due to computational constraints.
-On the real dataset, we found
+On the real dataset, we found that regularization had little impact, as we did not reach model complexities in which overfitting occured.
 % Comparing the Mean Squared Error (MSE) for the different methods, we found that Ridge and Lasso outperform OLS for small datasets and high polynomial degrees, but that OLS, given the optimal polynomial degree for the dataset, yields the lowest MSE.
@@
 
-In this paper, we will provide a detailed inspection of the Ordinary Least Squares (OLS), Ridge and Lasso linear regression methods, it's hyperparameters and varying model complexity with the aim of making an informed decision about model selection when fitting to different datasets.
-The datasets consist of synthetic data created by the Franke function, and real datasets of terrain in Norway.
+In this paper, we will provide a detailed inspection of the Ordinary Least Squares (OLS), Ridge and Lasso linear regression methods, their hyperparameters and varying model complexities with the aim of making an informed decision about model selection when fitting to different datasets.
+The datasets consist of synthetic data created by the Franke function, and a real dataset of terrain in Norway.
 We inform our decisions by plotting the mean squared error (MSE), our cost function, over different polynomial degrees describing model complexity, over different resampling techniques and over different values of lambda, which is the hyperparameter responsible for a suitable regularization.
@@
 First, we will go more into depth about the methods and the theory behind them, along with a brief description of our implementation.
-We will further continue to the results, alongside a critical discussion of the results.
+We will further continue to the results, alongside a critical discussion of their implications.
 Finally, we will give a conclusion on the optimal model selection for the different datasets.
@@
 \subsection{Implementation}
-\subsubsection{Preprocessing: Creating the design matrix and normalizing our data}
+\subsubsection{Preprocessing: Creating the design matrix and scaling our data}
 We have two datasets, one synthetic and one real.
@@
 For polynomial degree $10$ and an $x$ value which reaches $400$, we will have a design matrix with values from $0$ to $400^{10}$.
-This large range of values makes our linear regression perform poorly (WHY?).
+This large range of values makes our linear regression perform poorly.
 We have therefore chosen to scale our design matrix using Scikit's \texttt{MinMaxScaler}, which scales each column to be in the range of $(0,1)$.
@@
 
-\lstinputlisting[language=Python, firstline=238, lastline=245]{../src/utils.py}
+\lstinputlisting[language=Python, firstline=242, lastline=249]{../src/utils.py}
 
@@
 We would expect to see higher polynomial degrees give us better and better prediction, up to a certain point.\footnote{And indeed, this seems to be the case, as if we increase the availiable datapoints considerably, as seen in Fig.~\ref{frankiehigherdatapoints} in the appendix, we can reach much higher polynomial degrees without overfitting} When we reach degrees higher than around 20, we see that our test error increases slightly, a sign of overfitting.
-This is also to be expected, as even if the continual Franke function could be expanded into a series of polynomials, we are working with a limited dataset of only 400 samples (and only 320 of those in our training set).
-As we reach higher and higher degrees, we are more and more likely to see unlucky train-test-splits where the best fit for the training data leads to large errors for the test data.
+This is also to be expected, as even if the continous Franke function could be expanded into a series of polynomials, we are working with a limited dataset of only 400 samples (and only 80\% of those in our training set).
+As we reach higher and higher degrees, we are more and more likely to see unfortunate train-test-splits where the best fit for the training data leads to large errors for the test data.
+Regardless, our fit for lower model complexity is remarkably good.
 See Fig.~\ref{frankenonoise} for a comparison between the Franke function and our best prediction, which shows that they are essentially identical.
@@
 The MSE for test in these plots is the average MSE for 100 bootstraps.
-For a comprehensive list of the shell command required to reproduce this and every other plot, see Table.~\ref{allparamstable} in the appendix.
+For a comprehensive list of the shell command required to reproduce this and every other plot, see Table~\ref{allparamstable} in the appendix.
 
@@
 Without a regularization hyperparameter, parameters fluctuate wildly and our prediction suffers.
-Fig.~\ref{frankenoisemse} shows our MSE up to polynomial degree 24 with an added stochastic noise $\epsilon \sim N(0,0.05)$, where we see that we are overfitting severely.
+Fig.~\ref{frankenoisemse} shows our MSE up to polynomial degree 24 with an added stochastic noise $\epsilon \sim N(0,0.05)$, where we see that we are overfitting severely starting at around degree 7.
 See Fig.~\ref{frankienoisecomparison} in the appendix for test MSE for more and more noisy target values.
@@
 
-Table.
-~\ref{ols_mse_table} shows the best polynomial degree and MSE for regular and noisy data.
+Table ~\ref{ols_mse_table} shows the best polynomial degree and MSE for regular and noisy data.
 Here we see that the best degree without noise is 9, which shows that even without noise, the low amount of data leads to a slight overfit even at very small polynomial degrees.
-This is not surprising, as at $N = 10$ we have $66$ features, which is a substantial amount compared to our measly 320 training datapoints.
+This is not surprising, as at $N = 9$ we have $55$ features, which is a substantial amount compared to our measly 320 training datapoints.
 We could of course arbitrarily increase our dataset to ludicrous levels, but doing linear regression with this dataset gives valuable insight into how our performance is dependent on our availiable data.
@@
 Using Ridge regression, we see from Fig.~\ref{frankiedifflambdas} in the appendix that larger and larger values of $\lambda$ efficiently prevents overfitting and reduces the variance between predictions.
-Looking at the beta progression with for the same values of $\lambda$ (Fig.~\ref{frankeridgebetas} in the appendix), we see that the regularization parameter is successfully shrinking the beta values.
-Using Scikit's gridsearch, we find the optimal $\lambda$ to be $1.43 \cdot 10^{-7}$ and the optimal polynomial degree to be 13, with an average MSE of $0.003027$.
+Looking at the beta progression for the same values of $\lambda$ (Fig.~\ref{frankeridgebetas} in the appendix), we see that the regularization parameter is successfully shrinking the beta values.
+Using Scikit's gridsearch, we find the optimal $\lambda$ to be $6.10 \cdot 10^{-8}$ and the optimal polynomial degree to be 13, with an average MSE of $0.003027$.
 This is a good improvement of our OLS fit, and shows that we are able to push our model complexity significantly higher and gain a moderate MSE improvement with a correctly chosen regularization parameter.
-Looking at the beta progression for this value of lambda Fig.~\ref{frankebestlambdabetas}, we see that the beta values settle and do not vary greatly between polynomial degrees.
+Looking at the beta progression for this value of lambda (see Fig.~\ref{frankebestlambdabetas}), we see that the beta values settle and do not vary greatly between polynomial degrees.
 Fig.~\ref{frankebestlambdamse} shows the MSE score, which shows how we are successfully avoiding overfitting.
-Table.~\ref{ols_vs_ridge_vs_lasso_table} shows the relevant values.
+Table~\ref{ols_vs_ridge_vs_lasso_table} shows the relevant values.
 
@@
 We could expect that Ridge would allow us to reach higher polynomial degrees without overfitting, by shrinking features of the higher polynomial degrees that are not supported but allowing us to keep other high degree terms (for example, allowing us to use $x^5y^4$, but discard $x^9$, if only $x^5y^4$ was supported by the data).
-But we can see from the Table.
-~\ref{ols_vs_ridge_vs_lasso_table} that we are not able to push into higher degrees.
+But we can see from the Table ~\ref{ols_vs_ridge_vs_lasso_table} that we are not able to push into higher degrees.
 A possible explanation for this is that every feature is useful, due to the polynomial expansion of the Franke function.
@@
 Looking at Lasso regression, we find that it has the same capability to reduce overfitting as Ridge, as can be seen in Fig.~\ref{frankedifflambdaslasso} in the appendix.
-However, looking at Table.~\ref{ols_vs_ridge_vs_lasso_table}, we see that Lasso falls short when finding a good MSE.
+However, looking at Table~\ref{ols_vs_ridge_vs_lasso_table}, we see that Lasso falls short when finding a good MSE.
 Intuitively, it does not make sense for Lasso to be worse than OLS, because Lasso with $\lambda = 0$ is equivalent to OLS.
@@
 OLS & 0.05 & 0.0038302 & 7 & - \\
-Ridge & 0.05 & 0.003027 & 13 & $1.43 \cdot 10^{-7}$ \\
-Lasso & 0.05 & 0.008291 & 11 & $7.847 \cdot 10^{-7}$ \\
+Ridge & 0.05 & 0.003027 & 13 & $6.10 \cdot 10^{-8}$ \\
+Lasso & 0.05 & 0.008298 & 12 & $1.38 \cdot 10^{-5}$ \\
 \hline
@@
 This is about two orders of magnitude higher than our lowest MSE for the Franke function without noise, which had values on a similar scale.\footnote{The Franke function without noise has a range of about $(0,1.2)$ and our MinMaxed terrain data has been scaled to a range of $(0,1)$.
-While not exactly the same, these are on the same order of magnitude, and we believe that comparing the MSE scores is reasonable} In fact, our MSE is comparable to the Franke function with a stochastic noise $\epsilon \sim N(0,0.1)$ (Table.~\ref{ols_mse_table_dif_data}), which implies that our terrain is very irregular and difficult to approximate.
-Looking at the plot for polynomial degree 19 (Fig.~\ref{real1surfacecomp} in the appendix), we see that our poor MSE is reflected visually in our fit as well, as the surfaces do not look particularly similar.
+While not exactly the same, these are on the same order of magnitude, and we believe that comparing the MSE scores is reasonable} In fact, our MSE is comparable to the Franke function with a stochastic noise $\epsilon \sim N(0,0.1)$ (Table~\ref{ols_mse_table_dif_data}), which implies that our terrain is very irregular and difficult to approximate.
+Looking at the plot for polynomial degree 19 (Fig.~\ref{real1surfacecomp} in the appendix), we see that our poor MSE is reflected visually in our fit as well, as we can only make out general similarities in shape, not finer details.
 Comparing the best polynomial degree of the best fit with the synthetic data, we see that our best fit is reached much later.
 This implies that we are not overfitting yet, and indeed the bias variance decomposition confirms this, as we see that variance is essentially inconsequential (Fig.~\ref{realbiasvarianceOLS}).
-The fact that we are not overfitting can be seen when we increase the polynomial degree to very large amounts, seeing only a very modest decrease in MSE to $0.01381$ (Fig.~\ref{realOLSmse40} in the appendix).
-Comparing this to the tenfold decrease in MSE when increasing the polynomial degree for the Franke function, we see a clear difference.
+The fact that we are not overfitting can be seen when we increase the polynomial degree to very large amounts, seeing only a very modest decrease in MSE to $0.01381$ at degree 40 (Fig.~\ref{realOLSmse40} in the appendix).
+Comparing this to the tenfold decrease in MSE when increasing the polynomial degree for the Franke function to 50, we see a clear difference.
 Looking at the beta progression for selected betas (Fig.\ref{realOLSbetaprog40} in the appendix), we see that the parameters stabilize, which again implies that we are not overfitting, but simply reaching our best possible prediction for polynomial regression.
-The fact that we are not overfitting is further reinforced by looking at the bias-variance plot in Fig.~\ref{realbiasvarianceOLS}, which shows no increase in variance.
 It is reasonable to expect that our model will not overfit, due to the large increase in data we now use for the fit.
@@
 
-\begin{table}[t]
+\begin{table}
 \caption{Best MSE and polynomial degree for OLS on synthetic and real data}
@@
 For very large values, we see that we are shrinking the parameters too much, and we see a slight increase in MSE, but overall we do not see much of an effect.
-We could expect to see that Ridge shows improvement at higher polynomial degrees, when OLS starts to overfit, but due to computational constraints we have not been able to push much further than 9 degrees.
-Using scikit's GridSearch, we get that the optimal lambda is 0, which is exactly what we would expect on a model that does not overfit.
-See Table.~\ref{ols_vs_ridge_vs_lasso_table_real}.
+We could expect to see that Ridge shows improvement at higher polynomial degrees, when OLS starts to overfit, but due to computational constraints we have not been able to push our grid search much further than 9 degrees.
+Using scikit's GridSearch, we find that the optimal lambda is 0, which is exactly what we would expect on a model that does not overfit, as Ridge regression with a lambda equal to 0 is equivalent with OLS.
+See Table~\ref{ols_vs_ridge_vs_lasso_table_real}.
 
@@
 Ridge & 0.01538 & 9 & 0 \\
-Lasso & 0.01538 & 9 & 0 \\
+Lasso & 0.01685 & 9 & $2.154 \cdot 10^{-7}$ \\
 \hline
@@
 In this paper we have compared Ordinary Least Squares, Ridge and Lasso regression for both synthetic and real data.
-We have found that linear regression was well suited for approximating the synthetic Franke function, but less suited for the real elevation data.
+We have seen the value of using resampling methods such as bootstrap and cross validation to gain valuable insights about the quality of our predictions.
+Using resampling methods, we have found that linear regression was well suited for approximating the synthetic Franke function, but less suited for the real elevation data.
 We have seen that a correctly chosen regularization parameter can allow us to push our model complexity higher than we can with OLS, and thus give us a better MSE than OLS.
-% but that they have little to offer in this case, due to our ability to easily reduce or increase the complexity of our model, and due to the nature of our data.
-We have also seen the value of using regression methods such as bootstrap and cross validation to learn valuable insights about the quality of our predictions.
+Looking at the real data, we have seen that regularization does not offer much value when we are not overfitting our dataset.
+Thus all three linear regression methods perform similarly on this dataset.
 

### 2022-10-11 src/task_e_biasvariance.py
final commit

@@ -38,9 +38,7 @@ if plot_only_best_lambda:
     ridge = Ridge(fit_intercept=False)
     lam, best_MSE, best_poly = find_best_lambda(X, z, ridge, lambdas, N, K)
     lambdas = [lam]
-    f"Optimal lambda = {lam}, best MSE = {best_MSE}, best polynomial = {best_poly}"
 
-raise ValueError
 # for lambdas
 for i in range(len(lambdas)):
     if not plot_only_best_lambda:

### 2022-10-11 src/task_f_biasvariance.py

@@ -38,7 +38,6 @@ if plot_only_best_lambda:
     lam, best_MSE, best_poly = find_best_lambda(X, z, lasso, lambdas, N, K)
     lambdas = [lam]
 
-raise ValueError
 # loop through different lambda values
 for i in range(len(lambdas)):
     if not plot_only_best_lambda:

### 2022-10-11 src/task_e_biasvariance.py
finaler commit

@@ -11,7 +11,7 @@ np.random.seed(42069)
 # parameters
 K = 20
 bootstraps = 100
-plot_only_best_lambda = True
+plot_only_best_lambda = False
 lambdas = np.logspace(-10, 0, 4)
 
 (

### 2022-10-11 src/task_f_biasvariance.py

@@ -12,7 +12,7 @@ np.random.seed(42069)
 # parameters
 K = 20
 bootstraps = 100
-plot_only_best_lambda = True
+plot_only_best_lambda = False
 lambdas = np.logspace(-10, 0, 4)
 (
     betas_to_plot,
